---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hexify)
```

```{r cover-image, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4, fig.align='center'}
# Generate cover image: multi-resolution grids over different regions
library(sf)
library(ggplot2)

# Get world basemap
countries <- hexify_world

# Generate grids at different resolutions for specific regions
# Coarse grid: South America
grid_coarse <- hexify_grid_rect(
  minlon = -80, maxlon = -35,
  minlat = -40, maxlat = 10,
  area = 200000
)

# Medium grid: Europe/Africa
grid_medium <- hexify_grid_rect(
  minlon = -10, maxlon = 40,
  minlat = -10, maxlat = 50,
  area = 30000
)

# Fine grid: East Asia
grid_fine <- hexify_grid_rect(
  minlon = 90, maxlon = 140,
  minlat = 10, maxlat = 50,
  area = 8000
)

ggplot() +
  geom_sf(data = countries, fill = "gray95", color = "gray80", linewidth = 0.15) +
  geom_sf(data = grid_coarse, fill = NA, color = "#1B9E77", linewidth = 0.5) +
  geom_sf(data = grid_medium, fill = NA, color = "#D95F02", linewidth = 0.4) +
  geom_sf(data = grid_fine, fill = NA, color = "#7570B3", linewidth = 0.3) +
  coord_sf(xlim = c(-100, 150), ylim = c(-50, 60)) +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", color = NA))
```

# Spatial Analysis Done Right

You want to do spatial statistics, and it involves binning points into grid cells.

**The problem with rectangular grids**: A rectangular lat-lon grid introduces severe distortions. At the equator, a 1° cell covers ~12,300 km². Near the poles, the same 1° cell covers a tiny fraction of that area. This breaks any analysis that assumes equal sampling effort or comparable cell sizes.

**The solution**: Discrete global grids partition Earth's surface into cells of **equal area**, regardless of latitude. hexify implements the ISEA (Icosahedral Snyder Equal Area) projection, providing hexagonal cells that are all the same size from the equator to the Arctic.

### Why Equal-Area Matters

```{r equal-area-demo, message=FALSE}
# Same-sized cells at different latitudes
test_points <- data.frame(
  location = c("Equator", "Mid-latitude", "Arctic"),
  lon = c(0, 0, 0),
  lat = c(0, 45, 70)
)

result <- hexify(test_points, lon = "lon", lat = "lat", area = 1000)

# All cells have the same area, regardless of latitude
result[, c("location", "lat", "cell_area")]
```

With hexify, a 1000 km² cell at the equator is the same size as a 1000 km² cell in Norway.


## Installation

```{r, eval = FALSE}
# Install from GitHub
remotes::install_github("gcol33/hexify")
```

**Required packages**: `sf` (for spatial operations)

## What hexify Does

hexify assigns geographic coordinates to equal-area hexagonal grid cells. Given a data frame with longitude/latitude columns, `hexify()` returns:

- `cell_id`: Unique cell identifier (DGGRID-compatible SEQNUM)
- `cell_cen_lon`, `cell_cen_lat`: Cell center coordinates
- `cell_area`: Actual cell area in km²
- `cell_diag`: Cell diagonal in km

Supports apertures 3, 4, 7, and mixed 4/3:

| Aperture | Grid Type | Cell Count Formula | Use Case |
|----------|-----------|-------------------|----------|
| 3 | ISEA3H | 10 × 3^res + 2 | Default, dggridR compatible, fine control |
| 4 | ISEA4H | 10 × 4^res + 2 | Power-of-2 scaling |
| 7 | ISEA7H | 10 × 7^res + 2 | Rapid cell growth |
| "4/3" | ISEA43H | Mixed | Balance of fast start + fine control |

## Quick Start

### Basic Usage

```{r basic-usage}
# Sample data: European cities
cities <- data.frame(
  name = c("Vienna", "Paris", "Madrid", "Berlin", "Rome"),
  lon = c(16.37, 2.35, -3.70, 13.40, 12.50),
  lat = c(48.21, 48.86, 40.42, 52.52, 41.90)
)

# Assign to ~1000 km² hexagonal cells
result <- hexify(cities, lon = "lon", lat = "lat", area = 1000)
result
```

### With sf Objects

```{r sf-usage, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)

# Create sf object (any CRS works - hexify transforms automatically)
pts <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)

# hexify handles CRS transformation automatically
result_sf <- hexify(pts, area = 1000)
class(result_sf)
```

## Real-World Example: Species Occurrence Data

This example demonstrates the typical workflow: loading point data, assigning to grid cells, aggregating, and visualizing. We'll simulate bird observation data across Europe and Africa.

```{r bird-setup, message=FALSE, warning=FALSE}
library(sf)

# Simulate bird observation data
set.seed(123)
n_obs <- 3000

# Generate observations with realistic spatial clustering
# More observations in temperate regions, fewer in deserts/arctic
birds <- data.frame(
  lon = c(
    rnorm(800, mean = 10, sd = 15),    # Western Europe
    rnorm(600, mean = 25, sd = 10),    # Eastern Europe
    rnorm(400, mean = 20, sd = 20),    # Mediterranean
    rnorm(500, mean = 0, sd = 15),     # West Africa
    rnorm(400, mean = 35, sd = 10),    # East Africa
    rnorm(300, mean = -5, sd = 8)      # Atlantic coast
  ),
  lat = c(
    rnorm(800, mean = 50, sd = 8),     # Western Europe
    rnorm(600, mean = 55, sd = 6),     # Eastern Europe
    rnorm(400, mean = 42, sd = 5),     # Mediterranean
    rnorm(500, mean = 10, sd = 10),    # West Africa
    rnorm(400, mean = -5, sd = 12),    # East Africa
    rnorm(300, mean = 35, sd = 10)     # Atlantic coast
  ),
  species = sample(c("Passer domesticus", "Turdus merula", "Parus major",
                     "Columba palumbus", "Sturnus vulgaris"), n_obs, replace = TRUE)
)

# Clip to valid ranges
birds$lon <- pmax(-30, pmin(60, birds$lon))
birds$lat <- pmax(-35, pmin(70, birds$lat))
```

### Assign Points to Grid Cells

```{r bird-assign}
# Assign each observation to a ~25,000 km² cell
birds_gridded <- hexify(birds, lon = "lon", lat = "lat", area = 25000)

# Count observations per cell
obs_counts <- aggregate(
  species ~ cell_id + cell_cen_lon + cell_cen_lat,
  data = birds_gridded,
  FUN = length
)
names(obs_counts)[4] <- "n_observations"

# Species richness per cell
richness <- aggregate(
  species ~ cell_id,
  data = birds_gridded,
  FUN = function(x) length(unique(x))
)
names(richness)[2] <- "n_species"

obs_counts <- merge(obs_counts, richness, by = "cell_id")
head(obs_counts)
```

### Generate Cell Polygons

```{r bird-polygons, message=FALSE}
# Get grid parameters
grid_info <- hexify_grid(area = 25000, aperture = 3)

# Generate polygons for cells with data
cell_polys <- hexify_cell_to_sf(
  obs_counts$cell_id,
  resolution = grid_info$resolution,
  aperture = 3
)
cell_polys$n_observations <- obs_counts$n_observations
cell_polys$n_species <- obs_counts$n_species
```

### Visualize Results

```{r bird-plot, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
library(ggplot2)

# Get relevant countries
region <- hexify_world[hexify_world$continent %in% c("Europe", "Africa"), ]

ggplot() +
  geom_sf(data = region, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = cell_polys, aes(fill = n_observations), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", name = "Observations", trans = "sqrt") +
  coord_sf(xlim = c(-30, 60), ylim = c(-35, 70)) +
  labs(
    title = "Bird Observations in Equal-Area Hexagonal Cells",
    subtitle = sprintf("ISEA3H grid at resolution %d (~%.0f km² cells)",
                       grid_info$resolution, grid_info$area)
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_line(color = "gray90")
  )
```

Notice how the hexagons appear different sizes on the flat map—that's the projection distortion, not the actual cell sizes. On the sphere, every cell has exactly the same area.

### Species Richness Map

```{r richness-plot, fig.width=8, fig.height=6, message=FALSE}
ggplot() +
  geom_sf(data = region, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = cell_polys, aes(fill = n_species), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "mako", name = "Species\nRichness", direction = -1) +
  coord_sf(xlim = c(-30, 60), ylim = c(-35, 70)) +
  labs(
    title = "Species Richness per Grid Cell",
    subtitle = "Number of unique species observed in each equal-area cell"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_line(color = "gray90")
  )
```

## Random Sampling Earth

Discrete global grids provide two elegant methods for uniform spatial sampling.

### Method 1: Sample Points, Get Cells

Generate uniformly distributed lat-lon points and retrieve their grid cells:

```{r random-sample-method1}
N <- 100

# Uniform points on sphere (see Wolfram MathWorld: Sphere Point Picking)
u <- runif(N)
v <- runif(N)
theta <- 2 * pi * u
phi <- acos(2 * v - 1)
lon <- (theta * 180 / pi) - 180
lat <- (phi * 180 / pi) - 90

sample_df <- data.frame(lon = lon, lat = lat)

# Assign to ~100,000 km² cells
sample_gridded <- hexify(sample_df, lon = "lon", lat = "lat", area = 100000)

# Get unique cells
unique_cells <- unique(sample_gridded$cell_id)
cat(sprintf("Sampled %d unique cells from %d points\n", length(unique_cells), N))
```

### Method 2: Sample Cell IDs Directly

For aperture 3, the maximum cell ID at resolution r is `10 × 3^r + 2`:

```{r random-sample-method2}
# Grid parameters
grid <- hexify_grid(area = 100000, aperture = 3)
max_cell <- 10 * (3^grid$resolution) + 2

cat(sprintf("Resolution %d has %d total cells\n", grid$resolution, max_cell))

# Sample random cell IDs
N <- 100
random_cells <- sample(1:max_cell, N, replace = FALSE)

# Get cell centers
cell_centers <- hexify_cell_to_lonlat(random_cells,
                                       resolution = grid$resolution,
                                       aperture = 3)

head(cell_centers)
```

### Visualize Random Sample

```{r random-sample-plot, fig.width=7, fig.height=4, message=FALSE}
# Generate polygons for sampled cells
sample_polys <- hexify_cell_to_sf(
  random_cells,
  resolution = grid$resolution,
  aperture = 3
)

sample_polys_wrapped <- st_wrap_dateline(
  sample_polys,
  options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180"),
  quiet = TRUE
)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = sample_polys_wrapped, fill = alpha("forestgreen", 0.5),
          color = "darkgreen", linewidth = 0.4) +
  labs(title = sprintf("Random Sample of %d Cells (~%.0f km² each)", N, grid$area)) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Grid Generation

### Grid Over a Rectangular Region

```{r grid-rect, fig.width=7, fig.height=5, message=FALSE}
# Generate a grid covering Western Europe
europe_grid <- hexify_grid_rect(
  minlon = -10, maxlon = 25,
  minlat = 35, maxlat = 60,
  area = 5000
)

# Get European countries for context
europe <- hexify_world[hexify_world$continent == "Europe", ]

ggplot() +
  geom_sf(data = europe, fill = "gray95", color = "gray60") +
  geom_sf(data = europe_grid, fill = NA, color = "steelblue", linewidth = 0.4) +
  coord_sf(xlim = c(-10, 25), ylim = c(35, 60)) +
  labs(title = "Hexagonal Grid over Western Europe (~5,000 km² cells)") +
  theme_minimal()
```

### Grid Over a Polygon (Shapefile)

Clip a grid to any sf polygon boundary:

```{r grid-polygon, fig.width=6, fig.height=6, message=FALSE}
# Get France boundary
france <- hexify_world[hexify_world$name == "France", ]

# Generate grid covering France's bounding box
france_bbox <- st_bbox(france)
france_grid <- hexify_grid_rect(
  minlon = france_bbox["xmin"], maxlon = france_bbox["xmax"],
  minlat = france_bbox["ymin"], maxlat = france_bbox["ymax"],
  area = 2000
)

# Clip grid to France boundary
france_grid_clipped <- st_intersection(france_grid, st_geometry(france))

ggplot() +
  geom_sf(data = france, fill = "gray95", color = "gray40", linewidth = 0.5) +
  geom_sf(data = france_grid_clipped, fill = alpha("steelblue", 0.3),
          color = "steelblue", linewidth = 0.3) +
  labs(title = "Hexagonal Grid Clipped to France (~2,000 km² cells)") +
  theme_minimal()
```

## Export to External Formats

Use sf's `st_write()` to export grids for use in GIS software:

```{r export-formats, eval=FALSE}
library(sf)

# Generate a grid
grid <- hexify_grid_rect(
  minlon = -10, maxlon = 25,
  minlat = 35, maxlat = 60,
  area = 10000
)

# Export to various formats
st_write(grid, "europe_grid.gpkg", layer = "hexgrid")     # GeoPackage
st_write(grid, "europe_grid.shp")                         # Shapefile
st_write(grid, "europe_grid.geojson")                     # GeoJSON
st_write(grid, "europe_grid.kml", layer = "hexgrid")      # KML (Google Earth)
```

## Hierarchical Indexing

hexify supports hierarchical cell indices with parent-child relationships. This enables multi-scale analysis and efficient spatial queries.

### Parent-Child Relationships

```{r hierarchy-demo}
# Get index for a location
idx <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 3)
cat("Index at resolution 5:", idx, "\n")

# Get parent (coarser resolution)
parent_idx <- hexify_get_parent(idx, aperture = 3)
cat("Parent (resolution 4):", parent_idx, "\n")

# Get children (finer resolution)
children <- hexify_get_children(idx, aperture = 3)
cat("Children (resolution 6):", paste(children, collapse = ", "), "\n")
```

### Index Structure

The index encodes the hierarchical path from icosahedron face to cell:

```{r index-structure}
idx <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 3)

cat("Full index:", idx, "\n")
cat("Face (first 2 digits):", substr(idx, 1, 2), "\n")
cat("Path (remaining digits):", substr(idx, 3, nchar(idx)), "\n")
cat("Resolution:", hexify_get_resolution(idx, aperture = 3), "\n")
```

### Multi-Scale Aggregation

```{r multiscale, message=FALSE}
# Sample data
set.seed(42)
obs <- data.frame(
  lon = runif(500, 10, 20),
  lat = runif(500, 45, 55),
  value = rnorm(500, 100, 20)
)

# Aggregate at multiple scales
scales <- c(500, 2000, 10000)  # km²

for (area in scales) {
  gridded <- hexify(obs, lon = "lon", lat = "lat", area = area)
  n_cells <- length(unique(gridded$cell_id))
  mean_per_cell <- nrow(obs) / n_cells
  cat(sprintf("Area %6.0f km²: %3d cells, %.1f obs/cell\n",
              area, n_cells, mean_per_cell))
}
```

## Caveats

### Pentagon Cells

At every resolution, the ISEA grid contains **12 pentagonal cells** with area 5/6 that of hexagons. These are located at the icosahedron vertices:

```{r pentagon-locations, fig.width=7, fig.height=4, message=FALSE}
# Pentagon locations (icosahedron vertices in standard ISEA orientation)
pentagon_coords <- data.frame(
  type = c("Pole", "Pole", rep("Vertex", 10)),
  lon = c(0, 0, seq(0, 324, by = 36)),
  lat = c(90, -90, rep(c(26.57, -26.57), 5))
)

# Assign to grid and get polygons
grid_info <- hexify_grid(area = 500000, aperture = 3)
pentagon_coords$cell_id <- hexify_lonlat_to_cell(
  pentagon_coords$lon, pentagon_coords$lat,
  resolution = grid_info$resolution, aperture = 3
)

pentagon_polys <- hexify_cell_to_sf(
  pentagon_coords$cell_id,
  resolution = grid_info$resolution,
  aperture = 3
)

pentagon_polys_wrapped <- st_wrap_dateline(
  pentagon_polys,
  options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180"),
  quiet = TRUE
)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = pentagon_polys_wrapped, fill = alpha("purple", 0.6),
          color = "purple", linewidth = 0.8) +
  labs(
    title = "Pentagon Cell Locations",
    subtitle = "12 pentagonal cells at icosahedron vertices (area = 5/6 of hexagons)"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

For most analyses, pentagons are not a concern:

1. They're a tiny minority (12 out of millions of cells at high resolutions)
2. They're in predictable locations (poles + 10 evenly-spaced low-latitude points)
3. The area difference (5/6) is small and can be corrected if needed

### Multi-Scale Analysis

Hexagonal grids do **not** nest perfectly—cells at one resolution partially overlap cells at other resolutions. For hierarchical analysis, use the index-based parent/child functions rather than spatial containment.

### Integer Limits

Cell IDs are stored as integers. For resolutions above 15 (aperture 3), cell IDs may exceed R's integer limit (2^31-1). For very high resolutions, use index strings via `hexify_lonlat_to_index()`.

## Resolution Reference

### ISEA3H (Aperture 3)

| Res | Number of Cells | Cell Area (km²) | Mean Spacing (km) |
|----:|----------------:|----------------:|------------------:|
|   0 |              12 |  51,006,562.17  |                   |
|   1 |              32 |  17,002,187.39  |         4,320.49  |
|   2 |              92 |   5,667,395.80  |         2,539.69  |
|   3 |             272 |   1,889,131.93  |         1,480.02  |
|   4 |             812 |     629,710.64  |           855.42  |
|   5 |           2,432 |     209,903.55  |           494.96  |
|   6 |           7,292 |      69,967.85  |           285.65  |
|   7 |          21,872 |      23,322.62  |           165.06  |
|   8 |          65,612 |       7,774.21  |            95.26  |
|   9 |         196,832 |       2,591.40  |            55.02  |
|  10 |         590,492 |         863.80  |            31.76  |
|  11 |       1,771,472 |         287.93  |            18.34  |
|  12 |       5,314,412 |          95.98  |            10.59  |
|  13 |      15,943,232 |          31.99  |             6.11  |
|  14 |      47,829,692 |          10.66  |             3.53  |
|  15 |     143,489,072 |           3.55  |             2.04  |

### Choosing Resolution by Area

```{r resolution-choice}
# Find resolution for target areas
targets <- c(100, 1000, 10000, 100000)

cat("Target Area → Actual Resolution & Area\n")
cat("--------------------------------------\n")
for (target in targets) {
  grid <- hexify_grid(area = target, aperture = 3)
  cat(sprintf("%7.0f km² → res %2d = %9.1f km²\n",
              target, grid$resolution, grid$area))
}
```

### Comparing Apertures

```{r aperture-comparison}
target <- 1000  # km²

cat(sprintf("Target: ~%d km² cells\n\n", target))
for (ap in c(3, 4, 7)) {
  grid <- hexify_grid(area = target, aperture = ap)
  n_cells <- 10 * (ap^grid$resolution) + 2
  cat(sprintf("Aperture %d: res %2d → %8.1f km² (%s cells)\n",
              ap, grid$resolution, grid$area,
              format(n_cells, big.mark = ",")))
}
```

## dggridR Compatibility

hexify produces identical cell assignments to dggridR:

```{r dggridr-compat, eval = FALSE}
library(dggridR)
library(hexify)

# dggridR reference
dggs <- dggridR::dgconstruct(res = 10, aperture = 3)
ref <- dggridR::dgGEO_to_SEQNUM(dggs, cities$lon, cities$lat)

# hexify result (res 10 ≈ 864 km²)
result <- hexify(cities, lon = "lon", lat = "lat", area = 864)

# Verify identical
all(result$cell_id == ref$seqnum)
#> TRUE
```

## Function Reference

### Main Functions

| Function | Description |
|----------|-------------|
| `hexify()` | Assign points to grid cells (main entry point) |
| `hexify_grid()` | Create grid specification |
| `hexify_grid_rect()` | Generate grid polygons over rectangular region |
| `hexify_grid_global()` | Generate global grid polygons |
| `hexify_to_sf()` | Convert hexify result to sf points or polygons |
| `hexify_cell_to_sf()` | Generate polygons from cell IDs |

### Coordinate Conversion

| Function | Description |
|----------|-------------|
| `hexify_lonlat_to_cell()` | lon/lat → cell ID |
| `hexify_cell_to_lonlat()` | cell ID → cell center lon/lat |
| `hexify_lonlat_to_index()` | lon/lat → hierarchical index string |
| `hexify_index_to_lonlat()` | index string → cell center lon/lat |

### Hierarchical Operations

| Function | Description |
|----------|-------------|
| `hexify_get_parent()` | Get parent cell (coarser resolution) |
| `hexify_get_children()` | Get child cells (finer resolution) |
| `hexify_get_resolution()` | Get resolution level from index |

## See Also

- `vignette("theory")` — Mathematical foundations (projection, apertures, space-filling curves)
- `vignette("workflows")` — Detailed workflow examples for common analysis tasks

## Session Info

```{r session-info}
sessionInfo()
```
