---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hexify)
```

## Installation

```{r, eval = FALSE}
# Install from GitHub
remotes::install_github("gcol33/hexify")
```

**Suggested packages** (for extended functionality):

- `sf`: Spatial data handling, CRS transformations
- `dggridR`: Polygon generation, full DGGRID functionality

## What hexify Does

hexify assigns geographic coordinates to equal-area hexagonal grid cells using the ISEA3H (Icosahedral Snyder Equal Area, Aperture 3, Hexagon) discrete global grid system.

Given a data frame with longitude/latitude columns, `hexify()` returns:

- `hex_id`: Unique cell identifier (integer SEQNUM)
- `hex_cen_lon`: Cell center longitude
- `hex_cen_lat`: Cell center latitude

## Quick Examples

### Basic Usage

```{r}
# Sample data: European cities
cities <- data.frame(
  name = c("Vienna", "Paris", "Madrid", "Berlin", "Rome"),
  lon = c(16.37, 2.35, -3.70, 13.40, 12.50),
  lat = c(48.21, 48.86, 40.42, 52.52, 41.90)
)

# Assign to ~1000 km² hexagonal cells
result <- hexify(cities, lon = "lon", lat = "lat", area = 1000)
result
```

### Using Spacing Instead of Area

Specify the long diagonal of hexagons in kilometers:

```{r}
# 50 km spacing (long diagonal)
result <- hexify(cities, lon = "lon", lat = "lat", spacing = 50)
result[, c("name", "hex_id")]
```

### With sf Objects

```{r, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)

# Create sf object (any CRS works)
pts <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)

# hexify handles CRS transformation automatically
result_sf <- hexify(pts, area = 1000)

# sf class is preserved
class(result_sf)
```

## Grid Specification

### Area-Based

Specify target cell area in km²:

```{r}
# Fine grid (~100 km²)
fine <- hexify(cities, lon = "lon", lat = "lat", area = 100)

# Coarse grid (~10,000 km²)
coarse <- hexify(cities, lon = "lon", lat = "lat", area = 10000)

# Compare: more cells at finer resolution
cat("Fine grid cell IDs:", fine$hex_id, "\n")
cat("Coarse grid cell IDs:", coarse$hex_id, "\n")
```

### Spacing-Based

Specify the long diagonal of hexagons in km:

```{r}
# Relationship: area ≈ 0.866 * spacing²
# 50 km spacing → ~2165 km² cells
result <- hexify(cities, lon = "lon", lat = "lat", spacing = 50)
```

## Resolution Reference

| Resolution | Total Cells | Area (km²) | Spacing (km) |
|------------|-------------|------------|--------------|
| 5 | 2,432 | 209,903 | 695 |
| 6 | 7,292 | 69,968 | 401 |
| 7 | 21,872 | 23,323 | 232 |
| 8 | 65,612 | 7,774 | 134 |
| 9 | 196,832 | 2,591 | 77 |
| 10 | 590,492 | 864 | 45 |
| 11 | 1,771,472 | 288 | 26 |
| 12 | 5,314,412 | 96 | 15 |

## dggridR Compatibility

hexify produces identical cell assignments to dggridR:

```{r, eval = FALSE}
library(dggridR)
library(hexify)

# dggridR reference
dggs <- dggridR::dgconstruct(res = 10, aperture = 3)
ref <- dggridR::dgGEO_to_SEQNUM(dggs, cities$lon, cities$lat)

# hexify result (res 10 ≈ 864 km²)
result <- hexify(cities, lon = "lon", lat = "lat", area = 864)

# Verify identical
all(result$hex_id == ref$seqnum)
#> TRUE
```

## Aggregating Data by Grid Cell

Common use case: aggregate point data to grid cells.

```{r}
# Sample occurrence data
occurrences <- data.frame(
  species = c("A", "A", "A", "B", "B", "C"),
  lon = c(16.37, 16.40, 2.35, 2.35, 13.40, 12.50),
  lat = c(48.21, 48.22, 48.86, 48.86, 52.52, 41.90)
)

# Assign to grid
occ_grid <- hexify(occurrences, lon = "lon", lat = "lat", area = 1000)

# Count occurrences per cell
cell_counts <- aggregate(species ~ hex_id, data = occ_grid, FUN = length)
names(cell_counts)[2] <- "n_occurrences"
cell_counts
```

### Species Richness per Cell

```{r}
# Count unique species per cell
richness <- aggregate(species ~ hex_id, data = occ_grid,
                      FUN = function(x) length(unique(x)))
names(richness)[2] <- "n_species"
richness
```

## Function Reference

### hexify()
```r
hexify(data, lon = "lon", lat = "lat", area = NULL, spacing = NULL,
       aperture = 3L, resround = "nearest")
```

| Parameter | Description | Default |
|-----------|-------------|---------|
| `data` | data.frame or sf object | *required* |
| `lon` | Longitude column name (ignored for sf) | `"lon"` |
| `lat` | Latitude column name (ignored for sf) | `"lat"` |
| `area` | Target cell area in km² | `NULL` |
| `spacing` | Long diagonal in km | `NULL` |
| `aperture` | Grid aperture (only 3 supported) | `3L` |
| `resround` | Resolution rounding: `"nearest"`, `"up"`, `"down"` | `"nearest"` |

**Returns**: Input data with added columns `hex_id`, `hex_cen_lon`, `hex_cen_lat`.

## Troubleshooting

**"Column 'lon' not found" error**

- Check column names match `lon` and `lat` parameters
- For sf objects, coordinates are extracted automatically

**Different cell IDs than dggridR**

- Ensure same aperture (hexify currently only supports aperture 3)
- Match resolution via area: use `dgearthstat(dggs)$area_km` from dggridR

**Very large hex_id values**

- Normal for high resolutions; R stores as integer up to 2^31-1
- For res > 15, consider using string indices via `hexify_lonlat_to_index()`

## See Also

- `vignette("theory")` - Mathematical foundations
- `vignette("workflows")` - Real-world analysis examples

## Session Info

```{r}
sessionInfo()
```
