---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hexify)
```

## Spatial Analysis Done Right

You want to do spatial statistics, and it involves binning points into grid cells.

**The problem with rectangular grids**: Using a rectangular grid (like a standard lat-long grid) introduces distortions. At the equator, a 1 degree cell covers about 12,300 km2. Near the poles, the same 1 degree cell covers a tiny fraction of that area. This breaks any analysis that assumes equal sampling effort or comparable cell sizes.

**The solution**: Discrete global grids partition Earths surface into cells of **equal area**, regardless of latitude. hexify implements the ISEA (Icosahedral Snyder Equal Area) projection, providing hexagonal cells that are all the same size from the equator to the Arctic.

### Why Equal-Area Matters

```{r equal-area-demo, message=FALSE}
# Same-sized cells at different latitudes
test_points <- data.frame(
  location = c("Equator", "Mid-latitude", "Arctic"),
  lon = c(0, 0, 0),
  lat = c(0, 45, 70)
)

result <- hexify(test_points, lon = "lon", lat = "lat", area = 1000)

# All cells have the same area, regardless of latitude
result[, c("location", "lat", "hex_area")]
```

With hexify, a 1000 km2 cell at the equator is the same size as a 1000 km2 cell in Norway.


## Installation

```{r, eval = FALSE}
# Install from GitHub
remotes::install_github("gcol33/hexify")
```

**Suggested packages** (for extended functionality):

- `sf`: Spatial data handling, CRS transformations
- `dggridR`: Polygon generation, full DGGRID functionality

## What hexify Does

hexify assigns geographic coordinates to equal-area hexagonal grid cells using the ISEA (Icosahedral Snyder Equal Area) discrete global grid system. Supports apertures 3, 4, 7, and mixed 4/3:

- **Aperture 3** (ISEA3H): Default, compatible with dggridR
- **Aperture 4** (ISEA4H): Faster cell count growth per resolution
- **Aperture 7** (ISEA7H): Densest grid per resolution level
- **Aperture "4/3"** (ISEA43H): Mixed aperture (4 then 3)

Given a data frame with longitude/latitude columns, `hexify()` returns:

- `hex_id`: Unique cell identifier (integer cell ID)
- `hex_cen_lon`: Cell center longitude
- `hex_cen_lat`: Cell center latitude
- `hex_area`: Actual cell area in km²
- `hex_diag`: Actual cell diagonal in km

## Quick Examples

### Basic Usage

```{r}
# Sample data: European cities
cities <- data.frame(
  name = c("Vienna", "Paris", "Madrid", "Berlin", "Rome"),
  lon = c(16.37, 2.35, -3.70, 13.40, 12.50),
  lat = c(48.21, 48.86, 40.42, 52.52, 41.90)
)

# Assign to ~1000 km² hexagonal cells
result <- hexify(cities, lon = "lon", lat = "lat", area = 1000)
result
```

### Using Diagonal Instead of Area

Specify the long diagonal of hexagons in kilometers:

```{r}
# 50 km diagonal (long diagonal)
result <- hexify(cities, lon = "lon", lat = "lat", diagonal = 50)
result[, c("name", "hex_id")]
```

### With sf Objects

```{r, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)

# Create sf object (any CRS works)
pts <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)

# hexify handles CRS transformation automatically
result_sf <- hexify(pts, area = 1000)

# sf class is preserved
class(result_sf)
```

## Grid Specification

### Area-Based

Specify target cell area in km²:

```{r}
# Fine grid (~100 km²)
fine <- hexify(cities, lon = "lon", lat = "lat", area = 100)

# Coarse grid (~10,000 km²)
coarse <- hexify(cities, lon = "lon", lat = "lat", area = 10000)

# Compare: more cells at finer resolution
cat("Fine grid cell IDs:", fine$hex_id, "\n")
cat("Coarse grid cell IDs:", coarse$hex_id, "\n")
```

### Diagonal-Based

Specify the long diagonal of hexagons in km:

```{r}
# Relationship: area ≈ 0.866 * diagonal²
# 50 km diagonal → ~2165 km² cells
result <- hexify(cities, lon = "lon", lat = "lat", diagonal = 50)
```

## Aperture Options

| Aperture | Grid Type | Cell Count Formula | Use Case |
|----------|-----------|-------------------|----------|
| 3 | ISEA3H | 10 × 3^res + 2 | Default, compatible with dggridR |
| 4 | ISEA4H | 10 × 4^res + 2 | Faster cell count growth |
| 7 | ISEA7H | 10 × 7^res + 2 | Densest grid per resolution |
| "4/3" | ISEA43H | 10 × 4^m × 3^(res-m) + 2 | Mixed aperture |

```{r}
# Using different apertures
result_ap3 <- hexify(cities, lon = "lon", lat = "lat", area = 1000, aperture = 3)
result_ap4 <- hexify(cities, lon = "lon", lat = "lat", area = 1000, aperture = 4)
result_ap7 <- hexify(cities, lon = "lon", lat = "lat", area = 1000, aperture = 7)

# Compare cell IDs (different numbering systems)
data.frame(
  city = cities$name,
  ap3 = result_ap3$hex_id,
  ap4 = result_ap4$hex_id,
  ap7 = result_ap7$hex_id
)
```

### Mixed Aperture (ISEA43H)

Mixed aperture uses aperture 4 for the first levels, then switches to aperture 3:

```{r}
# Mixed aperture grid
result_mixed <- hexify(cities, lon = "lon", lat = "lat", area = 1000, aperture = "4/3")
result_mixed[, c("name", "hex_id", "hex_area")]
```

## Resolution Reference (ISEA3H)

The following table shows the number of cells, their area, and spacing statistics for the ISEA3H grid type (aperture 3):

| Res | Number of Cells | Cell Area (km2) | Mean Spacing (km) |
|----:|----------------:|----------------:|------------------:|
|   0 |              12 |  51,006,562.17  |                   |
|   1 |              32 |  17,002,187.39  |         4,320.49  |
|   2 |              92 |   5,667,395.80  |         2,539.69  |
|   3 |             272 |   1,889,131.93  |         1,480.02  |
|   4 |             812 |     629,710.64  |           855.42  |
|   5 |           2,432 |     209,903.55  |           494.96  |
|   6 |           7,292 |      69,967.85  |           285.65  |
|   7 |          21,872 |      23,322.62  |           165.06  |
|   8 |          65,612 |       7,774.21  |            95.26  |
|   9 |         196,832 |       2,591.40  |            55.02  |
|  10 |         590,492 |         863.80  |            31.76  |
|  11 |       1,771,472 |         287.93  |            18.34  |
|  12 |       5,314,412 |          95.98  |            10.59  |
|  13 |      15,943,232 |          31.99  |             6.11  |
|  14 |      47,829,692 |          10.66  |             3.53  |
|  15 |     143,489,072 |           3.55  |             2.04  |

## Caveats

### Pentagon Cells

At every resolution, the ISEA grid contains **12 pentagonal cells** which each have an area exactly 5/6 that of the hexagonal cells. In the standard orientation, these are located at:

- North Pole (90N)
- South Pole (90S)
- 10 cells at latitudes +/-26.57 degrees spaced evenly around the globe

For most analyses, this is not a concern:

1. Pentagons are a tiny minority (12 out of millions of cells at high resolutions)
2. They are positioned in predictable, often low-data locations
3. The area difference (5/6) is small and can be corrected if needed

### Multi-Scale Analysis

In hexagonal grids, cells from one resolution level are only **partially contained** by cells of other levels (unlike square grids). This means direct hierarchical aggregation is not straightforward. For nested analyses, consider using the same resolution throughout or carefully handling partial overlaps.

### Integer Limits

Cell IDs are stored as integers. For resolutions above 15, cell IDs may exceed Rs integer limit (2^31-1). For very high resolutions, use the index-based functions that return string representations.


## dggridR Compatibility

hexify produces identical cell assignments to dggridR:

```{r, eval = FALSE}
library(dggridR)
library(hexify)

# dggridR reference
dggs <- dggridR::dgconstruct(res = 10, aperture = 3)
ref <- dggridR::dgGEO_to_SEQNUM(dggs, cities$lon, cities$lat)

# hexify result (res 10 ≈ 864 km²)
result <- hexify(cities, lon = "lon", lat = "lat", area = 864)

# Verify identical
all(result$hex_id == ref$seqnum)
#> TRUE
```

## Aggregating Data by Grid Cell

Common use case: aggregate point data to grid cells.

```{r}
# Sample occurrence data
occurrences <- data.frame(
  species = c("A", "A", "A", "B", "B", "C"),
  lon = c(16.37, 16.40, 2.35, 2.35, 13.40, 12.50),
  lat = c(48.21, 48.22, 48.86, 48.86, 52.52, 41.90)
)

# Assign to grid
occ_grid <- hexify(occurrences, lon = "lon", lat = "lat", area = 1000)

# Count occurrences per cell
cell_counts <- aggregate(species ~ hex_id, data = occ_grid, FUN = length)
names(cell_counts)[2] <- "n_occurrences"
cell_counts
```

### Species Richness per Cell

```{r}
# Count unique species per cell
richness <- aggregate(species ~ hex_id, data = occ_grid,
                      FUN = function(x) length(unique(x)))
names(richness)[2] <- "n_species"
richness
```

## Polygon Generation

Generate hexagon polygons for visualization and spatial analysis:

### Basic Polygon Generation

```{r, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)

# Get hex_ids from hexify result
result <- hexify(cities, lon = "lon", lat = "lat", area = 5000)

# Generate polygons (requires resolution and aperture)
polys <- hexify_cell_to_sf(result$hex_id, resolution = 8, aperture = 3)
polys
```

### Plotting Hexagons

```{r, eval = requireNamespace("sf", quietly = TRUE), fig.width=7, fig.height=5}
# Plot the hexagons
plot(st_geometry(polys), col = "lightblue", border = "darkblue",
     main = "European Cities - Hexagonal Grid")
```

### Grid Over a Region

```{r, eval = requireNamespace("sf", quietly = TRUE), fig.width=7, fig.height=5}
# Generate a grid covering Western Europe
europe_grid <- hexify_grid_rect(
  minlon = -5, maxlon = 20,
  minlat = 40, maxlat = 55,
  area = 10000
)

plot(st_geometry(europe_grid), border = "gray",
     main = "Hexagonal Grid over Western Europe")
```


### Visualization with ggplot2

```{r ggplot-example, eval = requireNamespace("sf", quietly = TRUE) && requireNamespace("ggplot2", quietly = TRUE), fig.width=7, fig.height=5}
library(ggplot2)
library(sf)

# Create sample data with values
set.seed(42)
sample_data <- data.frame(
  lon = runif(100, -5, 20),
  lat = runif(100, 40, 55),
  value = rnorm(100, 50, 15)
)

# Assign to grid and aggregate
gridded <- hexify(sample_data, lon = "lon", lat = "lat", area = 5000)
cell_means <- aggregate(value ~ hex_id + hex_cen_lon + hex_cen_lat,
                        data = gridded, FUN = mean)

# Get grid info for polygons
grid_info <- hexify_grid(area = 5000, aperture = 3)

# Generate polygons for cells with data
polys <- hexify_cell_to_sf(cell_means[["hex_id"]],
                           resolution = grid_info[["resolution"]],
                           aperture = 3)
polys[["value"]] <- cell_means[["value"]]

# Plot with ggplot2
ggplot() +
  geom_sf(data = polys, aes(fill = value), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", name = "Mean Value") +
  labs(title = "Aggregated Values in Equal-Area Hexagonal Cells",
       subtitle = "ISEA3H grid at ~5000 km2 resolution") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Function Reference

### hexify()
```r
hexify(data, lon = "lon", lat = "lat", area = NULL, diagonal = NULL,
       aperture = 3L, mixed_aperture_level = NULL, resround = "nearest")
```

| Parameter | Description | Default |
|-----------|-------------|---------|
| `data` | data.frame or sf object | *required* |
| `lon` | Longitude column name (ignored for sf) | `"lon"` |
| `lat` | Latitude column name (ignored for sf) | `"lat"` |
| `area` | Target cell area in km² | `NULL` |
| `diagonal` | Long diagonal in km | `NULL` |
| `aperture` | Grid aperture: 3, 4, 7, or "4/3" | `3L` |
| `mixed_aperture_level` | For "4/3": number of aperture-4 levels | `NULL` (auto) |
| `resround` | Resolution rounding: `"nearest"`, `"up"`, `"down"` | `"nearest"` |

**Returns**: Input data with added columns `hex_id`, `hex_cen_lon`, `hex_cen_lat`, `hex_area`, `hex_diag`.

### hexify_cell_to_sf()
```r
hexify_cell_to_sf(hex_id, resolution, aperture = 3L, return_sf = TRUE)
```

| Parameter | Description | Default |
|-----------|-------------|---------|
| `hex_id` | Integer vector of cell IDs | *required* |
| `resolution` | Grid resolution level | *required* |
| `aperture` | Grid aperture: 3, 4, or 7 | `3L` |
| `return_sf` | Return sf object (TRUE) or data frame (FALSE) | `TRUE` |

**Returns**: sf object with POLYGON geometries, or data frame with vertex coordinates.

### hexify_grid_rect()
```r
hexify_grid_rect(minlon, maxlon, minlat, maxlat, area, aperture = 3L)
```

Generates a grid of hexagon polygons covering a rectangular region.

## Troubleshooting

**"Column 'lon' not found" error**

- Check column names match `lon` and `lat` parameters
- For sf objects, coordinates are extracted automatically

**Different cell IDs than dggridR**

- Ensure same aperture (hexify supports apertures 3, 4, 7, and "4/3")
- Match resolution via area: use `dgearthstat(dggs)$area_km` from dggridR

**Very large hex_id values**

- Normal for high resolutions; R stores as integer up to 2^31-1
- For res > 15, consider using string indices via `hexify_lonlat_to_index()`

## See Also

- `vignette("theory")` - Mathematical foundations
- `vignette("workflows")` - Real-world analysis examples

## Session Info

```{r}
sessionInfo()
```
