---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hexify)
```

```{r cover-image, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4, fig.align='center'}
# Generate cover image: multi-resolution grids over different regions
library(sf)
library(ggplot2)

# Get world basemap
countries <- hexify_world

# Generate grids at different resolutions for specific regions
# Coarse grid: South America
grid_sa <- hex_grid(area_km2 = 200000)
sa_hexes <- grid_rect(c(-80, -40, -35, 10), grid_sa)

# Medium grid: Europe/Africa
grid_eu <- hex_grid(area_km2 = 30000)
eu_hexes <- grid_rect(c(-10, -10, 40, 50), grid_eu)

# Fine grid: East Asia
grid_asia <- hex_grid(area_km2 = 8000)
asia_hexes <- grid_rect(c(90, 10, 140, 50), grid_asia)

ggplot() +
  geom_sf(data = countries, fill = "gray95", color = "gray80", linewidth = 0.15) +
  geom_sf(data = sa_hexes, fill = NA, color = "#1B9E77", linewidth = 0.5) +
  geom_sf(data = eu_hexes, fill = NA, color = "#D95F02", linewidth = 0.4) +
  geom_sf(data = asia_hexes, fill = NA, color = "#7570B3", linewidth = 0.3) +
  coord_sf(xlim = c(-100, 150), ylim = c(-50, 60)) +
  theme_void() +
  theme(panel.background = element_rect(fill = "white", color = NA))
```

# Spatial Analysis Done Right

You want to do spatial statistics, and it involves binning points into grid cells.

**The problem with rectangular grids**: A rectangular lat-lon grid introduces severe distortions. At the equator, a 1 degree cell covers ~12,300 km squared. Near the poles, the same 1 degree cell covers a tiny fraction of that area. This breaks any analysis that assumes equal sampling effort or comparable cell sizes.

**The solution**: Discrete global grids partition Earth's surface into cells of **equal area**, regardless of latitude. hexify implements the ISEA (Icosahedral Snyder Equal Area) projection, providing hexagonal cells that are all the same size from the equator to the Arctic.

### Why Equal-Area Matters

```{r equal-area-demo, message=FALSE}
# Same-sized cells at different latitudes
test_points <- data.frame(
  location = c("Equator", "Mid-latitude", "Arctic"),
  lon = c(0, 0, 0),
  lat = c(0, 45, 70)
)

grid <- hex_grid(area_km2 = 1000)
result <- hexify(test_points, lon = "lon", lat = "lat", grid = grid)

# All cells have the same area, regardless of latitude
cat(sprintf("Grid area: %.1f km2\n", grid@area_km2))
cat(sprintf("Points assigned to %d unique cells\n", n_cells(result)))
```

With hexify, a 1000 km squared cell at the equator is the same size as a 1000 km squared cell in Norway.


## Installation

```{r, eval = FALSE}
# Install from GitHub
remotes::install_github("gcol33/hexify")
```

**Required packages**: `sf` (for spatial operations)

## Core Concepts: HexGridInfo and HexData

hexify uses two S4 classes to make spatial workflows clean and error-free:

- **`HexGridInfo`**: A grid specification that stores all parameters (aperture, resolution, area). Define once, reuse everywhere.
- **`HexData`**: Your data + the grid that was used. Carries the grid reference so downstream operations "just work."

### HexGridInfo Slots

| Slot | Type | Description |
|------|------|-------------|
| `aperture` | character | Grid aperture ("3", "4", "7", or "4/3") |
| `resolution` | integer | Resolution level (0-30) |
| `area_km2` | numeric | Cell area in km squared |
| `diagonal_km` | numeric | Cell diagonal in km |
| `crs` | integer | Coordinate reference system (EPSG code) |

### HexData Slots

| Slot | Type | Description |
|------|------|-------------|
| `data` | data.frame/sf | Your original data (unchanged) |
| `grid` | HexGridInfo | The grid specification used |
| `cell_id` | numeric | Cell ID for each row |
| `cell_center` | matrix | Cell center coordinates (lon, lat) |

## Quick Start

### Basic Usage

```{r basic-usage}
# Sample data: European cities
cities <- data.frame(
  name = c("Vienna", "Paris", "Madrid", "Berlin", "Rome"),
  lon = c(16.37, 2.35, -3.70, 13.40, 12.50),
  lat = c(48.21, 48.86, 40.42, 52.52, 41.90)
)

# Create a grid specification
grid <- hex_grid(area_km2 = 10000)
grid

# Assign cities to hexagonal cells
result <- hexify(cities, lon = "lon", lat = "lat", grid = grid)
result
```

### Accessing HexData

```{r hexdata-access}
# Get the grid specification
grid_info(result)

# Get unique cell IDs
cells(result)

# Count unique cells
n_cells(result)

# Access all cell IDs (one per row)
result@cell_id

# Access cell centers
head(result@cell_center)

# Extract original data as data.frame
head(as.data.frame(result))
```

### Define Grid Once, Reuse Everywhere

The key workflow: create a `HexGridInfo` once, then apply it to multiple datasets.

```{r grid-reuse}
# Define grid specification
my_grid <- hex_grid(area_km2 = 5000, aperture = 3)
my_grid

# Apply same grid to different datasets
set.seed(123)
birds <- data.frame(lon = runif(100, -10, 30), lat = runif(100, 35, 60))
mammals <- data.frame(x = runif(50, -10, 30), y = runif(50, 35, 60))

birds_hex <- hexify(birds, lon = "lon", lat = "lat", grid = my_grid)
mammals_hex <- hexify(mammals, lon = "x", lat = "y", grid = my_grid)

# Both use identical grid - safe to combine
cat("Birds:", n_cells(birds_hex), "cells\n")
cat("Mammals:", n_cells(mammals_hex), "cells\n")
```

### With sf Objects

```{r sf-usage, eval = requireNamespace("sf", quietly = TRUE)}
library(sf)

# Create sf object (any CRS works - hexify transforms automatically)
pts <- st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)

# hexify handles CRS transformation automatically
result_sf <- hexify(pts, area_km2 = 10000)
class(result_sf)
```

## Visualization

hexify provides multiple ways to visualize your gridded data.

### Quick Plot with Base R

The simplest approach uses the built-in `plot()` method:

```{r plot-basic, fig.width=7, fig.height=5}
# Basic plot with world basemap
plot(result, basemap = TRUE, main = "European Cities")
```

### Customize Plot Appearance

```{r plot-custom, fig.width=7, fig.height=5}
# Custom colors and styling
plot(result,
     basemap = TRUE,
     grid_fill = "steelblue",
     grid_border = "darkblue",
     grid_alpha = 0.6,
     basemap_fill = "ivory",
     basemap_border = "gray50",
     main = "Custom Styling")
```

### Show Original Points

```{r plot-points, fig.width=7, fig.height=5}
# Show original points overlaid on cells
plot(result,
     basemap = TRUE,
     show_points = TRUE,
     point_color = "red",
     point_size = 1.5,
     main = "Cities with Cell Overlay")
```

### ggplot2 with hexify_ggplot()

For ggplot2 users, `hexify_ggplot()` returns a ggplot object:

```{r ggplot-basic, fig.width=7, fig.height=5, message=FALSE}
library(ggplot2)

# Basic ggplot
hexify_ggplot(result, basemap = TRUE, title = "European Cities")
```

### Heatmaps with hexify_heatmap()

For choropleth-style visualizations with aggregated data:

```{r heatmap-demo, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
# Simulate observation data
set.seed(42)
n_obs <- 500
obs_data <- data.frame(
  lon = c(rnorm(300, 10, 8), rnorm(200, 0, 10)),
  lat = c(rnorm(300, 48, 5), rnorm(200, 52, 6)),
  count = rpois(n_obs, lambda = 50)
)

# Hexify and aggregate
grid <- hex_grid(area_km2 = 5000)
obs_hex <- hexify(obs_data, lon = "lon", lat = "lat", grid = grid)

# Aggregate counts per cell
obs_df <- as.data.frame(obs_hex)
obs_df$cell_id <- obs_hex@cell_id
cell_totals <- aggregate(count ~ cell_id, data = obs_df, FUN = sum)

# Create heatmap
hexify_heatmap(
  obs_hex,
  value = "count",
  basemap = "world",
  colors = "YlOrRd",
  title = "Observation Counts",
  legend_title = "Count",
  xlim = c(-20, 35),
  ylim = c(35, 65)
)
```

### Simple Map with hexify_map()

For quick base R visualization:

```{r hexify-map, fig.width=7, fig.height=5}
# Quick map visualization
hexify_map(
  result,
  basemap = "world",
  fill = "steelblue",
  border = "darkblue",
  main = "European Cities Grid"
)
```

### World Map Helper

```{r plot-world, fig.width=7, fig.height=4}
# Quick world map
plot_world(fill = "lightgray", border = "gray50")
```

## Grid Generation

### Grid Over a Rectangular Region

```{r grid-rect, fig.width=7, fig.height=5, message=FALSE}
library(sf)

# Create grid specification
grid <- hex_grid(area_km2 = 5000)

# Generate hexagons over Western Europe
europe_hexes <- grid_rect(c(-10, 35, 25, 60), grid)

# Get European countries for context
europe <- hexify_world[hexify_world$continent == "Europe", ]

ggplot() +
  geom_sf(data = europe, fill = "gray95", color = "gray60") +
  geom_sf(data = europe_hexes, fill = NA, color = "steelblue", linewidth = 0.4) +
  coord_sf(xlim = c(-10, 25), ylim = c(35, 60)) +
  labs(title = sprintf("Hexagonal Grid (~%.0f km2 cells)", grid@area_km2)) +
  theme_minimal()
```

### Grid Over a Polygon (Shapefile)

Clip a grid to any sf polygon boundary:

```{r grid-polygon, fig.width=6, fig.height=6, message=FALSE}
# Get France boundary
france <- hexify_world[hexify_world$name == "France", ]

# Generate grid covering mainland France
grid <- hex_grid(area_km2 = 2000)
france_grid <- grid_rect(c(-5, 41, 10, 52), grid)

# Clip grid to France boundary
france_grid_clipped <- st_intersection(france_grid, st_geometry(france))

ggplot() +
  geom_sf(data = france, fill = "gray95", color = "gray40", linewidth = 0.5) +
  geom_sf(data = france_grid_clipped, fill = alpha("steelblue", 0.3),
          color = "steelblue", linewidth = 0.3) +
  coord_sf(xlim = c(-5, 10), ylim = c(41, 52)) +
  labs(title = sprintf("Hexagonal Grid Clipped to France (~%.0f km2 cells)", grid@area_km2)) +
  theme_minimal()
```

### Global Grid

```{r grid-global, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
# Coarse global grid (be careful with fine grids - many cells!)
grid <- hex_grid(area_km2 = 500000)
global_hexes <- grid_global(grid)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray90", color = "gray70", linewidth = 0.2) +
  geom_sf(data = global_hexes, fill = NA, color = "darkgreen", linewidth = 0.3) +
  labs(title = sprintf("Global Hexagonal Grid (~%.0f km2 cells)", grid@area_km2)) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Real-World Example: Species Occurrence Data

This example demonstrates the typical workflow: loading point data, assigning to grid cells, aggregating, and visualizing.

```{r bird-setup, message=FALSE, warning=FALSE}
library(sf)

# Simulate bird observation data
set.seed(123)
n_obs <- 3000

# Generate observations with realistic spatial clustering
birds <- data.frame(
  lon = c(
    rnorm(800, mean = 10, sd = 15),    # Western Europe
    rnorm(600, mean = 25, sd = 10),    # Eastern Europe
    rnorm(400, mean = 20, sd = 20),    # Mediterranean
    rnorm(500, mean = 0, sd = 15),     # West Africa
    rnorm(400, mean = 35, sd = 10),    # East Africa
    rnorm(300, mean = -5, sd = 8)      # Atlantic coast
  ),
  lat = c(
    rnorm(800, mean = 50, sd = 8),     # Western Europe
    rnorm(600, mean = 55, sd = 6),     # Eastern Europe
    rnorm(400, mean = 42, sd = 5),     # Mediterranean
    rnorm(500, mean = 10, sd = 10),    # West Africa
    rnorm(400, mean = -5, sd = 12),    # East Africa
    rnorm(300, mean = 35, sd = 10)     # Atlantic coast
  ),
  species = sample(c("Passer domesticus", "Turdus merula", "Parus major",
                     "Columba palumbus", "Sturnus vulgaris"), n_obs, replace = TRUE)
)

# Clip to valid ranges
birds$lon <- pmax(-30, pmin(60, birds$lon))
birds$lat <- pmax(-35, pmin(70, birds$lat))
```

### Assign Points to Grid Cells

```{r bird-assign}
# Create grid and assign observations
grid <- hex_grid(area_km2 = 25000)
birds_hex <- hexify(birds, lon = "lon", lat = "lat", grid = grid)

# Extract data with cell IDs
birds_gridded <- as.data.frame(birds_hex)
birds_gridded$cell_id <- birds_hex@cell_id

# Count observations per cell
obs_counts <- aggregate(
  species ~ cell_id,
  data = birds_gridded,
  FUN = length
)
names(obs_counts)[2] <- "n_observations"

# Species richness per cell
richness <- aggregate(
  species ~ cell_id,
  data = birds_gridded,
  FUN = function(x) length(unique(x))
)
names(richness)[2] <- "n_species"

obs_counts <- merge(obs_counts, richness, by = "cell_id")
head(obs_counts)
```

### Generate Cell Polygons and Visualize

```{r bird-plot, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
library(ggplot2)

# Generate polygons for cells with data
cell_polys <- cell_to_sf(obs_counts$cell_id, grid)
cell_polys <- merge(cell_polys, obs_counts, by = "cell_id")

# Get relevant countries
region <- hexify_world[hexify_world$continent %in% c("Europe", "Africa"), ]

ggplot() +
  geom_sf(data = region, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = cell_polys, aes(fill = n_observations), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "plasma", name = "Observations", trans = "sqrt") +
  coord_sf(xlim = c(-30, 60), ylim = c(-35, 70)) +
  labs(
    title = "Bird Observations in Equal-Area Hexagonal Cells",
    subtitle = sprintf("ISEA3H grid at resolution %d (~%.0f km2 cells)",
                       grid@resolution, grid@area_km2)
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_line(color = "gray90")
  )
```

### Species Richness Map

```{r richness-plot, fig.width=8, fig.height=6, message=FALSE}
ggplot() +
  geom_sf(data = region, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = cell_polys, aes(fill = n_species), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "mako", name = "Species\nRichness", direction = -1) +
  coord_sf(xlim = c(-30, 60), ylim = c(-35, 70)) +
  labs(
    title = "Species Richness per Grid Cell",
    subtitle = "Number of unique species observed in each equal-area cell"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_line(color = "gray90")
  )
```

## Random Sampling Earth

Discrete global grids provide elegant methods for uniform spatial sampling.

### Sample Cell IDs Directly

```{r random-sample}
# Grid parameters
grid <- hex_grid(area_km2 = 100000, aperture = 3)
max_cell <- 10 * (3^grid@resolution) + 2

cat(sprintf("Resolution %d has %d total cells\n", grid@resolution, max_cell))

# Sample random cell IDs
N <- 100
random_cells <- sample(1:max_cell, N, replace = FALSE)

# Get cell centers
cell_centers <- cell_to_lonlat(random_cells, grid)
head(cell_centers)
```

### Visualize Random Sample

```{r random-sample-plot, fig.width=7, fig.height=4, message=FALSE}
# Generate polygons for sampled cells
sample_polys <- cell_to_sf(random_cells, grid)

sample_polys_wrapped <- st_wrap_dateline(
  sample_polys,
  options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180"),
  quiet = TRUE
)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = sample_polys_wrapped, fill = alpha("forestgreen", 0.5),
          color = "darkgreen", linewidth = 0.4) +
  labs(title = sprintf("Random Sample of %d Cells (~%.0f km2 each)", N, grid@area_km2)) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Export to External Formats

Use sf's `st_write()` to export grids for use in GIS software:

```{r export-formats, eval=FALSE}
library(sf)

# Generate a grid
grid <- hex_grid(area_km2 = 10000)
europe <- grid_rect(c(-10, 35, 25, 60), grid)

# Export to various formats
st_write(europe, "europe_grid.gpkg", layer = "hexgrid")     # GeoPackage
st_write(europe, "europe_grid.shp")                         # Shapefile
st_write(europe, "europe_grid.geojson")                     # GeoJSON
st_write(europe, "europe_grid.kml", layer = "hexgrid")      # KML (Google Earth)
```

## Caveats

### Pentagon Cells

At every resolution, the ISEA grid contains **12 pentagonal cells** with area 5/6 that of hexagons. These are located at the icosahedron vertices:

```{r pentagon-locations, fig.width=7, fig.height=4, message=FALSE}
# Pentagon locations (icosahedron vertices in standard ISEA orientation)
pentagon_coords <- data.frame(
  type = c("Pole", "Pole", rep("Vertex", 10)),
  lon = c(0, 0, seq(0, 324, by = 36)),
  lat = c(90, -90, rep(c(26.57, -26.57), 5))
)

# Assign to grid and get polygons
grid <- hex_grid(area_km2 = 500000)
pentagon_cells <- lonlat_to_cell(pentagon_coords$lon, pentagon_coords$lat, grid)

pentagon_polys <- cell_to_sf(pentagon_cells, grid)
pentagon_polys_wrapped <- st_wrap_dateline(
  pentagon_polys,
  options = c("WRAPDATELINE=YES", "DATELINEOFFSET=180"),
  quiet = TRUE
)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = pentagon_polys_wrapped, fill = alpha("purple", 0.6),
          color = "purple", linewidth = 0.8) +
  labs(
    title = "Pentagon Cell Locations",
    subtitle = "12 pentagonal cells at icosahedron vertices (area = 5/6 of hexagons)"
  ) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

For most analyses, pentagons are not a concern:

1. They're a tiny minority (12 out of millions of cells at high resolutions)
2. They're in predictable locations (poles + 10 evenly-spaced low-latitude points)
3. The area difference (5/6) is small and can be corrected if needed

### Multi-Scale Analysis

Hexagonal grids do **not** nest perfectly - cells at one resolution partially overlap cells at other resolutions. For hierarchical analysis, use grid-based aggregation rather than spatial containment.

### Integer Limits

Cell IDs are stored as integers. For resolutions above 15 (aperture 3), cell IDs may exceed R's integer limit (2^31-1). Use appropriate numeric types for very high resolutions.

## Resolution Reference

### ISEA3H (Aperture 3)

```{r resolution-table, echo=FALSE}
comparison <- hexify_compare_resolutions(aperture = 3, res_range = 0:15)
comparison$n_cells_fmt <- ifelse(
  comparison$n_cells > 1e6,
  sprintf("%.1fM", comparison$n_cells / 1e6),
  ifelse(comparison$n_cells > 1e3,
         sprintf("%.1fK", comparison$n_cells / 1e3),
         as.character(comparison$n_cells))
)
knitr::kable(
  comparison[, c("resolution", "n_cells_fmt", "cell_area_km2", "cell_spacing_km")],
  col.names = c("Resolution", "# Cells", "Cell Area (km2)", "Spacing (km)"),
  digits = 1
)
```

### Choosing Resolution by Area

```{r resolution-choice}
# Find resolution for target areas
targets <- c(100, 1000, 10000, 100000)

cat("Target Area -> Actual Resolution & Area\n")
cat("--------------------------------------\n")
for (target in targets) {
  grid <- hex_grid(area_km2 = target, aperture = 3)
  cat(sprintf("%7.0f km2 -> res %2d = %9.1f km2\n",
              target, grid@resolution, grid@area_km2))
}
```

### Comparing Apertures

```{r aperture-comparison}
target <- 1000  # km2

cat(sprintf("Target: ~%d km2 cells\n\n", target))
for (ap in c(3, 4, 7)) {
  grid <- hex_grid(area_km2 = target, aperture = ap)
  n_cells <- 10 * (ap^grid@resolution) + 2
  cat(sprintf("Aperture %d: res %2d -> %8.1f km2 (%s cells)\n",
              ap, grid@resolution, grid@area_km2,
              format(n_cells, big.mark = ",")))
}
```

## dggridR Compatibility

hexify produces identical cell assignments to dggridR:

```{r dggridr-compat, eval = FALSE}
library(dggridR)
library(hexify)

# dggridR reference
dggs <- dggridR::dgconstruct(res = 10, aperture = 3)
ref <- dggridR::dgGEO_to_SEQNUM(dggs, cities$lon, cities$lat)

# hexify result (res 10 = 864 km2)
grid <- hex_grid(resolution = 10, aperture = 3)
result <- hexify(cities, lon = "lon", lat = "lat", grid = grid)

# Verify identical
all(result@cell_id == ref$seqnum)
#> TRUE
```

## Function Reference

### S4 Classes

| Class | Description |
|-------|-------------|
| `HexGridInfo` | Grid specification (aperture, resolution, area) |
| `HexData` | Hexified data with grid reference |

### Constructors

| Function | Description |
|----------|-------------|
| `hex_grid()` | Create a HexGridInfo specification |
| `hexify()` | Assign points to grid cells, returns HexData |

### HexData Methods

| Method | Description |
|--------|-------------|
| `grid_info(x)` | Extract HexGridInfo from HexData |
| `cells(x)` | Get unique cell IDs |
| `n_cells(x)` | Count unique cells |
| `as.data.frame(x)` | Extract underlying data frame |
| `x$column` | Access columns directly |
| `x[i, j]` | Subset rows/columns |

### Grid Generation

| Function | Description |
|----------|-------------|
| `grid_rect()` | Generate grid polygons over rectangular region |
| `grid_global()` | Generate global grid polygons |
| `cell_to_sf()` | Generate polygons from cell IDs |
| `as_sf()` | Convert HexData to sf object |

### Coordinate Conversion

| Function | Description |
|----------|-------------|
| `lonlat_to_cell()` | lon/lat -> cell ID |
| `cell_to_lonlat()` | cell ID -> cell center lon/lat |

### Visualization

| Function | Description |
|----------|-------------|
| `plot()` | Base R plot method for HexData |
| `hexify_ggplot()` | ggplot2 plotting for HexData |
| `hexify_map()` | Quick map visualization |
| `hexify_heatmap()` | ggplot2 choropleth heatmap |
| `plot_world()` | Quick world map |

### Grid Statistics

| Function | Description |
|----------|-------------|
| `hexify_compare_resolutions()` | Compare resolution levels |
| `dgearthstat()` | Get grid statistics |

## See Also

- `vignette("theory")` - Mathematical foundations (projection, apertures, space-filling curves)
- `vignette("workflows")` - Detailed workflow examples for common analysis tasks

## Session Info

```{r session-info}
sessionInfo()
```
