---
title: "Getting Started with hexify"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with hexify}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What is hexify?

**hexify** is an R package for working with discrete global grid systems (DGGS)
based on the ISEA (Icosahedral Snyder Equal Area) projection. It provides:

- Equal-area hexagonal cells covering the entire globe
- Hierarchical indexing with multiple apertures (3, 4, 7)
- Efficient point-to-cell assignment
- Cell center and boundary computation

## Installation

```{r eval=FALSE}
# From GitHub (development version)
devtools::install_github("username/hexify")
```

## Basic Usage

### 1. Assign points to grid cells

The most common use case: given longitude/latitude coordinates, find which
hexagonal cell each point belongs to.

```{r}
library(hexify)

# Sample coordinates (European cities)
cities <- data.frame(
  name = c("Vienna", "Paris", "London", "Berlin", "Rome"),
  lon = c(16.37, 2.35, -0.12, 13.40, 12.50),
  lat = c(48.21, 48.86, 51.51, 52.52, 41.90)
)

# Assign to grid cells (aperture 3, resolution 5)
indices <- hexify_lonlat_to_index(
  cities$lon, cities$lat,
  resolution = 5,
  aperture = 3
)

# Add cell indices to data
cities$cell_id <- indices
print(cities)
```

### 2. Get cell centers

Convert cell indices back to geographic coordinates (cell centers):

```{r}
# Get center of each cell
centers <- hexify_index_to_lonlat(cities$cell_id, aperture = 3)
print(centers)
```

### 3. Using the grid object interface

For more complex workflows, create a grid specification:

```{r}
# Create grid with target cell area of 1000 km²
grid <- hexify_construct(area = 1000, aperture = 3)
print(grid)

# Use with the grid object
cells <- hexify_geo_to_cell(grid, cities$lon, cities$lat)
print(cells)
```

## Aperture Selection

hexify supports three aperture values, each with different characteristics:

| Aperture | Refinement | Cell Count Growth | Best For |
|----------|------------|-------------------|----------|
| 3 | 3:1 | Slower | Fine resolution control |
| 4 | 4:1 | Medium | Standard applications |
| 7 | 7:1 | Faster | Coarse-to-fine jumps |

```{r}
# Compare cell counts at resolution 5
for (ap in c(3, 4, 7)) {
  idx <- hexify_lonlat_to_index(0, 0, resolution = 5, aperture = ap)
  cat(sprintf("Aperture %d: index = %s\n", ap, idx))
}
```

## Index Types

Each aperture has an optimal index encoding:

- **Aperture 3**: Z3 encoding (default)
- **Aperture 4**: Z-order (Morton) encoding (default)
- **Aperture 7**: Z7 encoding (default)

The index type affects how cell indices are represented as strings:

```{r}
# Z3 index (aperture 3)
idx_z3 <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 3)
cat("Z3 index:", idx_z3, "\n")

# Z-order index (aperture 4)
idx_zo <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 4)
cat("Z-order index:", idx_zo, "\n")
```

## Integer Sequential Numbers

For database storage, use integer sequential numbers instead of string indices:

```{r}
# Convert to integer seqnum
seqnum <- hexify_lonlat_to_seqnum(
  cities$lon, cities$lat,
  resolution = 5, aperture = 3
)
print(seqnum)

# Convert back to coordinates
coords <- hexify_seqnum_to_lonlat(seqnum, resolution = 5, aperture = 3)
print(coords)
```

## Hierarchical Operations

Navigate the grid hierarchy:

```{r}
# Get an index
idx <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 3)

# Get parent (coarser resolution)
parent <- hexify_get_parent(idx, aperture = 3)
cat("Index:", idx, "\n")
cat("Parent:", parent, "\n")

# Get children (finer resolution)
children <- hexify_get_children(idx, aperture = 3)
cat("Children:", paste(children, collapse = ", "), "\n")
```

## dggridR Compatibility

If you're familiar with dggridR, hexify provides compatible function names:

```{r}
# dggridR-style API
dggs <- dgconstruct(area = 1000, aperture = 3)
result <- dgGEO_to_SEQNUM(dggs, in_lon_deg = 16.37, in_lat_deg = 48.21)
print(result$seqnum)
```

## Next Steps

- See `vignette("theory")` for mathematical details
- See `vignette("workflows")` for real-world examples
- See `vignette("apertures")` for aperture comparison
