---
title: "Mathematical Foundations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hexify)
```

## Overview

hexify implements the **ISEA (Icosahedral Snyder Equal Area)** discrete global grid system. This vignette explains the mathematical foundations: the projection geometry, coordinate systems, aperture subdivision, and cell indexing.

## The Problem: Equal-Area Grids on a Sphere

Any projection from a sphere to a plane must distort *something*. For spatial statistics, we need cells of equal area regardless of location. Standard latitude-longitude grids fail badly: a 1° cell at the equator covers ~12,300 km², while the same cell near the poles covers a tiny fraction of that.

The ISEA projection solves this by:

1. Inscribing a regular icosahedron in the sphere
2. Projecting each spherical "cap" onto its corresponding flat triangular face using a modified Lambert equal-area projection
3. Overlaying a hexagonal grid on the resulting planar triangles

## The Lambert Azimuthal Equal-Area Projection

The foundation of Snyder's projection is Lambert's azimuthal equal-area projection. The projection maps a sphere of radius $R$ to a tangent plane while preserving area (Snyder, 1987, p. 182).

### Definition

For a projection centered at point $S$ with spherical coordinates $(\lambda_0, \phi_1)$, a point $P$ at angular distance $c$ from the center projects to distance $\rho = 2R\sin(c/2)$ from the origin on the plane. This chord distance equals the straight-line distance through 3D space from $S$ to $P$.

The projection is **not** perspective-based; it is a synthetic mathematical construction designed to achieve equal-area and azimuthal properties simultaneously.

```{r lambert-geometry, echo=FALSE, fig.width=7, fig.height=5.5}
# Accurate illustration of Lambert azimuthal equal-area projection
par(mar = c(1, 1, 2, 1), bg = "white")
plot(NULL, xlim = c(-2.5, 2.5), ylim = c(-1.2, 2.8), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Lambert Azimuthal Equal-Area Projection")

# Draw the sphere (circle in cross-section)
theta <- seq(0, 2*pi, length.out = 200)
R <- 1  # sphere radius
lines(R * cos(theta), R * sin(theta) + R, lwd = 2, col = "gray30")

# Draw the tangent plane (horizontal line at y = 0)
lines(c(-2.3, 2.3), c(0, 0), lwd = 2, col = "gray30")

# Mark the tangent point S (south pole of projection)
points(0, 0, pch = 19, cex = 1.5, col = "black")
text(0, -0.25, "S (tangent point)", cex = 0.9)

# Mark the center of sphere O
points(0, R, pch = 3, cex = 1.2, col = "gray50")
text(0.15, R + 0.15, "O", cex = 0.9, col = "gray50")

# Show a point P on the sphere and its projection P'
# Angular distance from S (colatitude) = 60 degrees
phi <- 60 * pi/180
P_x <- R * sin(phi)
P_y <- R * (1 - cos(phi))

# The chord distance d from S to P
d <- sqrt(P_x^2 + P_y^2)

# Lambert formula: projected distance = d (chord distance)
proj_x <- d

# Draw point P on sphere
points(P_x, P_y, pch = 19, cex = 1.3, col = "#E63946")
text(P_x + 0.15, P_y + 0.15, "P", cex = 1, col = "#E63946", font = 2)

# Draw the chord from S to P
lines(c(0, P_x), c(0, P_y), lwd = 2, col = "#E63946", lty = 2)

# Draw projected point P' on plane
points(proj_x, 0, pch = 19, cex = 1.3, col = "#457B9D")
text(proj_x, -0.25, "P'", cex = 1, col = "#457B9D", font = 2)

# Draw vertical dashed line from P down to show projection concept
lines(c(P_x, proj_x), c(P_y, 0), lwd = 1, col = "gray60", lty = 3)

# Label the chord distance
mid_x <- P_x / 2
mid_y <- P_y / 2
text(mid_x - 0.2, mid_y + 0.15, "d", cex = 1.1, col = "#E63946", font = 3)

# Label the projected distance (same as d)
text(proj_x / 2, -0.35, "d", cex = 1.1, col = "#457B9D", font = 3)

# Add angular distance arc
arc_r <- 0.4
arc_theta <- seq(3*pi/2, 3*pi/2 + phi, length.out = 30)
lines(arc_r * cos(arc_theta), R + arc_r * sin(arc_theta),
      lwd = 1.5, col = "gray50")
text(0.25, R - 0.55, expression(c), cex = 1, col = "gray50")

# Second example point at different angle
phi2 <- 100 * pi/180
P2_x <- R * sin(phi2)
P2_y <- R * (1 - cos(phi2))
d2 <- sqrt(P2_x^2 + P2_y^2)

points(P2_x, P2_y, pch = 19, cex = 1.1, col = "#2A9D8F")
text(P2_x + 0.12, P2_y + 0.15, expression(P[2]), cex = 0.9, col = "#2A9D8F")
lines(c(0, P2_x), c(0, P2_y), lwd = 1.5, col = "#2A9D8F", lty = 2)
points(d2, 0, pch = 19, cex = 1.1, col = "#2A9D8F")
text(d2, -0.25, expression(P[2]*"'"), cex = 0.9, col = "#2A9D8F")

# Annotation box
rect(-2.4, 2.1, 2.4, 2.7, col = "gray95", border = NA)
text(0, 2.5, "Lambert's rule: Place P' at distance d from S,", cex = 0.9)
text(0, 2.25, "where d is the straight-line (chord) distance from S to P", cex = 0.9)
```

### Forward Formulas

For a point at angular distance $c$ from the projection center, the projected distance is:

$$\rho = 2R \sin\left(\frac{c}{2}\right)$$

For the oblique aspect centered at $(\lambda_0, \phi_1)$, the full formulas are (Snyder, 1987, eq. 24-2 to 24-4, p. 185):

$$k' = \sqrt{\frac{2}{1 + \sin\phi_1\sin\phi + \cos\phi_1\cos\phi\cos(\lambda - \lambda_0)}}$$

$$x = R \cdot k' \cdot \cos\phi \cdot \sin(\lambda - \lambda_0)$$

$$y = R \cdot k' \cdot [\cos\phi_1\sin\phi - \sin\phi_1\cos\phi\cos(\lambda - \lambda_0)]$$

### Equal-Area Proof

The projection preserves area because the radial and tangential scale factors satisfy $h' \cdot k' = 1$ at every point. At angular distance $c$ from center (Snyder, 1987, eq. 24-22, 24-23, p. 188):

$$h' = \cos\left(\frac{c}{2}\right), \quad k' = \sec\left(\frac{c}{2}\right)$$

Therefore $h' \cdot k' = \cos(c/2) \cdot \sec(c/2) = 1$, confirming the equal-area property.

```{r lambert-area-preservation, echo=FALSE, fig.width=8, fig.height=4}
# Show equal-area property with concentric rings
par(mfrow = c(1, 2), mar = c(2, 1, 3, 1), bg = "white")

R <- 1

# Left: Sphere view (orthographic, looking from above at tangent point)
plot(NULL, xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Sphere (view from above S)")

# Draw outer circle (equator as seen from S)
theta <- seq(0, 2*pi, length.out = 100)
lines(cos(theta), sin(theta), lwd = 2)

# Draw concentric latitude circles and shade bands
n_bands <- 5
cols <- c("#264653", "#2A9D8F", "#E9C46A", "#F4A261", "#E76F51")

for (i in n_bands:1) {
  phi_outer <- (i / n_bands) * (pi/2)
  phi_inner <- ((i-1) / n_bands) * (pi/2)

  r_outer <- sin(phi_outer)
  r_inner <- sin(phi_inner)

  theta_seq <- seq(0, 2*pi, length.out = 100)
  if (i == 1) {
    polygon(r_outer * cos(theta_seq), r_outer * sin(theta_seq),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  } else {
    polygon(c(r_outer * cos(theta_seq), rev(r_inner * cos(theta_seq))),
            c(r_outer * sin(theta_seq), rev(r_inner * sin(theta_seq))),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  }
}
points(0, 0, pch = 19, cex = 1.2)
text(0, -1.2, "Equal-area bands on sphere", cex = 0.85)

# Right: Lambert projection
plot(NULL, xlim = c(-1.6, 1.6), ylim = c(-1.6, 1.6), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Lambert Projection")

for (i in n_bands:1) {
  phi_outer <- (i / n_bands) * (pi/2)
  phi_inner <- ((i-1) / n_bands) * (pi/2)

  # Lambert projection: r = 2*sin(phi/2)
  r_outer <- 2 * sin(phi_outer / 2)
  r_inner <- 2 * sin(phi_inner / 2)

  theta_seq <- seq(0, 2*pi, length.out = 100)
  if (i == 1) {
    polygon(r_outer * cos(theta_seq), r_outer * sin(theta_seq),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  } else {
    polygon(c(r_outer * cos(theta_seq), rev(r_inner * cos(theta_seq))),
            c(r_outer * sin(theta_seq), rev(r_inner * sin(theta_seq))),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  }
}

r_eq <- 2 * sin(pi/4)
lines(r_eq * cos(theta), r_eq * sin(theta), lwd = 2, lty = 2, col = "gray50")

points(0, 0, pch = 19, cex = 1.2)
text(0, -1.5, "Same bands after projection (areas preserved)", cex = 0.85)
```

Each colored band has equal area on the sphere. After Lambert projection, shapes change (outer bands stretch radially, compress tangentially) but areas remain equal.

### Inverse Formulas

The inverse mapping $(x, y) \to (\lambda, \phi)$ recovers geographic coordinates from planar coordinates (Snyder, 1987, eq. 24-14 to 24-16, p. 187):

$$\rho = \sqrt{x^2 + y^2}$$

$$c = 2\arcsin\left(\frac{\rho}{2R}\right)$$

where $c$ is the angular distance from the projection center. Then:

$$\phi = \arcsin\left[\cos c \cdot \sin\phi_1 + \frac{y \cdot \sin c \cdot \cos\phi_1}{\rho}\right]$$

$$\lambda = \lambda_0 + \arctan\left[\frac{x \cdot \sin c}{\rho\cos\phi_1\cos c - y\sin\phi_1\sin c}\right]$$

If $\rho = 0$, the point is at the projection center: $\phi = \phi_1$, $\lambda = \lambda_0$.

### Limitations

**Antipode singularity:** The point diametrically opposite the projection center (angular distance 180°) maps to infinity and must be excluded from the domain.

**Shape distortion:** While area is preserved exactly, shapes distort increasingly with distance from center. The maximum angular distortion $\omega$ at distance $c$ is (Snyder, 1987, eq. 24-24, p. 188):

$$\sin\left(\frac{\omega}{2}\right) = \frac{k'^2 - 1}{k'^2 + 1}$$

At $c = 90°$, distortion reaches $\omega \approx 70.5°$. ISEA grids limit this by using icosahedral faces subtending only ~72° from face center.

**No conformality:** No projection can be both equal-area and conformal (angle-preserving)---a fundamental constraint from differential geometry (Snyder, 1987, p. 16-18).

## The Icosahedron

Lambert's projection works for a single tangent plane covering at most a hemisphere. To cover the entire globe with minimal distortion, Snyder used **20 tangent planes**---one for each face of a regular icosahedron (Snyder, 1992, p. 10).

### Geometry

A regular icosahedron has:

- **20 equilateral triangular faces** (each covering ~1/20 of Earth's surface)
- **12 vertices** (where 5 faces meet---these become pentagon cells)
- **30 edges**

```{r icosahedron-concept, echo=FALSE, fig.width=8, fig.height=4}
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1), bg = "white")

# Left: conceptual diagram
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Icosahedron inscribed in sphere")

theta <- seq(0, 2*pi, length.out = 100)
lines(cos(theta), sin(theta), lwd = 2, col = "gray70")

pent_r <- 0.38
pent_angles <- seq(pi/2, pi/2 + 2*pi, length.out = 6)[-6]
pent_x <- pent_r * cos(pent_angles)
pent_y <- pent_r * sin(pent_angles)

outer_r <- 0.95
outer_angles <- seq(pi/2 + pi/5, pi/2 + pi/5 + 2*pi, length.out = 6)[-6]
outer_x <- outer_r * cos(outer_angles)
outer_y <- outer_r * sin(outer_angles)

face_cols <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A", "#F4A261")
for (i in 1:5) {
  j <- (i %% 5) + 1
  polygon(c(pent_x[i], pent_x[j], outer_x[i]),
          c(pent_y[i], pent_y[j], outer_y[i]),
          col = adjustcolor(face_cols[i], 0.4),
          border = "gray30", lwd = 1.5)
}

polygon(pent_x, pent_y, col = adjustcolor("gray80", 0.3),
        border = "gray30", lwd = 1.5)

points(c(0, pent_x, outer_x), c(0, pent_y, outer_y), pch = 19, cex = 0.8)

text(0, -1.3, "Each triangular face = 1/20 of Earth", cex = 0.85)

# Right: show projection concept
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-0.5, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Projection to one face")

arc_theta <- seq(-pi/3, pi/3, length.out = 50)
lines(1.3 * sin(arc_theta), 1.3 * cos(arc_theta) + 0.5, lwd = 2, col = "gray50")

tri_x <- c(-0.9, 0, 0.9, -0.9)
tri_y <- c(0, 1.5, 0, 0)
polygon(tri_x, tri_y, col = adjustcolor("#457B9D", 0.3),
        border = "#264653", lwd = 2)

points(0, 0.5, pch = 19, cex = 1.2, col = "#E63946")
text(0.25, 0.5, "face center", cex = 0.8, adj = 0)

sphere_pts <- data.frame(
  x = c(-0.4, 0.3, 0.1),
  y = c(1.1, 0.9, 1.3)
)
for (i in 1:nrow(sphere_pts)) {
  points(sphere_pts$x[i], sphere_pts$y[i], pch = 19, cex = 0.8, col = "#2A9D8F")
  arrows(sphere_pts$x[i], sphere_pts$y[i] - 0.1,
         sphere_pts$x[i] * 0.7, 0.4,
         length = 0.08, lwd = 1.5, col = "#2A9D8F")
}

text(0, -0.35, "Lambert projection applied to each face", cex = 0.85)
```

### Vertex Latitude Derivation

The 12 vertices are located at (Coxeter, 1973, p. 52-53):

| Location | Latitude | Longitudes |
|----------|----------|------------|
| North pole | +90° | 0° |
| Upper ring | $+\arctan(1/2) \approx +26.565°$ | 0°, 72°, 144°, 216°, 288° |
| Lower ring | $-\arctan(1/2) \approx -26.565°$ | 36°, 108°, 180°, 252°, 324° |
| South pole | −90° | 0° |

The latitude $\arctan(1/2)$ arises from the golden ratio geometry. An icosahedron can be constructed from three mutually perpendicular golden rectangles ($1 \times \varphi$, where $\varphi = (1+\sqrt{5})/2$). The non-polar vertices have $z$-coordinate $1/s$ where $s = \sqrt{1 + \varphi^2}$, yielding $\tan\phi = 1/2$.

### Standard ISEA Orientation

The default orientation places vertex 0 at longitude 11.25°, latitude 58.28252559°, with azimuth 0°. This places icosahedron vertices (pentagon cells) predominantly over oceans (Sahr et al., 2003, p. 123).

```{r face-centers, echo=FALSE, fig.width=8, fig.height=4}
library(sf)
library(ggplot2)

centers <- hexify_face_centers()

center_pts <- st_as_sf(
  data.frame(face = 0:19, lon = centers$lon * 180 / pi, lat = centers$lat * 180 / pi),
  coords = c("lon", "lat"), crs = 4326
)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = center_pts, color = "#E63946", size = 3) +
  geom_sf_text(data = center_pts, aes(label = face), nudge_y = 6, size = 2.5) +
  labs(title = "ISEA Icosahedron Face Centers",
       subtitle = "20 triangular faces, numbered 0-19") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Snyder's ISEA Projection

Snyder extended the Lambert projection to the icosahedron by introducing an azimuth-adjustment transformation that ensures seamless transitions between adjacent faces while maintaining the equal-area property (Snyder, 1992, p. 12).

### Key Constants

| Constant | Symbol | Value | Source |
|----------|--------|-------|--------|
| Edge-to-center angle | $E_l$ | 37.37736814° | Snyder (1992, Table 1, p. 14) |
| Geometric angle | $G$ | 36° | 360°/10 (icosahedral 5-fold symmetry) |
| Scale factor | $R_1$ | 0.9103832815 | Snyder (1992, Table 1, p. 14) |

### Forward Projection Steps

The complete algorithm comprises seven steps (Snyder, 1992, p. 13-15):

**Step 1: Compute angular distance and azimuth** from face center $(\lambda_0, \phi_0)$ to point $(\lambda, \phi)$:

$$z = \arccos(\sin\phi_0 \sin\phi + \cos\phi_0 \cos\phi \cos(\lambda - \lambda_0))$$
$$\text{Az} = \arctan2(\cos\phi \sin(\lambda - \lambda_0), \cos\phi_0 \sin\phi - \sin\phi_0 \cos\phi \cos(\lambda - \lambda_0))$$

**Step 2: Reduce azimuth** to [0°, 120°) by exploiting 3-fold symmetry.

**Step 3: Compute auxiliary angle** $\delta_z$ (Snyder, 1992, eq. 8, p. 14):
$$\delta_z = \arctan\left(\frac{\tan E_l}{\cos \text{Az} + \cot 30° \cdot \sin \text{Az}}\right)$$

**Step 4: Compute auxiliary angle** $h$ (Snyder, 1992, eq. 9, p. 14):
$$h = \arccos(\sin \text{Az} \sin G \cos E_l - \cos \text{Az} \cos G)$$

**Step 5: Compute adjusted azimuth** $\text{Az}'$ (Snyder, 1992, eq. 10-11, p. 14):
$$A_G = \text{Az} + G + h - \pi$$
$$\text{Az}' = \arctan\left(\frac{2 A_G}{R_1^2 \tan^2 E_l - 2 A_G \cot 30°}\right)$$

**Step 6: Compute radial distance** (Snyder, 1992, eq. 12-13, p. 14-15):
$$f = \frac{\tan E_l}{2(\cos \text{Az}' + \cot 30° \cdot \sin \text{Az}') \sin(\delta_z / 2)}$$
$$\rho = 2 R_1 f \sin(z / 2)$$

**Step 7: Convert to Cartesian** with sector offset restored.

### The Inverse Projection

The inverse projection cannot be solved analytically because the azimuth adjustment contains transcendental functions. A Newton-Raphson iteration finds the spherical azimuth Az from the planar azimuth Az':

$$f(\text{Az}) = \text{agh} - \text{Az} - G + (\pi - h) = 0$$

where $h = \arccos(\sin \text{Az} \sin G \cos E_l - \cos \text{Az} \cos G)$.

```{r newton-raphson, echo=FALSE, fig.width=7, fig.height=5}
par(mar = c(4, 4, 3, 1), bg = "white")

f <- function(x) x^3 - 2*x - 2
f_prime <- function(x) 3*x^2 - 2

x_seq <- seq(0.5, 2.5, length.out = 100)
plot(x_seq, f(x_seq), type = "l", lwd = 2, col = "gray30",
     xlab = "Azimuth estimate", ylab = "f(Az) - target",
     main = "Newton-Raphson Iteration for Inverse Projection",
     ylim = c(-3, 4))
abline(h = 0, lty = 2, col = "gray50")

x0 <- 2.3
cols <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A")
x_current <- x0

for (i in 1:4) {
  y_current <- f(x_current)
  slope <- f_prime(x_current)
  x_next <- x_current - y_current / slope

  points(x_current, y_current, pch = 19, cex = 1.2, col = cols[i])
  text(x_current, y_current + 0.4, paste0("x", i-1), cex = 0.8, col = cols[i])

  if (i < 4) {
    segments(x_current, y_current, x_next, 0, lty = 2, col = cols[i], lwd = 1.5)
    segments(x_next, 0, x_next, f(x_next), lty = 3, col = cols[i+1])
  }

  x_current <- x_next
}

root <- 1.769
points(root, 0, pch = 4, cex = 2, lwd = 2, col = "black")
text(root, -0.5, "solution", cex = 0.9)

legend("topleft",
       legend = c("f(Az)", "Iteration steps", "Converged solution"),
       lty = c(1, 2, NA), pch = c(NA, 19, 4),
       col = c("gray30", "#457B9D", "black"),
       lwd = c(2, 1.5, 2), bty = "n", cex = 0.85)
```

The iteration exhibits **quadratic convergence**, typically reaching machine precision in 3-5 iterations. hexify provides four precision modes:

| Mode | Tolerance | Typical Iterations | Use Case |
|------|-----------|-------------------|----------|
| fast | $10^{-10}$ | 3-4 | Interactive visualization |
| default | $10^{-12}$ | 4-5 | General applications (~1 m accuracy) |
| high | $10^{-14}$ | 5-6 | High-precision geodesy |
| ultra | $10^{-15}$ | 6-7 | Research |

## Aperture and Resolution

**Aperture** defines how cells subdivide at each resolution level---it's the ratio of parent cell area to child cell area (Sahr et al., 2003, p. 124).

```{r aperture-diagrams, echo=FALSE, fig.width=9, fig.height=3.5}
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1), bg = "white")

hex_vertices <- function(cx, cy, r, flat_top = TRUE) {
  offset <- if (flat_top) 0 else pi/6
  angles <- seq(offset, 2*pi + offset, length.out = 7)
  list(x = cx + r * cos(angles), y = cy + r * sin(angles))
}

# Aperture 3
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Aperture 3: Triangle (30° rotation)")

h_parent <- hex_vertices(0, 0, 1.7, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "gray40", lwd = 2, col = NA)

child_r <- 0.85
child_positions <- list(
  c(0, 0.6),
  c(-0.52, -0.3),
  c(0.52, -0.3)
)
child_cols <- c("#E63946", "#457B9D", "#2A9D8F")

for (i in 1:3) {
  h <- hex_vertices(child_positions[[i]][1], child_positions[[i]][2],
                    child_r, flat_top = FALSE)
  polygon(h$x, h$y, col = adjustcolor(child_cols[i], 0.5),
          border = "gray30", lwd = 1.5)
  text(child_positions[[i]][1], child_positions[[i]][2], i,
       cex = 1.2, font = 2)
}
text(0, -1.85, expression("N = 10" %*% 3^r + 2), cex = 0.85)

# Aperture 4
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Aperture 4: Rhombic (no rotation)")

h_parent <- hex_vertices(0, 0, 1.7, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "gray40", lwd = 2, col = NA)

child_r4 <- 0.7
positions4 <- list(
  c(-0.45, 0.45), c(0.45, 0.45),
  c(-0.45, -0.45), c(0.45, -0.45)
)
child_cols4 <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A")

for (i in 1:4) {
  h <- hex_vertices(positions4[[i]][1], positions4[[i]][2],
                    child_r4, flat_top = TRUE)
  polygon(h$x, h$y, col = adjustcolor(child_cols4[i], 0.5),
          border = "gray30", lwd = 1.5)
  text(positions4[[i]][1], positions4[[i]][2], i, cex = 1.2, font = 2)
}
text(0, -1.85, expression("N = 10" %*% 4^r + 2), cex = 0.85)

# Aperture 7
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Aperture 7: Rosette (19.1° rotation)")

h_parent <- hex_vertices(0, 0, 1.7, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "gray40", lwd = 2, col = NA)

child_r7 <- 0.52
h <- hex_vertices(0, 0, child_r7, flat_top = TRUE)
polygon(h$x, h$y, col = adjustcolor("#F4A261", 0.5),
        border = "gray30", lwd = 1.5)
text(0, 0, "1", cex = 1, font = 2)

ring_r <- 0.82
ring_cols <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A", "#9B5DE5", "#00BBF9")
for (i in 1:6) {
  ang <- pi/2 + (i-1) * pi/3
  cx <- ring_r * cos(ang)
  cy <- ring_r * sin(ang)
  h <- hex_vertices(cx, cy, child_r7, flat_top = TRUE)
  polygon(h$x, h$y, col = adjustcolor(ring_cols[i], 0.5),
          border = "gray30", lwd = 1.5)
  text(cx, cy, i+1, cex = 0.9, font = 2)
}
text(0, -1.85, expression("N = 10" %*% 7^r + 2), cex = 0.85)
```

### Aperture Properties

| Aperture | Area Ratio | Linear Scale | Rotation per Level | Orientation |
|----------|------------|--------------|-------------------|-------------|
| 3 | 1:3 | $\sqrt{3} \approx 1.73$ | 30° | Alternates Class I/II |
| 4 | 1:4 | $2.0$ | 0° | Always Class I |
| 7 | 1:7 | $\sqrt{7} \approx 2.65$ | $\arctan(\sqrt{3/7}) \approx 19.1°$ | Class III |

The aperture 7 rotation angle $\arctan(\sqrt{3/7})$ arises from the geometric constraint that 7 hexagons in a rosette pattern (1 center + 6 ring) must maintain lattice consistency (DGGRID Manual, 2023).

### Cell Count Growth

```{r cell-counts}
cat("Resolution  Aperture 3    Aperture 4    Aperture 7\n")
cat("---------  ----------    ----------    ----------\n")
for (res in 0:8) {
  cells_ap3 <- 10 * 3^res + 2
  cells_ap4 <- 10 * 4^res + 2
  cells_ap7 <- 10 * 7^res + 2
  cat(sprintf("    %d      %10s    %10s    %10s\n",
              res,
              format(cells_ap3, big.mark = ","),
              format(cells_ap4, big.mark = ","),
              format(cells_ap7, big.mark = ",")))
}
```

## Orientation Classes

```{r orientation-classes, echo=FALSE, fig.width=8, fig.height=3}
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1), bg = "white")

hex_v <- function(cx, cy, r, rotation = 0) {
  angles <- seq(rotation, 2*pi + rotation, length.out = 7)
  list(x = cx + r * cos(angles), y = cy + r * sin(angles))
}

# Class I: Flat-top
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Class I (Flat-top, 0°)")
h <- hex_v(0, 0, 1.2, rotation = 0)
polygon(h$x, h$y, col = adjustcolor("#457B9D", 0.4), border = "gray30", lwd = 2)
lines(h$x[1:2], h$y[1:2], col = "#E63946", lwd = 3)
text(0, -1.35, "Aperture 4 (all res)\nAperture 3 (even res)", cex = 0.8)

# Class II: Pointy-top (30° rotation)
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Class II (Pointy-top, 30°)")
h <- hex_v(0, 0, 1.2, rotation = pi/6)
polygon(h$x, h$y, col = adjustcolor("#2A9D8F", 0.4), border = "gray30", lwd = 2)
points(h$x[1], h$y[1], pch = 19, cex = 1.5, col = "#E63946")
text(0, -1.35, "Aperture 3 (odd res)", cex = 0.8)

# Class III: Skewed
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Class III (Skewed, ~19.1°)")
h_parent <- hex_v(0, 0, 1.2, rotation = 0)
polygon(h_parent$x, h_parent$y, col = NA, border = "gray50", lwd = 2, lty = 2)
# Correct rotation: arctan(sqrt(3/7))
rot_angle <- atan(sqrt(3/7))
h_child <- hex_v(0, 0, 0.8, rotation = rot_angle)
polygon(h_child$x, h_child$y, col = adjustcolor("#E9C46A", 0.4),
        border = "gray30", lwd = 2)
arc_r <- 0.5
arc_theta <- seq(0, rot_angle, length.out = 20)
lines(arc_r * cos(arc_theta), arc_r * sin(arc_theta), col = "#E63946", lwd = 2)
text(0.6, 0.15, "19.1°", cex = 0.8, col = "#E63946")
text(0, -1.35, "Aperture 7 (all res)\nRotation accumulates", cex = 0.8)
```

For **aperture 3**, orientation alternates between Class I (flat-top) and Class II (pointy-top) at each resolution.
For **aperture 7**, each level adds a rotation of $\arctan(\sqrt{3/7}) \approx 19.1°$ (Sahr, 2008, p. 176).

## Pentagon Cells

Exactly **12 cells are pentagons** at every resolution. This is a topological necessity derived from Euler's formula (Coxeter, 1973, p. 10):

$$V - E + F = 2$$

For a tiling with $h$ hexagons and $p$ pentagons on a sphere:
- $V = (6h + 5p)/3$, $E = (6h + 5p)/2$, $F = h + p$

Substituting into Euler's formula and simplifying yields $p = 12$, independent of $h$.

```{r pentagon-locations, echo=FALSE, fig.width=8, fig.height=4}
pentagon_coords <- data.frame(
  type = c("Pole", "Pole", rep("Upper ring", 5), rep("Lower ring", 5)),
  lon = c(0, 0, 0, 72, 144, 216, 288, 36, 108, 180, 252, 324),
  lat = c(90, -90, rep(26.57, 5), rep(-26.57, 5))
)

pentagon_pts <- st_as_sf(pentagon_coords, coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = pentagon_pts, aes(color = type), size = 4) +
  scale_color_manual(values = c("Pole" = "#E63946",
                                 "Upper ring" = "#457B9D",
                                 "Lower ring" = "#2A9D8F")) +
  labs(title = "Pentagon Cell Locations",
       subtitle = "12 pentagonal cells at icosahedron vertices (area = 5/6 of hexagons)",
       color = "Location") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

| Location | Latitude | Longitudes |
|----------|----------|------------|
| Poles | ±90° | 0° |
| Upper ring | $\arctan(1/2) \approx 26.57°$ | 0°, 72°, 144°, 216°, 288° |
| Lower ring | $-\arctan(1/2) \approx -26.57°$ | 36°, 108°, 180°, 252°, 324° |

Pentagon area is exactly 5/6 of hexagonal cell area at the same resolution (Sahr et al., 2003, p. 125).

## Coordinate Systems

hexify uses a multi-stage coordinate pipeline (Sahr, 2008, p. 178):

| System | Components | Description |
|--------|------------|-------------|
| **GEO** | lon, lat | WGS84 degrees |
| **Icosa Triangle** | face (0-19), tx, ty | Snyder projection output |
| **Quad XY** | quad (0-11), qx, qy | Paired-triangle coordinates |
| **Quad IJ** | quad (0-11), i, j | Quantized grid indices |
| **SEQNUM** | integer | Global cell ID (1-based) |

### Triangle to Quad

The 20 triangular faces are paired into 12 "quads" (diamond-shaped regions). Each quad contains two adjacent triangular faces sharing an edge, simplifying grid indexing (DGGRID Manual, 2023).

### SEQNUM Assignment

The SEQNUM provides a unique integer for each cell. The formula is:

$$N(r) = 10 \times a^r + 2$$

where $a$ is the aperture and $r$ is the resolution. SEQNUMs are assigned to maintain compatibility with dggridR.

## Round-Trip Accuracy

```{r round-trip}
original_lon <- 16.37
original_lat <- 48.21

cat(sprintf("Original: (%.4f, %.4f)\n\n", original_lon, original_lat))

for (ap in c(3, 4, 7)) {
  res <- if (ap == 7) 6 else 10
  grid <- hex_grid(resolution = res, aperture = ap)
  cell_id <- lonlat_to_cell(original_lon, original_lat, grid)
  recovered <- cell_to_lonlat(cell_id, grid)
  error_km <- sqrt((recovered$lon - original_lon)^2 +
                   (recovered$lat - original_lat)^2) * 111
  cat(sprintf("Aperture %d (res %2d): cell %d -> (%.4f, %.4f), ~%.1f km from center\n",
              ap, res, cell_id,
              recovered$lon, recovered$lat, error_km))
}
```

## Summary of ISEA Properties

| Property | Description |
|----------|-------------|
| **Equal area** | All hexagonal cells identical; pentagons = 5/6 hex area |
| **Bounded distortion** | Max angular distortion ~17.3° at face edges |
| **Uniform topology** | Hexagons have 6 neighbors; pentagons have 5 |
| **12 pentagons** | Topological necessity from Euler's formula |

## References

- Coxeter, H.S.M. (1973). *Regular Polytopes* (3rd ed.). Dover Publications.

- DGGRID Manual (2023). *DGGRID Version 7.8 Documentation*. https://github.com/sahrk/DGGRID

- Sahr, K. (2008). Location coding on icosahedral aperture 3 hexagon discrete global grids. *Computers, Environment and Urban Systems*, 32(3), 174-187.

- Sahr, K., White, D., & Kimerling, A.J. (2003). Geodesic Discrete Global Grid Systems. *Cartography and Geographic Information Science*, 30(2), 121-134.

- Snyder, J.P. (1987). *Map Projections: A Working Manual*. U.S. Geological Survey Professional Paper 1395.

- Snyder, J.P. (1992). An equal-area map projection for polyhedral globes. *Cartographica*, 29(1), 10-21.
