---
title: "Mathematical Foundations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

hexify implements the **ISEA (Icosahedral Snyder Equal Area)** discrete global
grid system. This vignette explains the mathematical foundations.

## The Icosahedral Projection

### Why an Icosahedron?

A regular icosahedron has 20 triangular faces, each with identical geometry.
This makes it ideal for:

1. **Equal-area partitioning**: Each face covers approximately 1/20 of Earth's surface
2. **Low distortion**: Each triangular face can be projected with minimal distortion
3. **Hierarchical subdivision**: Triangular faces subdivide naturally

### The Snyder Projection

The Snyder Equal Area projection maps points on the sphere to the icosahedron:

1. **Determine the face**: Find which of 20 icosahedral faces contains the point
2. **Project to face plane**: Use the Snyder formulas to compute (x, y) on the face
3. **Normalize coordinates**: Convert to (tx, ty) in range approximately [0, 1]

```{r}
library(hexify)

# Project a point
result <- hexify_forward(lon = 16.37, lat = 48.21)
cat(sprintf("Face: %d, tx: %.4f, ty: %.4f\n",
            result["face"], result["tx"], result["ty"]))
```

### Icosahedron Orientation

The icosahedron can be oriented arbitrarily on the globe. The default ISEA orientation places vertex 0 at:
- Longitude: 45°/4
- Latitude: arctan(√5/2)

This orientation was chosen by Sahr et al. (2003) to minimize distortion over land masses while placing the 12 pentagonal cells (which occur at icosahedron vertices) in oceanic or low-population areas. The specific vertex latitude arctan(√5/2) arises from the geometry of a regular icosahedron inscribed in a sphere.

```{r}
# Get all face centers
centers <- hexify_face_centers()
head(centers, 5)
```

## Hexagonal Grid Generation

### From Triangles to Hexagons

Each triangular face is subdivided into hexagonal cells. The subdivision process:

1. Overlay a hexagonal lattice on each face
2. Assign each point to the nearest hexagon center
3. Index cells hierarchically

### Aperture and Resolution

**Aperture** determines how many child cells fit in one parent cell:

| Aperture | Children per Parent | Class Pattern | Description |
|----------|---------------------|---------------|-------------|
| 3 | 3 | I/II alternating | Fine resolution control |
| 4 | 4 | I only | Power-of-2 scaling |
| 7 | 7 | III only | Fastest cell growth |
| 4/3 | 4 then 3 | Mixed | Aperture 4 for first m levels, then 3 |

**Resolution** is the number of subdivision levels. Cell count formulas:

| Aperture | Cell Count Formula |
|----------|-------------------|
| 3 | 10 × 3^res + 2 |
| 4 | 10 × 4^res + 2 |
| 7 | 10 × 7^res + 2 |
| 4/3 | 10 × 4^m × 3^(res-m) + 2 |

```{r}
# Cell count growth for all apertures
cat("Resolution  Aperture 3    Aperture 4    Aperture 7\n")
cat("---------  ----------    ----------    ----------\n")
for (res in 0:5) {
  cells_ap3 <- 10 * 3^res + 2
  cells_ap4 <- 10 * 4^res + 2
  cells_ap7 <- 10 * 7^res + 2
  cat(sprintf("    %d      %9d     %9d     %9d\n",
              res, cells_ap3, cells_ap4, cells_ap7))
}
```

For **mixed aperture (4/3)**, the first `m` levels use aperture 4, then remaining levels use aperture 3:

```{r}
# Mixed aperture example: m=2 means first 2 levels use aperture 4
m <- 2
cat("Mixed 4/3 (m=2): first 2 levels aperture 4, rest aperture 3\n")
for (res in 0:5) {
  if (res <= m) {
    cells <- 10 * 4^res + 2
  } else {
    cells <- 10 * 4^m * 3^(res - m) + 2
  }
  cat(sprintf("  Res %d: %d cells\n", res, cells))
}
```

### Orientation Classes

DGGRID defines three orientation classes based on how child cells align with parent cells during subdivision:

- **Class I**: Child cell vertices align with parent cell vertices. Hexagons have a "flat top" orientation.
- **Class II**: Child cell centers align with parent cell vertices. Hexagons are rotated 30° ("pointy top").
- **Class III**: Neither alignment—child cells are rotated relative to parent cells (skewed grids).

Each aperture produces a characteristic class pattern:

| Aperture | Class Pattern | Description |
|----------|---------------|-------------|
| 3 | I → II → I → II... | Alternates between Class I and II at each resolution |
| 4 | I → I → I... | Always Class I (aligned subdivision) |
| 7 | III → III → III... | Always Class III (arctan(√3/5) rotation per level) |
| 4/3 | I → I → ... → I/II | Class I for aperture-4 levels, then alternates |

For **aperture 3**, odd resolutions (1, 3, 5, ...) produce Class I grids with flat-top hexagons, while even resolutions (2, 4, 6, ...) produce Class II grids with pointy-top hexagons.

For **aperture 4**, the 2×2 subdivision preserves alignment, so all resolutions are Class I.

For **aperture 7**, the subdivision involves a √7 scaling and arctan(√3/5) rotation, creating the characteristic Class III skewed pattern. This rotation accumulates: resolution 2 is rotated 2×arctan(√3/5) from resolution 0, and so on.

For **mixed aperture (4/3)**, the class pattern follows aperture 4 (Class I) for the first `m` levels, then switches to aperture 3's alternating I/II pattern for subsequent levels.

The class affects hexagon orientation and the coordinate system internals, but not the fundamental cell assignment or equal-area property.

## Space-Filling Curves

### Why Space-Filling Curves?

Space-filling curves provide:

1. **Compact representation**: Cell identity as a single integer/string
2. **Hierarchical encoding**: Parent-child relationships from index structure
3. **Spatial locality**: Nearby cells often have similar indices

Each aperture requires a different curve that matches its subdivision geometry.

### Z3 Encoding (Aperture 3)

For aperture 3, the Z3 curve uses base-3 (ternary) encoding. Each resolution level adds one ternary digit representing which of the 3 child cells contains the point.

```
Index = FF D₁D₂...Dₙ
where:
  FF = face number (00-19)
  Dᵢ = digit 0, 1, or 2 (base 3)
```

The three child positions form a triangular arrangement within each parent cell. The digit values correspond to:

- **0**: Center-bottom child
- **1**: Upper-left child
- **2**: Upper-right child

This encoding preserves hierarchy: truncating the last digit gives the parent cell index.

```{r}
# Z3 index structure
idx <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 3)
cat("Index:", idx, "\n")
cat("Face:", substr(idx, 1, 2), "\n")
cat("Path:", substr(idx, 3, nchar(idx)), "\n")

# Parent is obtained by dropping the last digit
parent_idx <- substr(idx, 1, nchar(idx) - 1)
cat("Parent:", parent_idx, "\n")
```

### Z-Order (Morton) Encoding (Aperture 4)

For aperture 4, the Morton curve (Z-order curve) interleaves the bits of i and j grid coordinates. This classic space-filling curve maps a 2D grid to a 1D index while preserving locality.

The 2×2 subdivision at each level creates 4 child cells arranged in a square:

```
  +---+---+
  | 2 | 3 |
  +---+---+
  | 0 | 1 |
  +---+---+
```

Each child position is encoded as 2 bits: the low bit is from i, the high bit is from j. For resolution r, the index uses 2r bits total.

```
For cell at grid position (i, j):
  bit 0 = i₀, bit 1 = j₀   (resolution 1)
  bit 2 = i₁, bit 3 = j₁   (resolution 2)
  ...
```

This interleaving creates the characteristic "Z" pattern that gives the curve its name.

```{r}
# Z-order index
idx4 <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 4)
cat("Z-order index:", idx4, "\n")

# The index encodes (quad, i, j) implicitly
coords <- hexify_index_to_cell(idx4, aperture = 4)
cat("Decoded - Quad:", coords$quad, "i:", coords$i, "j:", coords$j, "\n")
```

### Z7 Encoding (Aperture 7)

Aperture 7 uses base-7 encoding with a twist: each subdivision involves both scaling (by √7) and rotation (by arctan(√3/5)). The 7 child cells form a hexagonal rosette pattern:

```
      1
    6   2
      0
    5   3
      4
```

The center cell (0) is surrounded by 6 peripheral cells (1-6) arranged in a hexagonal ring. Unlike apertures 3 and 4, the child cells are rotated relative to the parent, which is why aperture 7 produces Class III grids.

The encoding uses Eisenstein integers (complex numbers of the form a + bω where ω = e^(2πi/3)) to handle the hexagonal geometry. Each digit 0-6 represents a position in the rosette.

```{r}
# Z7 index
idx7 <- hexify_lonlat_to_index(16.37, 48.21, resolution = 3, aperture = 7)
cat("Z7 index:", idx7, "\n")

# Hierarchy still works - parent is obtained by dropping last digit
parent7 <- substr(idx7, 1, nchar(idx7) - 1)
cat("Parent:", parent7, "\n")
```

### Mixed Aperture (4/3) Encoding

Mixed aperture grids combine the aperture 4 and aperture 3 encoding schemes. The first `m` resolution levels use Morton (Z-order) encoding, then subsequent levels use Z3 encoding.

This hybrid approach provides:
- Faster initial cell growth (aperture 4 at coarse scales)
- Finer resolution control (aperture 3 at fine scales)
- dggridR compatibility with ISEA43H grids

```{r, eval = FALSE}
# Mixed aperture index (via hexify() function)
result <- hexify(data.frame(lon = 16.37, lat = 48.21),
                 lon = "lon", lat = "lat", area = 1000, aperture = "4/3")
cat("Mixed 4/3 cell_id:", result$cell_id, "\n")
```

### Comparison of Encodings

| Property | Z3 (Aperture 3) | Morton (Aperture 4) | Z7 (Aperture 7) | Mixed (4/3) |
|----------|-----------------|---------------------|-----------------|-------------|
| Base | 3 (ternary) | 4 (quaternary) | 7 | 4 then 3 |
| Digits per level | 1 | 2 bits | 1 | Mixed |
| Child arrangement | Triangle | Square | Hex rosette | Square → Triangle |
| Rotation per level | 0° or 30° | 0° | arctan(√3/5) | Mixed |
| Locality | Good | Excellent | Good | Good |

## Inverse Projection

Converting cell coordinates back to lon/lat requires:

1. **Decode index**: Extract face and (i, j) coordinates
2. **Compute cell center**: Find center point in face coordinates
3. **Inverse Snyder**: Convert face (x, y) to geographic (lon, lat)

```{r}
# Round-trip test for all apertures
original_lon <- 16.37
original_lat <- 48.21

cat(sprintf("Original coordinates: (%.4f, %.4f)\n\n", original_lon, original_lat))

for (ap in c(3, 4, 7)) {
  # Use appropriate resolution for each aperture
  res <- if (ap == 7) 6 else 10
  idx <- hexify_lonlat_to_index(original_lon, original_lat,
                                 resolution = res, aperture = ap)
  recovered <- hexify_index_to_lonlat(idx, aperture = ap)
  error <- sqrt((recovered["lon"] - original_lon)^2 +
                (recovered["lat"] - original_lat)^2)
  cat(sprintf("Aperture %d (res %2d): recovered (%.4f, %.4f), error: %.6f°\n",
              ap, res, recovered["lon"], recovered["lat"], error))
}
```

## Cell Properties

### Equal Area

A key property of ISEA grids: all cells have approximately equal area.

```{r}
# Compare cell areas across apertures at similar resolutions
earth_area_km2 <- 510072000

cat("Aperture comparison at ~1000 km² target:\n\n")
for (ap in c(3, 4, 7)) {
  grid <- hexify_grid(area = 1000, aperture = ap)
  cat(sprintf("Aperture %d: resolution %d → %.1f km² (%.0f cells)\n",
              ap, grid$resolution, grid$area, grid$total_cells))
}

# Mixed aperture available via hexify() function, not hexify_grid()
cat("\n(Mixed aperture 4/3 is available via hexify() with aperture = '4/3')\n")
```

### Cell Shape

Cells are approximately hexagonal, but:

- Cells at face edges may be irregular
- Cell shape varies slightly with latitude due to projection
- Pentagon cells exist at icosahedron vertices (12 total)

## References

- Sahr, K., White, D., & Kimerling, A. J. (2003). Geodesic discrete global grid systems.
- Snyder, J. P. (1992). An equal-area map projection for polyhedral globes.
- DGGRID documentation: https://github.com/sahrk/DGGRID
