---
title: "Mathematical Foundations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hexify)
```

## Overview

hexify implements the **ISEA (Icosahedral Snyder Equal Area)** discrete global grid system. This vignette explains the mathematical foundations: the projection geometry, coordinate systems, aperture subdivision, and cell indexing.

## The Problem: Equal-Area Grids on a Sphere

Any projection from a sphere to a plane must distort *something*. For spatial statistics, we need cells of equal area regardless of location. Standard latitude-longitude grids fail badly: a 1° cell at the equator covers ~12,300 km², while the same cell near the poles covers a tiny fraction of that.

The ISEA projection solves this by:
1. Inscribing a regular icosahedron in the sphere
2. Projecting each spherical "cap" onto its corresponding flat triangular face using a modified Lambert equal-area projection
3. Overlaying a hexagonal grid on the resulting planar triangles

## The Lambert Azimuthal Equal-Area Projection

The foundation of Snyder's projection is Lambert's azimuthal equal-area projection, developed by Johann Heinrich Lambert in 1772.

### Geometric Definition

The key insight is elegantly simple. Consider a sphere with a plane tangent at point S:

```{r lambert-geometry, echo=FALSE, fig.width=7, fig.height=5.5}
# Accurate illustration of Lambert azimuthal equal-area projection
par(mar = c(1, 1, 2, 1), bg = "white")
plot(NULL, xlim = c(-2.5, 2.5), ylim = c(-1.2, 2.8), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Lambert Azimuthal Equal-Area Projection")

# Draw the sphere (circle in cross-section)
theta <- seq(0, 2*pi, length.out = 200)
R <- 1  # sphere radius
lines(R * cos(theta), R * sin(theta) + R, lwd = 2, col = "gray30")

# Draw the tangent plane (horizontal line at y = 0)
lines(c(-2.3, 2.3), c(0, 0), lwd = 2, col = "gray30")

# Mark the tangent point S (south pole of projection)
points(0, 0, pch = 19, cex = 1.5, col = "black")
text(0, -0.25, "S (tangent point)", cex = 0.9)

# Mark the center of sphere O
points(0, R, pch = 3, cex = 1.2, col = "gray50")
text(0.15, R + 0.15, "O", cex = 0.9, col = "gray50")

# Show a point P on the sphere and its projection P'
# Angular distance from S (colatitude) = 60 degrees
phi <- 60 * pi/180
P_x <- R * sin(phi)
P_y <- R * (1 - cos(phi))

# The chord distance d from S to P
d <- sqrt(P_x^2 + P_y^2)

# Lambert formula: projected distance = d (chord distance)
# Direction is same as azimuth from S
proj_x <- d  # projected onto plane at distance d from S

# Draw point P on sphere
points(P_x, P_y, pch = 19, cex = 1.3, col = "#E63946")
text(P_x + 0.15, P_y + 0.15, "P", cex = 1, col = "#E63946", font = 2)

# Draw the chord from S to P
lines(c(0, P_x), c(0, P_y), lwd = 2, col = "#E63946", lty = 2)

# Draw projected point P' on plane
points(proj_x, 0, pch = 19, cex = 1.3, col = "#457B9D")
text(proj_x, -0.25, "P'", cex = 1, col = "#457B9D", font = 2)

# Draw vertical dashed line from P down to show projection concept
lines(c(P_x, proj_x), c(P_y, 0), lwd = 1, col = "gray60", lty = 3)

# Label the chord distance
mid_x <- P_x / 2
mid_y <- P_y / 2
text(mid_x - 0.2, mid_y + 0.15, "d", cex = 1.1, col = "#E63946", font = 3)

# Label the projected distance (same as d)
text(proj_x / 2, -0.35, "d", cex = 1.1, col = "#457B9D", font = 3)

# Add angular distance arc
arc_r <- 0.4
arc_theta <- seq(3*pi/2, 3*pi/2 + phi, length.out = 30)
lines(arc_r * cos(arc_theta), R + arc_r * sin(arc_theta),
      lwd = 1.5, col = "gray50")
text(0.25, R - 0.55, expression(phi), cex = 1, col = "gray50")

# Second example point at different angle
phi2 <- 100 * pi/180
P2_x <- R * sin(phi2)
P2_y <- R * (1 - cos(phi2))
d2 <- sqrt(P2_x^2 + P2_y^2)

points(P2_x, P2_y, pch = 19, cex = 1.1, col = "#2A9D8F")
text(P2_x + 0.12, P2_y + 0.15, expression(P[2]), cex = 0.9, col = "#2A9D8F")
lines(c(0, P2_x), c(0, P2_y), lwd = 1.5, col = "#2A9D8F", lty = 2)
points(d2, 0, pch = 19, cex = 1.1, col = "#2A9D8F")
text(d2, -0.25, expression(P[2]*"'"), cex = 0.9, col = "#2A9D8F")

# Annotation box
rect(-2.4, 2.1, 2.4, 2.7, col = "gray95", border = NA)
text(0, 2.5, "Lambert's rule: Place P' at distance d from S,", cex = 0.9)
text(0, 2.25, "where d is the straight-line (chord) distance from S to P", cex = 0.9)
```

**The Lambert projection rule:**
For any point P on the sphere, measure the straight-line distance d through 3D space from the tangent point S to P (the chord distance). Place the projected point P' on the plane at distance d from S, in the same azimuthal direction.

### Why Chord Distance Preserves Area

For a point at angular distance $\phi$ from S (measured along the sphere surface), the chord distance is:

$$d = 2R \sin\left(\frac{\phi}{2}\right)$$

This specific relationship ensures that a small area element $dA$ on the sphere maps to an equal area $dA$ on the plane. The area-preserving property arises because the Jacobian determinant of this transformation equals 1.

```{r lambert-area-preservation, echo=FALSE, fig.width=8, fig.height=4}
# Show equal-area property with concentric rings
par(mfrow = c(1, 2), mar = c(2, 1, 3, 1), bg = "white")

R <- 1

# Left: Sphere view (orthographic, looking from above at tangent point)
plot(NULL, xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Sphere (view from above S)")

# Draw outer circle (equator as seen from S)
theta <- seq(0, 2*pi, length.out = 100)
lines(cos(theta), sin(theta), lwd = 2)

# Draw concentric latitude circles and shade bands
# Each band represents equal area on sphere
n_bands <- 5
cols <- c("#264653", "#2A9D8F", "#E9C46A", "#F4A261", "#E76F51")

for (i in n_bands:1) {
  # Angular distance from pole for this band
  phi_outer <- (i / n_bands) * (pi/2)
  phi_inner <- ((i-1) / n_bands) * (pi/2)

  # Project to plane coordinates (radius = sin of angular distance for orthographic)
  r_outer <- sin(phi_outer)
  r_inner <- sin(phi_inner)

  # Draw band
  theta_seq <- seq(0, 2*pi, length.out = 100)
  if (i == 1) {
    polygon(r_outer * cos(theta_seq), r_outer * sin(theta_seq),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  } else {
    polygon(c(r_outer * cos(theta_seq), rev(r_inner * cos(theta_seq))),
            c(r_outer * sin(theta_seq), rev(r_inner * sin(theta_seq))),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  }
}
points(0, 0, pch = 19, cex = 1.2)
text(0, -1.2, "Equal-area bands on sphere", cex = 0.85)

# Right: Lambert projection
plot(NULL, xlim = c(-1.6, 1.6), ylim = c(-1.6, 1.6), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Lambert Projection")

# Draw the same bands after Lambert projection
for (i in n_bands:1) {
  phi_outer <- (i / n_bands) * (pi/2)
  phi_inner <- ((i-1) / n_bands) * (pi/2)

  # Lambert projection: r = 2*sin(phi/2)
  r_outer <- 2 * sin(phi_outer / 2)
  r_inner <- 2 * sin(phi_inner / 2)

  theta_seq <- seq(0, 2*pi, length.out = 100)
  if (i == 1) {
    polygon(r_outer * cos(theta_seq), r_outer * sin(theta_seq),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  } else {
    polygon(c(r_outer * cos(theta_seq), rev(r_inner * cos(theta_seq))),
            c(r_outer * sin(theta_seq), rev(r_inner * sin(theta_seq))),
            col = adjustcolor(cols[i], 0.5), border = "gray40")
  }
}

# Draw maximum extent circle (equator projects to r = sqrt(2))
r_eq <- 2 * sin(pi/4)
lines(r_eq * cos(theta), r_eq * sin(theta), lwd = 2, lty = 2, col = "gray50")

points(0, 0, pch = 19, cex = 1.2)
text(0, -1.5, "Same bands after projection (areas preserved)", cex = 0.85)
```

Each colored band has equal area on the sphere. After Lambert projection, the shapes change (outer bands stretch radially, compress tangentially) but the areas remain equal. This is the defining property of an equal-area projection.

**Reference:** [Wolfram MathWorld: Lambert Azimuthal Equal-Area Projection](https://mathworld.wolfram.com/LambertAzimuthalEqual-AreaProjection.html)

## From Lambert to Snyder: The Icosahedron

Lambert's projection works for a single tangent plane covering at most a hemisphere. To cover the entire globe with minimal distortion, Snyder's key insight was to use **20 tangent planes**—one for each face of a regular icosahedron.

### The Icosahedron

A regular icosahedron has:
- **20 triangular faces** (each covering ~1/20 of Earth's surface ≈ 25.5 million km²)
- **12 vertices** (where 5 faces meet—these become pentagon cells)
- **30 edges**

```{r icosahedron-concept, echo=FALSE, fig.width=8, fig.height=4}
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1), bg = "white")

# Left: conceptual diagram
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Icosahedron inscribed in sphere")

# Draw sphere
theta <- seq(0, 2*pi, length.out = 100)
lines(cos(theta), sin(theta), lwd = 2, col = "gray70")

# Draw simplified icosahedron front view (5 visible triangular faces)
# Central pentagon
pent_r <- 0.38
pent_angles <- seq(pi/2, pi/2 + 2*pi, length.out = 6)[-6]
pent_x <- pent_r * cos(pent_angles)
pent_y <- pent_r * sin(pent_angles)

# Outer vertices
outer_r <- 0.95
outer_angles <- seq(pi/2 + pi/5, pi/2 + pi/5 + 2*pi, length.out = 6)[-6]
outer_x <- outer_r * cos(outer_angles)
outer_y <- outer_r * sin(outer_angles)

# Draw triangular faces
face_cols <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A", "#F4A261")
for (i in 1:5) {
  j <- (i %% 5) + 1
  polygon(c(pent_x[i], pent_x[j], outer_x[i]),
          c(pent_y[i], pent_y[j], outer_y[i]),
          col = adjustcolor(face_cols[i], 0.4),
          border = "gray30", lwd = 1.5)
}

# Central region
polygon(pent_x, pent_y, col = adjustcolor("gray80", 0.3),
        border = "gray30", lwd = 1.5)

# Mark vertices (future pentagons)
points(c(0, pent_x, outer_x), c(0, pent_y, outer_y), pch = 19, cex = 0.8)

text(0, -1.3, "Each triangular face = 1/20 of Earth", cex = 0.85)

# Right: show projection concept
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-0.5, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Projection to one face")

# Draw sphere arc (cross section)
arc_theta <- seq(-pi/3, pi/3, length.out = 50)
lines(1.3 * sin(arc_theta), 1.3 * cos(arc_theta) + 0.5, lwd = 2, col = "gray50")

# Draw triangular face (tangent plane)
tri_x <- c(-0.9, 0, 0.9, -0.9)
tri_y <- c(0, 1.5, 0, 0)
polygon(tri_x, tri_y, col = adjustcolor("#457B9D", 0.3),
        border = "#264653", lwd = 2)

# Mark face center
points(0, 0.5, pch = 19, cex = 1.2, col = "#E63946")
text(0.25, 0.5, "face center", cex = 0.8, adj = 0)

# Show points on sphere projecting down
sphere_pts <- data.frame(
  x = c(-0.4, 0.3, 0.1),
  y = c(1.1, 0.9, 1.3)
)
for (i in 1:nrow(sphere_pts)) {
  # Point on sphere
  points(sphere_pts$x[i], sphere_pts$y[i], pch = 19, cex = 0.8, col = "#2A9D8F")
  # Arrow down to plane
  arrows(sphere_pts$x[i], sphere_pts$y[i] - 0.1,
         sphere_pts$x[i] * 0.7, 0.4,
         length = 0.08, lwd = 1.5, col = "#2A9D8F")
}

text(0, -0.35, "Lambert projection applied to each face", cex = 0.85)
```

Because each face subtends a relatively small solid angle (~1/20 of the sphere), projection distortion within each face is limited. The maximum angular distortion is bounded at approximately 17.3°.

### Standard ISEA Orientation

The icosahedron orientation is chosen to place vertices (pentagon cells) in oceanic or low-population areas:

```{r face-centers, echo=FALSE, fig.width=8, fig.height=4}
library(sf)
library(ggplot2)

# Get face centers from hexify (returns radians)
centers <- hexify_face_centers()

# Create sf points (convert radians to degrees)
center_pts <- st_as_sf(
  data.frame(face = 0:19, lon = centers$lon * 180 / pi, lat = centers$lat * 180 / pi),
  coords = c("lon", "lat"), crs = 4326
)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = center_pts, color = "#E63946", size = 3) +
  geom_sf_text(data = center_pts, aes(label = face), nudge_y = 6, size = 2.5) +
  labs(title = "ISEA Icosahedron Face Centers",
       subtitle = "20 triangular faces, numbered 0-19") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

**Reference:** [Sahr, White & Kimerling (2003). Geodesic discrete global grid systems.](https://www.researchgate.net/publication/246557072_ISEA_discrete_global_grids)

## Snyder's Projection Formulas

Snyder adapted Lambert's approach for triangular faces. The projection involves:

1. Finding which face contains the point
2. Computing spherical coordinates relative to the face center
3. Applying transformation formulas that preserve area

### The Forward Projection

For a point P at geographic coordinates (lon, lat), relative to a face with center at (lon₀, lat₀):

**Step 1:** Compute angular distance z and azimuth Az from face center:
$$z = \arccos(\sin\phi_0 \sin\phi + \cos\phi_0 \cos\phi \cos(\lambda - \lambda_0))$$
$$Az = \arctan\left(\frac{\cos\phi \sin(\lambda - \lambda_0)}{\cos\phi_0 \sin\phi - \sin\phi_0 \cos\phi \cos(\lambda - \lambda_0)}\right)$$

**Step 2:** Reduce azimuth to [0°, 120°) sector and apply Snyder's transformation:
$$\delta_z = \arctan\left(\frac{\tan E_l}{\cos Az + \cot 30° \sin Az}\right)$$
$$h = \arccos(\sin Az \sin G \cos E_l - \cos Az \cos G)$$
$$A_G = Az + G + h - \pi$$
$$Az' = \arctan\left(\frac{2 A_G}{R_1^2 \tan^2 E_l - 2 A_G \cot 30°}\right)$$

**Step 3:** Compute radial distance:
$$f = \frac{\tan E_l}{2(\cos Az' + \cot 30° \sin Az') \sin(\delta_z/2)}$$
$$\rho = 2 R_1 f \sin(z/2)$$

**Step 4:** Convert to Cartesian:
$$x = \rho \sin Az' + x_0, \quad y = \rho \cos Az' + y_0$$

### Key Constants

| Constant | Value | Description |
|----------|-------|-------------|
| $E_l$ | 37.37736814° | Face center to edge angular distance |
| G | 36° | Geometric angle (360°/10) |
| $R_1$ | 0.9103832815 | Scaling factor for equal-area |

**Reference:** [Snyder (1992). An equal-area map projection for polyhedral globes.](https://www.researchgate.net/publication/296900899_Snyder_equal-area_map_projection_for_polyhedral_globes)

### The Inverse Projection (Newton-Raphson Iteration)

The inverse—converting planar coordinates back to geographic—cannot be solved analytically because the forward transformation involves transcendental functions. Instead, a Newton-Raphson iterative method is used.

```{r newton-raphson, echo=FALSE, fig.width=7, fig.height=5}
# Illustrate Newton-Raphson iteration concept
par(mar = c(4, 4, 3, 1), bg = "white")

# Define a sample function (simplified analog of the Snyder inverse problem)
f <- function(x) x^3 - 2*x - 2
f_prime <- function(x) 3*x^2 - 2

# Plot the function
x_seq <- seq(0.5, 2.5, length.out = 100)
plot(x_seq, f(x_seq), type = "l", lwd = 2, col = "gray30",
     xlab = "Azimuth estimate", ylab = "f(Az) - target",
     main = "Newton-Raphson Iteration for Inverse Projection",
     ylim = c(-3, 4))
abline(h = 0, lty = 2, col = "gray50")

# Show iteration steps
x0 <- 2.3  # initial guess
cols <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A")
x_current <- x0

for (i in 1:4) {
  y_current <- f(x_current)
  slope <- f_prime(x_current)
  x_next <- x_current - y_current / slope

  # Mark current point
  points(x_current, y_current, pch = 19, cex = 1.2, col = cols[i])
  text(x_current, y_current + 0.4, paste0("x", i-1), cex = 0.8, col = cols[i])

  # Draw tangent line to x-axis
  if (i < 4) {
    segments(x_current, y_current, x_next, 0, lty = 2, col = cols[i], lwd = 1.5)
    # Vertical line to next point
    segments(x_next, 0, x_next, f(x_next), lty = 3, col = cols[i+1])
  }

  x_current <- x_next
}

# Mark the root
root <- 1.769  # approximate root
points(root, 0, pch = 4, cex = 2, lwd = 2, col = "black")
text(root, -0.5, "solution", cex = 0.9)

legend("topleft",
       legend = c("f(Az)", "Iteration steps", "Converged solution"),
       lty = c(1, 2, NA), pch = c(NA, 19, 4),
       col = c("gray30", "#457B9D", "black"),
       lwd = c(2, 1.5, 2), bty = "n", cex = 0.85)
```

The iteration solves for the azimuth Az given the planar coordinates. Starting from an initial estimate, each iteration refines:

$$Az_{n+1} = Az_n - \frac{f(Az_n)}{f'(Az_n)}$$

Convergence typically occurs within 3-5 iterations to machine precision.

**Reference:** [PROJ Documentation: Icosahedral Snyder Equal Area](https://proj.org/en/stable/operations/projections/isea.html)

## Aperture and Resolution

**Aperture** determines how cells subdivide at each resolution level—it's the ratio of parent cell area to child cell area.

```{r aperture-diagrams, echo=FALSE, fig.width=9, fig.height=3.5}
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1), bg = "white")

# Helper for hexagon vertices
hex_vertices <- function(cx, cy, r, flat_top = TRUE) {
  offset <- if (flat_top) 0 else pi/6
  angles <- seq(offset, 2*pi + offset, length.out = 7)
  list(x = cx + r * cos(angles), y = cy + r * sin(angles))
}

# Aperture 3
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Aperture 3: Triangle arrangement")

# Parent hexagon outline
h_parent <- hex_vertices(0, 0, 1.7, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "gray40", lwd = 2, col = NA)

# 3 child cells in triangular arrangement
child_r <- 0.85
child_positions <- list(
  c(0, 0.6),           # top
  c(-0.52, -0.3),      # bottom-left
  c(0.52, -0.3)        # bottom-right
)
child_cols <- c("#E63946", "#457B9D", "#2A9D8F")

for (i in 1:3) {
  h <- hex_vertices(child_positions[[i]][1], child_positions[[i]][2],
                    child_r, flat_top = FALSE)  # rotated 30° from parent
  polygon(h$x, h$y, col = adjustcolor(child_cols[i], 0.5),
          border = "gray30", lwd = 1.5)
  text(child_positions[[i]][1], child_positions[[i]][2], i,
       cex = 1.2, font = 2)
}
text(0, -1.85, expression("Cell count: 10" %*% 3^res + 2), cex = 0.85)

# Aperture 4
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Aperture 4: Square arrangement")

h_parent <- hex_vertices(0, 0, 1.7, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "gray40", lwd = 2, col = NA)

# 4 child cells in 2x2 arrangement
child_r4 <- 0.7
positions4 <- list(
  c(-0.45, 0.45), c(0.45, 0.45),
  c(-0.45, -0.45), c(0.45, -0.45)
)
child_cols4 <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A")

for (i in 1:4) {
  h <- hex_vertices(positions4[[i]][1], positions4[[i]][2],
                    child_r4, flat_top = TRUE)  # same orientation as parent
  polygon(h$x, h$y, col = adjustcolor(child_cols4[i], 0.5),
          border = "gray30", lwd = 1.5)
  text(positions4[[i]][1], positions4[[i]][2], i, cex = 1.2, font = 2)
}
text(0, -1.85, expression("Cell count: 10" %*% 4^res + 2), cex = 0.85)

# Aperture 7
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Aperture 7: Rosette arrangement")

h_parent <- hex_vertices(0, 0, 1.7, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "gray40", lwd = 2, col = NA)

# 7 child cells: 1 center + 6 ring (rotated ~19.1° from parent)
child_r7 <- 0.52
# Center cell
h <- hex_vertices(0, 0, child_r7, flat_top = TRUE)
polygon(h$x, h$y, col = adjustcolor("#F4A261", 0.5),
        border = "gray30", lwd = 1.5)
text(0, 0, "1", cex = 1, font = 2)

# Ring of 6
ring_r <- 0.82
ring_cols <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A", "#9B5DE5", "#00BBF9")
for (i in 1:6) {
  ang <- pi/2 + (i-1) * pi/3
  cx <- ring_r * cos(ang)
  cy <- ring_r * sin(ang)
  # Rotated orientation (characteristic of aperture 7)
  h <- hex_vertices(cx, cy, child_r7, flat_top = TRUE)
  polygon(h$x, h$y, col = adjustcolor(ring_cols[i], 0.5),
          border = "gray30", lwd = 1.5)
  text(cx, cy, i+1, cex = 0.9, font = 2)
}
text(0, -1.85, expression("Cell count: 10" %*% 7^res + 2), cex = 0.85)
```

| Aperture | Children/Parent | Area Ratio | Best For |
|----------|-----------------|------------|----------|
| 3 | 3 | 3:1 | Fine resolution control, dggridR compatibility |
| 4 | 4 | 4:1 | Power-of-2 scaling, GIS workflows |
| 7 | 7 | 7:1 | Rapid cell growth, coarse analysis |

### Cell Count Growth

```{r cell-counts}
cat("Resolution  Aperture 3    Aperture 4    Aperture 7\n")
cat("---------  ----------    ----------    ----------\n")
for (res in 0:8) {
  cells_ap3 <- 10 * 3^res + 2
  cells_ap4 <- 10 * 4^res + 2
  cells_ap7 <- 10 * 7^res + 2
  cat(sprintf("    %d      %10s    %10s    %10s\n",
              res,
              format(cells_ap3, big.mark = ","),
              format(cells_ap4, big.mark = ","),
              format(cells_ap7, big.mark = ",")))
}
```

**Reference:** [Location coding on icosahedral aperture 3 hexagon discrete global grids](https://www.researchgate.net/publication/223763428_Location_coding_on_icosahedral_aperture_3_hexagon_discrete_global_grids)

## Orientation Classes

The aperture determines how child cells align with parent cells:

```{r orientation-classes, echo=FALSE, fig.width=8, fig.height=3}
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1), bg = "white")

hex_v <- function(cx, cy, r, rotation = 0) {
  angles <- seq(rotation, 2*pi + rotation, length.out = 7)
  list(x = cx + r * cos(angles), y = cy + r * sin(angles))
}

# Class I: Flat-top
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Class I (Flat-top)")
h <- hex_v(0, 0, 1.2, rotation = 0)
polygon(h$x, h$y, col = adjustcolor("#457B9D", 0.4), border = "gray30", lwd = 2)
# Edge at top
lines(h$x[1:2], h$y[1:2], col = "#E63946", lwd = 3)
text(0, -1.35, "Aperture 4 (all res)\nAperture 3 (odd res)", cex = 0.8)

# Class II: Pointy-top (30° rotation)
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Class II (Pointy-top)")
h <- hex_v(0, 0, 1.2, rotation = pi/6)
polygon(h$x, h$y, col = adjustcolor("#2A9D8F", 0.4), border = "gray30", lwd = 2)
# Vertex at top
points(h$x[1], h$y[1], pch = 19, cex = 1.5, col = "#E63946")
text(0, -1.35, "Aperture 3 (even res)", cex = 0.8)

# Class III: Skewed (arbitrary rotation)
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Class III (Skewed)")
# Show parent and child with different rotations
h_parent <- hex_v(0, 0, 1.2, rotation = 0)
polygon(h_parent$x, h_parent$y, col = NA, border = "gray50", lwd = 2, lty = 2)
# Child rotated by arctan(sqrt(3)/5) ≈ 19.1°
rot_angle <- atan(sqrt(3)/5)
h_child <- hex_v(0, 0, 0.8, rotation = rot_angle)
polygon(h_child$x, h_child$y, col = adjustcolor("#E9C46A", 0.4),
        border = "gray30", lwd = 2)
# Show rotation angle
arc_r <- 0.5
arc_theta <- seq(0, rot_angle, length.out = 20)
lines(arc_r * cos(arc_theta), arc_r * sin(arc_theta), col = "#E63946", lwd = 2)
text(0.6, 0.15, "19.1°", cex = 0.8, col = "#E63946")
text(0, -1.35, "Aperture 7 (all res)\nRotation accumulates", cex = 0.8)
```

For **aperture 3**, orientation alternates between Class I and II at each resolution level.
For **aperture 7**, each level adds a rotation of arctan(√3/5) ≈ 19.1°.

## Pentagon Cells

At every resolution, exactly **12 cells are pentagons** (not hexagons). These occur at the icosahedron vertices and have area exactly 5/6 that of hexagonal cells.

```{r pentagon-locations, echo=FALSE, fig.width=8, fig.height=4}
# Pentagon locations
pentagon_coords <- data.frame(
  type = c("Pole", "Pole", rep("Upper ring", 5), rep("Lower ring", 5)),
  lon = c(0, 0, 0, 72, 144, 216, 288, 36, 108, 180, 252, 324),
  lat = c(90, -90, rep(26.57, 5), rep(-26.57, 5))
)

pentagon_pts <- st_as_sf(pentagon_coords, coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = hexify_world, fill = "gray95", color = "gray70", linewidth = 0.2) +
  geom_sf(data = pentagon_pts, aes(color = type), size = 4) +
  scale_color_manual(values = c("Pole" = "#E63946",
                                 "Upper ring" = "#457B9D",
                                 "Lower ring" = "#2A9D8F")) +
  labs(title = "Pentagon Cell Locations",
       subtitle = "12 pentagonal cells at icosahedron vertices (area = 5/6 of hexagons)",
       color = "Location") +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

| Location | Latitude | Longitudes |
|----------|----------|------------|
| Poles | ±90° | 0° |
| Upper ring | 26.57° | 0°, 72°, 144°, 216°, 288° |
| Lower ring | -26.57° | 36°, 108°, 180°, 252°, 324° |

The latitude 26.57° = arctan(1/2) arises from the geometry of a regular icosahedron.

## Coordinate Systems

hexify uses several coordinate systems internally:

| System | Components | Description |
|--------|------------|-------------|
| **GEO** | lon, lat | WGS84 degrees |
| **Icosa Triangle** | face (0-19), tx, ty | Triangle face + normalized coords [0,1] |
| **Quad IJ** | quad (0-11), i, j | Quad + integer grid indices |
| **SEQNUM** | integer | Global cell ID (1-based, dggridR compatible) |

### Triangle to Quad

The 20 triangular faces are paired into 12 "quads" (rhombi). Each quad combines two adjacent triangles sharing an edge, simplifying the grid indexing.

## Round-Trip Accuracy

```{r round-trip}
# Round-trip accuracy test
original_lon <- 16.37
original_lat <- 48.21

cat(sprintf("Original: (%.4f, %.4f)\n\n", original_lon, original_lat))

for (ap in c(3, 4, 7)) {
  res <- if (ap == 7) 6 else 10
  grid <- hex_grid(resolution = res, aperture = ap)
  cell_id <- lonlat_to_cell(original_lon, original_lat, grid)
  recovered <- cell_to_lonlat(cell_id, grid)
  error_km <- sqrt((recovered$lon - original_lon)^2 +
                   (recovered$lat - original_lat)^2) * 111
  cat(sprintf("Aperture %d (res %2d): cell %d -> (%.4f, %.4f), ~%.1f km from center\n",
              ap, res, cell_id,
              recovered$lon, recovered$lat, error_km))
}
```

## Summary of ISEA Properties

| Property | Description |
|----------|-------------|
| **Equal area** | All hexagonal cells identical; pentagons = 5/6 hex area |
| **Bounded distortion** | Max angular distortion ≈ 17.3° at face edges |
| **Uniform topology** | Hexagons have 6 neighbors; pentagons have 5 |
| **No perfect nesting** | Cells at different resolutions partially overlap |

## References

- Snyder, J. P. (1992). [An equal-area map projection for polyhedral globes](https://www.researchgate.net/publication/296900899_Snyder_equal-area_map_projection_for_polyhedral_globes). *Cartographica*, 29(1), 10-21.

- Sahr, K., White, D., & Kimerling, A. J. (2003). [Geodesic discrete global grid systems](https://www.researchgate.net/publication/246557072_ISEA_discrete_global_grids). *Cartography and Geographic Information Science*, 30(2), 121-134.

- [Wolfram MathWorld: Lambert Azimuthal Equal-Area Projection](https://mathworld.wolfram.com/LambertAzimuthalEqual-AreaProjection.html)

- [PROJ Documentation: Icosahedral Snyder Equal Area](https://proj.org/en/stable/operations/projections/isea.html)

- [DGGRID software](https://github.com/sahrk/DGGRID) by Kevin Sahr
