---
title: "Mathematical Foundations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

hexify implements the **ISEA (Icosahedral Snyder Equal Area)** discrete global
grid system. This vignette explains the mathematical foundations with intuitive
geometric illustrations.

## The Problem: Mapping a Sphere to a Plane

The fundamental challenge in cartography is mapping the curved surface of Earth
onto a flat plane. Any such mapping must distort *something*—you cannot
perfectly preserve both area and shape simultaneously. For ecological and
statistical applications, **equal-area** projections are essential: we need
grid cells of equal size regardless of location.

### Why Equal-Area Matters

Consider counting species in grid cells:

- If cells vary in area, counts aren't comparable
- A cell twice as large will, on average, contain twice as many individuals
- Standard lat/lon grids fail badly—cells near poles are tiny compared to equatorial cells

The ISEA projection solves this by preserving area while minimizing shape distortion.

## The Lambert Azimuthal Equal-Area Projection

Before understanding Snyder's projection, we need to understand its foundation:
the **Lambert azimuthal equal-area projection**, developed by Johann Heinrich
Lambert in 1772.

### The Core Idea: Chord Distance

The key insight is beautifully simple. Imagine a sphere with a flat plane
tangent to it at point S (the "south pole" in polar aspect):

```{r lambert-concept, echo=FALSE, fig.width=6, fig.height=5, fig.cap="Lambert projection: Points on the sphere are mapped to the plane using chord distance."}
# Geometric illustration of Lambert azimuthal equal-area projection
par(mar = c(1, 1, 2, 1), bg = "white")
plot(NULL, xlim = c(-2.5, 2.5), ylim = c(-1.5, 2.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Lambert Azimuthal Equal-Area Projection")

# Draw the sphere (circle)
theta <- seq(0, 2*pi, length.out = 100)
lines(cos(theta), sin(theta) + 1, lwd = 2)

# Draw the tangent plane (horizontal line at bottom)
lines(c(-2.5, 2.5), c(0, 0), lwd = 2)

# Mark the tangent point S
points(0, 0, pch = 19, cex = 1.5)
text(0, -0.2, "S (tangent point)", cex = 0.9, pos = 1)

# Mark the center of sphere
points(0, 1, pch = 3, cex = 1)
text(0.2, 1, "O", cex = 0.9)

# Show several points on sphere and their projections
angles <- c(30, 60, 90, 120) * pi/180  # angles from bottom
colors <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A")

for (i in seq_along(angles)) {
  ang <- angles[i]
  # Point on sphere
  px <- sin(ang)
  py <- 1 - cos(ang)

  # Chord distance from S to P
  d <- sqrt(px^2 + py^2)

  # Projected point on plane (at distance d from S)
  proj_x <- d * sign(px)

  # Draw the point on sphere
  points(px, py, pch = 19, cex = 1.2, col = colors[i])

  # Draw chord from S to P (dashed)
  lines(c(0, px), c(0, py), lty = 2, col = colors[i], lwd = 1.5)

  # Draw projection line down to plane
  lines(c(px, proj_x), c(py, 0), lty = 3, col = colors[i], lwd = 1.5)

  # Mark projected point
  points(proj_x, 0, pch = 19, cex = 1.2, col = colors[i])

  # Label
  text(px + 0.15, py + 0.1, paste0("P", i), cex = 0.8, col = colors[i])
  text(proj_x, -0.15, paste0("P'", i), cex = 0.8, col = colors[i], pos = 1)
}

# Add annotation
text(-1.8, 2.2, "d = chord distance", cex = 0.85, adj = 0)
text(-1.8, 1.9, "P' placed at distance d from S", cex = 0.85, adj = 0)

# Legend for line types
legend("topright",
       legend = c("Chord (length = d)", "Projection to plane"),
       lty = c(2, 3), lwd = 1.5, cex = 0.8, bty = "n")
```

**How it works:**

1. Pick any point **P** on the sphere
2. Measure the straight-line (chord) distance **d** from **S** to **P**
3. Place the projected point **P'** on the plane at distance **d** from **S**,
   in the same direction

This simple rule—*use chord distance*—automatically preserves area! The
mathematical magic is that the area element transforms correctly.

### Why Does This Preserve Area?

The chord distance formula for a point at angular distance φ from S is:

$$d = 2R \sin\left(\frac{\phi}{2}\right)$$

where R is the sphere's radius. This specific relationship ensures that a small
patch of area dA on the sphere maps to an equal area dA on the plane.

```{r lambert-area, echo=FALSE, fig.width=7, fig.height=4, fig.cap="Equal-area property: Concentric circles on the sphere map to circles on the plane with areas preserved."}
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1), bg = "white")

# Left: sphere view (looking from above)
plot(NULL, xlim = c(-1.3, 1.3), ylim = c(-1.3, 1.3), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Sphere (polar view)")

# Draw concentric latitude circles
for (lat in seq(15, 75, by = 15)) {
  r <- cos(lat * pi/180)  # radius at this latitude
  theta <- seq(0, 2*pi, length.out = 100)
  lines(r * cos(theta), r * sin(theta),
        col = gray(0.3 + lat/150), lwd = 1.5)
}
# Outer circle (equator)
lines(cos(theta), sin(theta), lwd = 2)
points(0, 0, pch = 19, cex = 1)
text(0, 0.15, "Pole", cex = 0.8)

# Color bands to show equal-area strips
cols <- c("#264653", "#2A9D8F", "#E9C46A", "#F4A261", "#E76F51")
for (i in 1:5) {
  lat1 <- (i-1) * 15
  lat2 <- i * 15
  r1 <- cos(lat1 * pi/180)
  r2 <- cos(lat2 * pi/180)
  # Draw as polygon
  theta_seq <- seq(0, 2*pi, length.out = 100)
  polygon(c(r1*cos(theta_seq), rev(r2*cos(theta_seq))),
          c(r1*sin(theta_seq), rev(r2*sin(theta_seq))),
          col = adjustcolor(cols[i], 0.3), border = NA)
}

# Right: Lambert projection
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Lambert Projection")

# Project the same latitude circles
for (lat in seq(15, 75, by = 15)) {
  # Lambert radius: d = 2*sin(colatitude/2) for unit sphere
  colat <- (90 - lat) * pi/180
  r_proj <- 2 * sin(colat/2)
  theta <- seq(0, 2*pi, length.out = 100)
  lines(r_proj * cos(theta), r_proj * sin(theta),
        col = gray(0.3 + lat/150), lwd = 1.5)
}
# Equator projection
r_eq <- 2 * sin(pi/4)  # 90° colatitude
lines(r_eq * cos(theta), r_eq * sin(theta), lwd = 2)
points(0, 0, pch = 19, cex = 1)
text(0, 0.15, "Pole", cex = 0.8)

# Same colored bands
for (i in 1:5) {
  lat1 <- (i-1) * 15
  lat2 <- i * 15
  colat1 <- (90 - lat1) * pi/180
  colat2 <- (90 - lat2) * pi/180
  r1 <- 2 * sin(colat1/2)
  r2 <- 2 * sin(colat2/2)
  theta_seq <- seq(0, 2*pi, length.out = 100)
  polygon(c(r1*cos(theta_seq), rev(r2*cos(theta_seq))),
          c(r1*sin(theta_seq), rev(r2*sin(theta_seq))),
          col = adjustcolor(cols[i], 0.3), border = NA)
}

# Note about equal areas
mtext("Each colored band has the same area in both views",
      side = 1, line = 0, outer = TRUE, cex = 0.9)
```

The colored bands have equal area on the sphere. After Lambert projection,
the shapes change (bands near the edge are stretched radially, compressed
tangentially), but the *areas* remain equal.

## From Lambert to Snyder: Projecting onto an Icosahedron

Lambert's projection works for a single tangent plane. But what if we want to
cover the entire globe with minimal distortion? Snyder's insight: use **20
tangent planes**, one for each face of an icosahedron.

### The Icosahedron Advantage

```{r icosahedron-concept, echo=FALSE, fig.width=7, fig.height=4, fig.cap="An icosahedron inscribed in a sphere provides 20 nearly-flat triangular faces."}
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1), bg = "white")

# Left: show concept of single face projection
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-0.5, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Single Face Projection")

# Draw sphere arc
theta <- seq(-pi/3, pi/3, length.out = 50)
lines(1.5*sin(theta), 1.5*cos(theta) + 0.3, lwd = 2, col = "gray50")

# Draw triangular face
tri_x <- c(-0.8, 0, 0.8, -0.8)
tri_y <- c(0, 1.3, 0, 0)
polygon(tri_x, tri_y, col = adjustcolor("#457B9D", 0.3), border = "#457B9D", lwd = 2)

# Show center point
points(0, 0.43, pch = 19, cex = 1)
text(0.15, 0.43, "face center", cex = 0.8, adj = 0)

# Arrow showing projection direction
arrows(0, 1.7, 0, 1.4, length = 0.1, lwd = 2, col = "#E63946")
text(0, 1.85, "project down", cex = 0.8)

# Right: show 20 faces covering globe
plot(NULL, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "20 Faces Cover the Globe")

# Draw circle for globe
theta <- seq(0, 2*pi, length.out = 100)
lines(cos(theta), sin(theta), lwd = 2)

# Draw simplified icosahedron view (front faces)
# Pentagon in center
pent_angles <- seq(pi/2, pi/2 + 2*pi, length.out = 6)
pent_r <- 0.4
pent_x <- pent_r * cos(pent_angles)
pent_y <- pent_r * sin(pent_angles)

# Draw triangular faces radiating from center
outer_r <- 0.95
for (i in 1:5) {
  mid_ang <- (pent_angles[i] + pent_angles[i+1])/2
  tri_x <- c(pent_x[i], pent_x[i+1], outer_r*cos(mid_ang), pent_x[i])
  tri_y <- c(pent_y[i], pent_y[i+1], outer_r*sin(mid_ang), pent_y[i])
  col_idx <- (i %% 3) + 1
  cols <- c("#E63946", "#457B9D", "#2A9D8F")
  polygon(tri_x, tri_y, col = adjustcolor(cols[col_idx], 0.4),
          border = "gray30", lwd = 1)
}

# Central pentagon
polygon(pent_x, pent_y, col = adjustcolor("#E9C46A", 0.5),
        border = "gray30", lwd = 1)

text(0, -1.3, "Each triangle ≈ 1/20 of Earth's surface", cex = 0.85)
```

Why an icosahedron?

1. **20 identical faces**: Each triangular face covers ~1/20 of Earth (about 25.5 million km²)
2. **Low distortion**: Because each face is small and nearly flat, projection distortion stays below 17.3°
3. **Triangles subdivide naturally**: Each triangle can be recursively divided into smaller triangles

### Snyder's Projection: Modified Lambert for Triangles

Snyder adapted Lambert's method for projecting spherical triangles (the curved
patches on the globe) to planar triangles (the flat icosahedron faces):

```{r snyder-process, echo=FALSE, fig.width=7, fig.height=5, fig.cap="Snyder's projection maps curved spherical triangles to flat planar triangles while preserving area."}
par(mar = c(1, 1, 2, 1), bg = "white")
plot(NULL, xlim = c(-0.5, 10), ylim = c(-1, 5), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Snyder Equal-Area Projection Process")

# Step 1: Spherical triangle (curved edges)
# Draw as if looking at curved surface
curve_pts <- function(x1, y1, x2, y2, bulge = 0.15) {
  t <- seq(0, 1, length.out = 30)
  mid_x <- (x1 + x2)/2
  mid_y <- (y1 + y2)/2
  # Perpendicular offset for bulge
  dx <- x2 - x1
  dy <- y2 - y1
  len <- sqrt(dx^2 + dy^2)
  nx <- -dy/len * bulge * len
  ny <- dx/len * bulge * len

  # Quadratic bezier-like curve
  x <- (1-t)^2 * x1 + 2*(1-t)*t*(mid_x + nx) + t^2 * x2
  y <- (1-t)^2 * y1 + 2*(1-t)*t*(mid_y + ny) + t^2 * y2
  list(x = x, y = y)
}

# Spherical triangle vertices
s_v <- list(c(0.5, 4), c(2.5, 4), c(1.5, 2))
# Draw curved edges
for (i in 1:3) {
  j <- i %% 3 + 1
  crv <- curve_pts(s_v[[i]][1], s_v[[i]][2], s_v[[j]][1], s_v[[j]][2], 0.12)
  lines(crv$x, crv$y, lwd = 2, col = "#264653")
}
# Fill
polygon(c(s_v[[1]][1], s_v[[2]][1], s_v[[3]][1]),
        c(s_v[[1]][2], s_v[[2]][2], s_v[[3]][2]),
        col = adjustcolor("#2A9D8F", 0.3), border = NA)
points(1.5, 3.3, pch = 19, cex = 0.8)
text(1.5, 3.5, "P", cex = 0.9)
text(1.5, 1.5, "Spherical triangle\n(on globe)", cex = 0.85)

# Arrow
arrows(3, 3, 4, 3, length = 0.15, lwd = 2, col = "#E63946")
text(3.5, 3.5, "Snyder\nprojection", cex = 0.8)

# Step 2: Planar triangle (straight edges)
p_v <- list(c(4.5, 4), c(6.5, 4), c(5.5, 2))
polygon(c(p_v[[1]][1], p_v[[2]][1], p_v[[3]][1]),
        c(p_v[[1]][2], p_v[[2]][2], p_v[[3]][2]),
        col = adjustcolor("#457B9D", 0.3), border = "#264653", lwd = 2)
points(5.5, 3.3, pch = 19, cex = 0.8)
text(5.5, 3.5, "P'", cex = 0.9)
text(5.5, 1.5, "Planar triangle\n(on icosahedron face)", cex = 0.85)

# Arrow
arrows(7, 3, 8, 3, length = 0.15, lwd = 2, col = "#E63946")
text(7.5, 3.5, "Grid\noverlay", cex = 0.8)

# Step 3: Hexagonal grid
hex_center <- function(cx, cy, r) {
  angles <- seq(0, 2*pi, length.out = 7) + pi/6
  list(x = cx + r * cos(angles), y = cy + r * sin(angles))
}

# Draw triangle with hex grid
p_v2 <- list(c(8.5, 4), c(10.5, 4), c(9.5, 2))
polygon(c(p_v2[[1]][1], p_v2[[2]][1], p_v2[[3]][1]),
        c(p_v2[[1]][2], p_v2[[2]][2], p_v2[[3]][2]),
        col = adjustcolor("#E9C46A", 0.2), border = "#264653", lwd = 2)

# Add small hexagons
hex_size <- 0.25
hex_positions <- list(
  c(9.5, 3.2), c(9.2, 3.6), c(9.8, 3.6),
  c(9.5, 2.8), c(9.2, 2.4), c(9.8, 2.4)
)
for (pos in hex_positions) {
  hx <- hex_center(pos[1], pos[2], hex_size)
  polygon(hx$x, hx$y, border = "#264653", lwd = 1,
          col = adjustcolor("#F4A261", 0.3))
}
text(9.5, 1.5, "Hexagonal cells\n(equal area)", cex = 0.85)

# Area preservation note
rect(-0.3, -0.5, 10.3, 0.3, col = adjustcolor("#E8E8E8", 0.5), border = NA)
text(5, -0.1, "Area is preserved at each step: A(sphere patch) = A(plane patch) = Σ A(cells)",
     cex = 0.9, font = 3)
```

**The Snyder projection process:**

1. **Subdivide each spherical face** into 6 smaller triangles (from center to vertices)
2. **Apply modified Lambert** to each sub-triangle, preserving area
3. **Reassemble** the planar triangles to form the flat icosahedron face

The mathematical details involve computing intersection points of great circles
and carefully chosen scaling factors, but the principle is the same as
Lambert: preserve area by using the right distance relationship.

## The Complete Projection Pipeline

Now let's see how hexify implements this:

### Step 1: Determine the Face

For any point on Earth, first find which of the 20 icosahedral faces contains it.
The algorithm computes the angular distance from the point to each face center
and selects the closest face.

### Step 2: Project to Face Coordinates

Apply Snyder's equal-area formulas to compute (x, y) coordinates on the planar
triangle. The formulas involve:

- Computing the azimuth angle from face center to point
- Computing the angular distance from face center
- Applying the area-preserving transformation

### Step 3: Normalize and Quantize

Convert the (x, y) coordinates to normalized (tx, ty) in range [0, 1], then
quantize to grid cell indices based on the resolution.

```{r}
library(hexify)

# Project a point through the pipeline
result <- hexify_forward(lon = 16.37, lat = 48.21)
cat(sprintf("Face: %d, tx: %.4f, ty: %.4f\n",
            result["face"], result["tx"], result["ty"]))
```

### Icosahedron Orientation

The icosahedron can be oriented arbitrarily on the globe. The default ISEA orientation places vertex 0 at:

- Longitude: 45°/4
- Latitude: arctan(√5/2)

This orientation was chosen by Sahr et al. (2003) to minimize distortion over land masses while placing the 12 pentagonal cells (which occur at icosahedron vertices) in oceanic or low-population areas. The specific vertex latitude arctan(√5/2) arises from the geometry of a regular icosahedron inscribed in a sphere.

```{r}
# Get all face centers
centers <- hexify_face_centers()
head(centers, 5)
```

## Hexagonal Grid Generation

### From Triangles to Hexagons

Each triangular face is subdivided into hexagonal cells:

```{r hexagon-subdivision, echo=FALSE, fig.width=7, fig.height=5, fig.cap="Hexagonal grid overlaid on a triangular face. Hexagons tile the plane efficiently with equal-area cells."}
par(mar = c(1, 1, 2, 1), bg = "white")
plot(NULL, xlim = c(-0.5, 8), ylim = c(-0.5, 6), asp = 1,
     axes = FALSE, xlab = "", ylab = "",
     main = "Hexagonal Subdivision of Triangle")

# Helper function for hexagon vertices
hex_vertices <- function(cx, cy, r, flat_top = TRUE) {
  if (flat_top) {
    angles <- seq(0, 2*pi, length.out = 7)
  } else {
    angles <- seq(pi/6, 2*pi + pi/6, length.out = 7)
  }
  list(x = cx + r * cos(angles), y = cy + r * sin(angles))
}

# Draw the triangle boundary
tri_x <- c(0, 7, 3.5, 0)
tri_y <- c(0, 0, 6, 0)
polygon(tri_x, tri_y, border = "#264653", lwd = 3, col = NA)

# Create hexagonal grid
hex_r <- 0.5
hex_h <- hex_r * sqrt(3)

# Generate hex centers within triangle
hex_centers <- list()
for (row in 0:8) {
  y <- row * hex_h * 0.75
  offset <- if (row %% 2 == 0) 0 else hex_r * 0.75
  for (col in 0:12) {
    x <- col * hex_r * 1.5 + offset
    # Check if center is inside triangle (with margin)
    # Triangle vertices: (0,0), (7,0), (3.5,6)
    # Use barycentric coordinates
    denom <- (tri_y[2] - tri_y[3]) * (tri_x[1] - tri_x[3]) +
             (tri_x[3] - tri_x[2]) * (tri_y[1] - tri_y[3])
    a <- ((tri_y[2] - tri_y[3]) * (x - tri_x[3]) +
          (tri_x[3] - tri_x[2]) * (y - tri_y[3])) / denom
    b <- ((tri_y[3] - tri_y[1]) * (x - tri_x[3]) +
          (tri_x[1] - tri_x[3]) * (y - tri_y[3])) / denom
    c <- 1 - a - b
    margin <- 0.08
    if (a > margin && b > margin && c > margin) {
      hex_centers[[length(hex_centers) + 1]] <- c(x, y)
    }
  }
}

# Draw hexagons with alternating colors
colors <- c("#2A9D8F", "#E9C46A", "#F4A261")
for (i in seq_along(hex_centers)) {
  hx <- hex_vertices(hex_centers[[i]][1], hex_centers[[i]][2], hex_r * 0.95, flat_top = FALSE)
  col_idx <- (i %% 3) + 1
  polygon(hx$x, hx$y, col = adjustcolor(colors[col_idx], 0.4),
          border = "#264653", lwd = 0.8)
}

# Mark a few cell centers
set.seed(42)
sample_idx <- sample(seq_along(hex_centers), min(5, length(hex_centers)))
for (idx in sample_idx) {
  points(hex_centers[[idx]][1], hex_centers[[idx]][2], pch = 19, cex = 0.6)
}

# Label
text(3.5, -0.4, "Triangle face with hexagonal tessellation", cex = 0.9)

# Annotation
text(6.5, 5, "Each hexagon\n= equal area cell", cex = 0.8, adj = 0)
```

The hexagonal grid provides several advantages:

1. **Uniform neighbors**: Each hexagon has exactly 6 neighbors (unlike squares with 8)
2. **Isotropic**: Distance to all neighbors is equal
3. **Efficient packing**: Hexagons cover the plane with minimal perimeter per area

### Aperture and Resolution

**Aperture** determines how many child cells fit in one parent cell. This is
the key parameter controlling grid resolution:

```{r aperture-visual, echo=FALSE, fig.width=8, fig.height=4, fig.cap="Aperture controls how parent cells subdivide into children. Aperture 3 divides into 3 children, aperture 4 into 4, and aperture 7 into 7."}
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1), bg = "white")

# Helper for hexagon
hex_v <- function(cx, cy, r, flat_top = FALSE) {
  offset <- if (flat_top) 0 else pi/6
  angles <- seq(offset, 2*pi + offset, length.out = 7)
  list(x = cx + r * cos(angles), y = cy + r * sin(angles))
}

# Aperture 3
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Aperture 3")

# Parent hexagon (outline)
h_parent <- hex_v(0, 0, 1.8, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "#264653", lwd = 2, col = NA)

# Three child hexagons (triangular arrangement)
child_r <- 0.9
child_dist <- 0.7
child_angles <- c(pi/2, pi/2 + 2*pi/3, pi/2 + 4*pi/3)
child_cols <- c("#E63946", "#457B9D", "#2A9D8F")
for (i in 1:3) {
  cx <- child_dist * cos(child_angles[i])
  cy <- child_dist * sin(child_angles[i])
  h <- hex_v(cx, cy, child_r, flat_top = TRUE)
  polygon(h$x, h$y, col = adjustcolor(child_cols[i], 0.5),
          border = "#264653", lwd = 1.5)
  text(cx, cy, i, cex = 1.2, font = 2)
}
text(0, -1.9, "3 children per parent", cex = 0.85)

# Aperture 4
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Aperture 4")

# Parent hexagon
h_parent <- hex_v(0, 0, 1.8, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "#264653", lwd = 2, col = NA)

# Four child hexagons (2x2 arrangement)
child_r4 <- 0.75
positions <- list(c(-0.5, 0.5), c(0.5, 0.5), c(-0.5, -0.5), c(0.5, -0.5))
child_cols4 <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A")
for (i in 1:4) {
  cx <- positions[[i]][1]
  cy <- positions[[i]][2]
  h <- hex_v(cx, cy, child_r4, flat_top = TRUE)
  polygon(h$x, h$y, col = adjustcolor(child_cols4[i], 0.5),
          border = "#264653", lwd = 1.5)
  text(cx, cy, i, cex = 1.2, font = 2)
}
text(0, -1.9, "4 children per parent", cex = 0.85)

# Aperture 7
plot(NULL, xlim = c(-2, 2), ylim = c(-2, 2), asp = 1,
     axes = FALSE, xlab = "", ylab = "", main = "Aperture 7")

# Parent hexagon
h_parent <- hex_v(0, 0, 1.8, flat_top = TRUE)
polygon(h_parent$x, h_parent$y, border = "#264653", lwd = 2, col = NA)

# Seven child hexagons (rosette pattern: 1 center + 6 around)
child_r7 <- 0.55
# Center
h <- hex_v(0, 0, child_r7, flat_top = TRUE)
polygon(h$x, h$y, col = adjustcolor("#F4A261", 0.5),
        border = "#264653", lwd = 1.5)
text(0, 0, "1", cex = 1, font = 2)

# Surrounding 6
ring_r <- 0.85
ring_cols <- c("#E63946", "#457B9D", "#2A9D8F", "#E9C46A", "#9B5DE5", "#00BBF9")
for (i in 1:6) {
  ang <- pi/2 + (i-1) * pi/3
  cx <- ring_r * cos(ang)
  cy <- ring_r * sin(ang)
  h <- hex_v(cx, cy, child_r7, flat_top = TRUE)
  polygon(h$x, h$y, col = adjustcolor(ring_cols[i], 0.5),
          border = "#264653", lwd = 1.5)
  text(cx, cy, i+1, cex = 1, font = 2)
}
text(0, -1.9, "7 children per parent", cex = 0.85)
```

| Aperture | Children per Parent | Class Pattern | Description |
|----------|---------------------|---------------|-------------|
| 3 | 3 | I/II alternating | Fine resolution control |
| 4 | 4 | I only | Power-of-2 scaling |
| 7 | 7 | III only | Fastest cell growth |
| 4/3 | 4 then 3 | Mixed | Aperture 4 for first m levels, then 3 |

**Resolution** is the number of subdivision levels. Cell count formulas:

| Aperture | Cell Count Formula |
|----------|-------------------|
| 3 | 10 × 3^res + 2 |
| 4 | 10 × 4^res + 2 |
| 7 | 10 × 7^res + 2 |
| 4/3 | 10 × 4^m × 3^(res-m) + 2 |

```{r}
# Cell count growth for all apertures
cat("Resolution  Aperture 3    Aperture 4    Aperture 7\n")
cat("---------  ----------    ----------    ----------\n")
for (res in 0:5) {
  cells_ap3 <- 10 * 3^res + 2
  cells_ap4 <- 10 * 4^res + 2
  cells_ap7 <- 10 * 7^res + 2
  cat(sprintf("    %d      %9d     %9d     %9d\n",
              res, cells_ap3, cells_ap4, cells_ap7))
}
```

For **mixed aperture (4/3)**, the first `m` levels use aperture 4, then remaining levels use aperture 3:

```{r}
# Mixed aperture example: m=2 means first 2 levels use aperture 4
m <- 2
cat("Mixed 4/3 (m=2): first 2 levels aperture 4, rest aperture 3\n")
for (res in 0:5) {
  if (res <= m) {
    cells <- 10 * 4^res + 2
  } else {
    cells <- 10 * 4^m * 3^(res - m) + 2
  }
  cat(sprintf("  Res %d: %d cells\n", res, cells))
}
```

### Orientation Classes

DGGRID defines three orientation classes based on how child cells align with parent cells during subdivision:

- **Class I**: Child cell vertices align with parent cell vertices. Hexagons have a "flat top" orientation.
- **Class II**: Child cell centers align with parent cell vertices. Hexagons are rotated 30° ("pointy top").
- **Class III**: Neither alignment—child cells are rotated relative to parent cells (skewed grids).

Each aperture produces a characteristic class pattern:

| Aperture | Class Pattern | Description |
|----------|---------------|-------------|
| 3 | I → II → I → II... | Alternates between Class I and II at each resolution |
| 4 | I → I → I... | Always Class I (aligned subdivision) |
| 7 | III → III → III... | Always Class III (arctan(√3/5) rotation per level) |
| 4/3 | I → I → ... → I/II | Class I for aperture-4 levels, then alternates |

For **aperture 3**, odd resolutions (1, 3, 5, ...) produce Class I grids with flat-top hexagons, while even resolutions (2, 4, 6, ...) produce Class II grids with pointy-top hexagons.

For **aperture 4**, the 2×2 subdivision preserves alignment, so all resolutions are Class I.

For **aperture 7**, the subdivision involves a √7 scaling and arctan(√3/5) rotation, creating the characteristic Class III skewed pattern. This rotation accumulates: resolution 2 is rotated 2×arctan(√3/5) from resolution 0, and so on.

For **mixed aperture (4/3)**, the class pattern follows aperture 4 (Class I) for the first `m` levels, then switches to aperture 3's alternating I/II pattern for subsequent levels.

The class affects hexagon orientation and the coordinate system internals, but not the fundamental cell assignment or equal-area property.

## Space-Filling Curves

### Why Space-Filling Curves?

Space-filling curves provide:

1. **Compact representation**: Cell identity as a single integer/string
2. **Hierarchical encoding**: Parent-child relationships from index structure
3. **Spatial locality**: Nearby cells often have similar indices

Each aperture requires a different curve that matches its subdivision geometry.

### Z3 Encoding (Aperture 3)

For aperture 3, the Z3 curve uses base-3 (ternary) encoding. Each resolution level adds one ternary digit representing which of the 3 child cells contains the point.

```
Index = FF D₁D₂...Dₙ
where:
  FF = face number (00-19)
  Dᵢ = digit 0, 1, or 2 (base 3)
```

The three child positions form a triangular arrangement within each parent cell. The digit values correspond to:

- **0**: Center-bottom child
- **1**: Upper-left child
- **2**: Upper-right child

This encoding preserves hierarchy: truncating the last digit gives the parent cell index.

```{r}
# Z3 index structure
idx <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 3)
cat("Index:", idx, "\n")
cat("Face:", substr(idx, 1, 2), "\n")
cat("Path:", substr(idx, 3, nchar(idx)), "\n")

# Parent is obtained by dropping the last digit
parent_idx <- substr(idx, 1, nchar(idx) - 1)
cat("Parent:", parent_idx, "\n")
```

### Z-Order (Morton) Encoding (Aperture 4)

For aperture 4, the Morton curve (Z-order curve) interleaves the bits of i and j grid coordinates. This classic space-filling curve maps a 2D grid to a 1D index while preserving locality.

The 2×2 subdivision at each level creates 4 child cells arranged in a square:

```
  +---+---+
  | 2 | 3 |
  +---+---+
  | 0 | 1 |
  +---+---+
```

Each child position is encoded as 2 bits: the low bit is from i, the high bit is from j. For resolution r, the index uses 2r bits total.

```
For cell at grid position (i, j):
  bit 0 = i₀, bit 1 = j₀   (resolution 1)
  bit 2 = i₁, bit 3 = j₁   (resolution 2)
  ...
```

This interleaving creates the characteristic "Z" pattern that gives the curve its name.

```{r}
# Z-order index
idx4 <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 4)
cat("Z-order index:", idx4, "\n")

# The index encodes (quad, i, j) implicitly
coords <- hexify_index_to_cell(idx4, aperture = 4)
cat("Decoded - Quad:", coords$quad, "i:", coords$i, "j:", coords$j, "\n")
```

### Z7 Encoding (Aperture 7)

Aperture 7 uses base-7 encoding with a twist: each subdivision involves both scaling by √7 and rotation by arctan(√3/5). The 7 child cells form a hexagonal rosette pattern:

```
      1
    6   2
      0
    5   3
      4
```

The center cell (0) is surrounded by 6 peripheral cells (1-6) arranged in a hexagonal ring. Unlike apertures 3 and 4, the child cells are rotated relative to the parent, which is why aperture 7 produces Class III grids.

The encoding uses Eisenstein integers (complex numbers of the form a + bω where ω = e^(2πi/3)) to handle the hexagonal geometry. Each digit 0-6 represents a position in the rosette.

```{r}
# Z7 index
idx7 <- hexify_lonlat_to_index(16.37, 48.21, resolution = 3, aperture = 7)
cat("Z7 index:", idx7, "\n")

# Hierarchy still works - parent is obtained by dropping last digit
parent7 <- substr(idx7, 1, nchar(idx7) - 1)
cat("Parent:", parent7, "\n")
```

### Mixed Aperture (4/3) Encoding

Mixed aperture grids combine the aperture 4 and aperture 3 encoding schemes. The first `m` resolution levels use Morton (Z-order) encoding, then subsequent levels use Z3 encoding.

This hybrid approach provides:

- Faster initial cell growth (aperture 4 at coarse scales)
- Finer resolution control (aperture 3 at fine scales)
- dggridR compatibility with ISEA43H grids

```{r, eval = FALSE}
# Mixed aperture index (via hex_grid() with aperture = "4/3")
grid <- hex_grid(area_km2 = 1000, aperture = "4/3")
result <- hexify(data.frame(lon = 16.37, lat = 48.21),
                 lon = "lon", lat = "lat", grid = grid)
cat("Mixed 4/3 cell_id:", result@cell_id, "\n")
```

### Comparison of Encodings

| Property | Z3 (Aperture 3) | Morton (Aperture 4) | Z7 (Aperture 7) | Mixed (4/3) |
|----------|-----------------|---------------------|-----------------|-------------|
| Base | 3 (ternary) | 4 (quaternary) | 7 | 4 then 3 |
| Digits per level | 1 | 2 bits | 1 | Mixed |
| Child arrangement | Triangle | Square | Hex rosette | Square → Triangle |
| Rotation per level | 0° or 30° | 0° | arctan(√3/5) | Mixed |
| Locality | Good | Excellent | Good | Good |

## Inverse Projection

Converting cell coordinates back to lon/lat requires:

1. **Decode index**: Extract face and (i, j) coordinates
2. **Compute cell center**: Find center point in face coordinates
3. **Inverse Snyder**: Convert face (x, y) to geographic (lon, lat)

```{r}
# Round-trip test for all apertures
original_lon <- 16.37
original_lat <- 48.21

cat(sprintf("Original coordinates: (%.4f, %.4f)\n\n", original_lon, original_lat))

for (ap in c(3, 4, 7)) {
  # Use appropriate resolution for each aperture
  res <- if (ap == 7) 6 else 10
  idx <- hexify_lonlat_to_index(original_lon, original_lat,
                                 resolution = res, aperture = ap)
  recovered <- hexify_index_to_lonlat(idx, aperture = ap)
  error <- sqrt((recovered["lon"] - original_lon)^2 +
                (recovered["lat"] - original_lat)^2)
  cat(sprintf("Aperture %d (res %2d): recovered (%.4f, %.4f), error: %.6f°\n",
              ap, res, recovered["lon"], recovered["lat"], error))
}
```

## Cell Properties

### Equal Area

A key property of ISEA grids: all cells have approximately equal area.

```{r}
# Compare cell areas across apertures at similar resolutions
earth_area_km2 <- 510072000

cat("Aperture comparison at ~1000 km2 target:\n\n")
for (ap in c(3, 4, 7)) {
  grid <- hex_grid(area_km2 = 1000, aperture = ap)
  n_cells <- 10 * (ap^grid@resolution) + 2
  cat(sprintf("Aperture %d: resolution %d -> %.1f km2 (%.0f cells)\n",
              ap, grid@resolution, grid@area_km2, n_cells))
}

# Mixed aperture
cat("\n(Mixed aperture 4/3 is also available via hex_grid(aperture = '4/3'))\n")
```

### Cell Shape

Cells are approximately hexagonal, but:

- Cells at face edges may be irregular
- Cell shape varies slightly with latitude due to projection
- Pentagon cells exist at icosahedron vertices (12 total)

## References

- Sahr, K., White, D., & Kimerling, A. J. (2003). Geodesic discrete global grid systems.
- Snyder, J. P. (1992). An equal-area map projection for polyhedral globes.
- DGGRID documentation: https://github.com/sahrk/DGGRID
