---
title: "Mathematical Foundations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

hexify implements the **ISEA (Icosahedral Snyder Equal Area)** discrete global
grid system. This vignette explains the mathematical foundations.

## The Icosahedral Projection

### Why an Icosahedron?

A regular icosahedron has 20 triangular faces, each with identical geometry.
This makes it ideal for:

1. **Equal-area partitioning**: Each face covers approximately 1/20 of Earth's surface
2. **Low distortion**: Each triangular face can be projected with minimal distortion
3. **Hierarchical subdivision**: Triangular faces subdivide naturally

### The Snyder Projection

The Snyder Equal Area projection maps points on the sphere to the icosahedron:

1. **Determine the face**: Find which of 20 icosahedral faces contains the point
2. **Project to face plane**: Use the Snyder formulas to compute (x, y) on the face
3. **Normalize coordinates**: Convert to (tx, ty) in range approximately [0, 1]

```{r}
library(hexify)

# Project a point
result <- hexify_forward(lon = 16.37, lat = 48.21)
cat(sprintf("Face: %d, tx: %.4f, ty: %.4f\n",
            result["face"], result["tx"], result["ty"]))
```

### Icosahedron Orientation

The default ISEA3H orientation places vertex 0 at:
- Longitude: 11.25°
- Latitude: 58.28252559°

```{r}
# Get all face centers
centers <- hexify_face_centers()
head(centers, 5)
```

## Hexagonal Grid Generation

### From Triangles to Hexagons

Each triangular face is subdivided into hexagonal cells. The subdivision process:

1. Overlay a hexagonal lattice on each face
2. Assign each point to the nearest hexagon center
3. Index cells hierarchically

### Aperture and Resolution

**Aperture** determines how many child cells fit in one parent cell:

- **Aperture 3**: Each parent contains 3 children (Class I/II alternating)
- **Aperture 4**: Each parent contains 4 children (Class I only)
- **Aperture 7**: Each parent contains 7 children (Class III)

**Resolution** is the number of subdivision levels. Total cells ≈ 20 × aperture^resolution

```{r}
# Cell count growth
for (res in 0:5) {
  cells_ap3 <- 20 * 3^res
  cells_ap4 <- 20 * 4^res
  cells_ap7 <- 20 * 7^res
  cat(sprintf("Res %d: ap3=%d, ap4=%d, ap7=%d\n",
              res, cells_ap3, cells_ap4, cells_ap7))
}
```

### Class I vs Class II

For aperture 3, the hexagon orientation alternates:

- **Class I** (odd resolution): Hexagons have a flat top
- **Class II** (even resolution): Hexagons are rotated 30°

This affects the coordinate system but not the cell assignment.

## Space-Filling Curves

### Why Space-Filling Curves?

Space-filling curves provide:

1. **Compact representation**: Cell identity as a single integer/string
2. **Hierarchical encoding**: Parent-child relationships from index structure
3. **Spatial locality**: Nearby cells often have similar indices

### Z3 Encoding (Aperture 3)

For aperture 3, the Z3 curve encodes each resolution level as a base-3 digit:

```
Index = FF D₁D₂...Dₙ
where:
  FF = face number (00-19)
  Dᵢ = digit 0, 1, or 2 (base 3)
```

```{r}
# Z3 index structure
idx <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 3)
cat("Index:", idx, "\n")
cat("Face:", substr(idx, 1, 2), "\n")
cat("Path:", substr(idx, 3, nchar(idx)), "\n")
```

### Z-Order Encoding (Aperture 4)

For aperture 4, the Morton (Z-order) curve interleaves i and j coordinates:

```{r}
# Z-order index
idx4 <- hexify_lonlat_to_index(16.37, 48.21, resolution = 5, aperture = 4)
cat("Z-order index:", idx4, "\n")
```

### Z7 Encoding (Aperture 7)

Aperture 7 uses a specialized encoding that handles the 19.1° rotation:

```{r}
# Z7 index
idx7 <- hexify_lonlat_to_index(16.37, 48.21, resolution = 3, aperture = 7)
cat("Z7 index:", idx7, "\n")
```

## Inverse Projection

Converting cell coordinates back to lon/lat requires:

1. **Decode index**: Extract face and (i, j) coordinates
2. **Compute cell center**: Find center point in face coordinates
3. **Inverse Snyder**: Convert face (x, y) to geographic (lon, lat)

```{r}
# Round-trip test
original_lon <- 16.37
original_lat <- 48.21

idx <- hexify_lonlat_to_index(original_lon, original_lat,
                               resolution = 10, aperture = 3)
recovered <- hexify_index_to_lonlat(idx, aperture = 3)

cat(sprintf("Original:  (%.4f, %.4f)\n", original_lon, original_lat))
cat(sprintf("Recovered: (%.4f, %.4f)\n", recovered["lon"], recovered["lat"]))
cat(sprintf("Error: %.6f degrees\n",
            sqrt((recovered["lon"] - original_lon)^2 +
                 (recovered["lat"] - original_lat)^2)))
```

## Cell Properties

### Equal Area

A key property of ISEA grids: all cells have approximately equal area.

```{r}
# Approximate cell area calculation
earth_area_km2 <- 510072000
resolution <- 10
aperture <- 3

total_cells <- 20 * aperture^resolution
cell_area <- earth_area_km2 / total_cells

cat(sprintf("Resolution %d, Aperture %d:\n", resolution, aperture))
cat(sprintf("  Total cells: %.0f\n", total_cells))
cat(sprintf("  Cell area: %.2f km²\n", cell_area))
```

### Cell Shape

Cells are approximately hexagonal, but:

- Cells at face edges may be irregular
- Cell shape varies slightly with latitude due to projection
- Pentagon cells exist at icosahedron vertices (12 total)

## References

- Sahr, K., White, D., & Kimerling, A. J. (2003). Geodesic discrete global grid systems.
- Snyder, J. P. (1992). An equal-area map projection for polyhedral globes.
- DGGRID documentation: https://github.com/sahrk/DGGRID
