---
title: "Practical Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Practical Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates practical workflows for common spatial analysis tasks
using hexify's discrete global grid system.

## Workflow 1: Point Aggregation

### Problem

You have a large dataset of point observations and want to aggregate them into
equal-area cells for analysis or visualization.

### Solution

```{r}
library(hexify)

# Simulate species occurrence data
set.seed(42)
observations <- data.frame(
  species = sample(c("Species A", "Species B", "Species C"), 1000, replace = TRUE),
  lon = runif(1000, -10, 30),  # Europe
  lat = runif(1000, 35, 60)
)

# Assign each observation to a grid cell
observations$cell_id <- hexify_lonlat_to_index(
  observations$lon, observations$lat,
  resolution = 5, aperture = 3
)

# Count observations per cell
cell_counts <- aggregate(
  species ~ cell_id,
  data = observations,
  FUN = length
)
names(cell_counts)[2] <- "count"

head(cell_counts)
```

### Get cell centers for mapping

```{r}
# Add cell center coordinates
cell_centers <- hexify_index_to_lonlat(cell_counts$cell_id, aperture = 3)
cell_counts$center_lon <- cell_centers[, "lon"]
cell_counts$center_lat <- cell_centers[, "lat"]

head(cell_counts)
```

## Workflow 2: Multi-Resolution Analysis

### Problem

You want to analyze data at multiple spatial scales without reprocessing.

### Solution

Use the hierarchical structure of the grid.

```{r}
# Start with fine-resolution cells
fine_cells <- unique(observations$cell_id)

# Get parent cells (coarser resolution)
parent_cells <- sapply(fine_cells, function(idx) {
  hexify_get_parent(idx, aperture = 3)
})

# Get resolution of each level
fine_res <- hexify_get_resolution(fine_cells[1], aperture = 3)
coarse_res <- hexify_get_resolution(parent_cells[1], aperture = 3)

cat(sprintf("Fine resolution: %d (%d unique cells)\n",
            fine_res, length(unique(fine_cells))))
cat(sprintf("Coarse resolution: %d (%d unique cells)\n",
            coarse_res, length(unique(parent_cells))))
```

### Build hierarchy table

```{r}
# Create parent-child relationship table
hierarchy <- data.frame(
  child = fine_cells,
  parent = parent_cells
)

# Aggregate counts to parent level
parent_counts <- aggregate(
  count ~ parent,
  data = merge(
    data.frame(child = observations$cell_id),
    hierarchy
  ) |> merge(cell_counts, by.x = "child", by.y = "cell_id"),
  FUN = sum
)

head(parent_counts)
```

## Workflow 3: Spatial Joins

### Problem

You have two datasets with different point locations and want to join them
based on shared grid cells.

### Solution

```{r}
# Dataset 1: Weather stations
stations <- data.frame(
  station_id = paste0("ST", 1:50),
  lon = runif(50, -10, 30),
  lat = runif(50, 35, 60),
  temperature = rnorm(50, 15, 5)
)

# Dataset 2: Cities
cities <- data.frame(
  city = c("Vienna", "Paris", "London", "Berlin", "Rome",
           "Madrid", "Prague", "Warsaw", "Budapest", "Amsterdam"),
  lon = c(16.37, 2.35, -0.12, 13.40, 12.50,
          -3.70, 14.42, 21.01, 19.04, 4.90),
  lat = c(48.21, 48.86, 51.51, 52.52, 41.90,
          40.42, 50.08, 52.23, 47.50, 52.37)
)

# Assign both to the same grid
resolution <- 4  # Coarse for joining disparate points

stations$cell_id <- hexify_lonlat_to_index(
  stations$lon, stations$lat,
  resolution = resolution, aperture = 3
)

cities$cell_id <- hexify_lonlat_to_index(
  cities$lon, cities$lat,
  resolution = resolution, aperture = 3
)

# Join by cell_id
city_weather <- merge(
  cities[, c("city", "cell_id")],
  aggregate(temperature ~ cell_id, data = stations, FUN = mean),
  by = "cell_id",
  all.x = TRUE
)

city_weather
```

## Workflow 4: Database Storage

### Problem

You need to store cell assignments in a database and want compact integer IDs.

### Solution

Use sequential numbers (seqnum) instead of string indices.

```{r}
# Convert to integer seqnum
observations$seqnum <- hexify_lonlat_to_seqnum(
  observations$lon, observations$lat,
  resolution = 5, aperture = 3
)

# Compare storage requirements
cat("Index string length:", nchar(observations$cell_id[1]), "characters\n")
cat("Seqnum:", observations$seqnum[1], "(integer)\n")

# Round-trip verification
coords <- hexify_seqnum_to_lonlat(
  observations$seqnum[1:5],
  resolution = 5, aperture = 3
)
print(coords)
```

## Workflow 5: Choosing Resolution

### Problem

You want cells of approximately a specific area.

### Solution

Use `hexify_construct()` to find the appropriate resolution.

```{r}
# Target: 100 km² cells
grid_100 <- hexify_construct(area = 100, aperture = 3)
cat(sprintf("For ~100 km² cells: resolution %d (actual area: %.1f km²)\n",
            grid_100$resolution, grid_100$area))

# Target: 1000 km² cells
grid_1000 <- hexify_construct(area = 1000, aperture = 3)
cat(sprintf("For ~1000 km² cells: resolution %d (actual area: %.1f km²)\n",
            grid_1000$resolution, grid_1000$area))

# Target: 10000 km² cells
grid_10000 <- hexify_construct(area = 10000, aperture = 3)
cat(sprintf("For ~10000 km² cells: resolution %d (actual area: %.1f km²)\n",
            grid_10000$resolution, grid_10000$area))
```

### Compare apertures for target area

```{r}
target_area <- 500  # km²

for (ap in c(3, 4, 7)) {
  grid <- hexify_construct(area = target_area, aperture = ap)
  cat(sprintf("Aperture %d: resolution %d, area %.1f km²\n",
              ap, grid$resolution, grid$area))
}
```

## Workflow 6: Working with sf

### Problem

You want to create sf polygons for visualization.

### Solution

```{r eval=FALSE}
library(sf)

# Get unique cells
unique_cells <- unique(observations$cell_id)

# Get cell boundaries (if implemented)
# boundaries <- hexify_cell_to_boundary(grid, unique_cells)

# Alternative: create point layer with cell centers
cell_sf <- st_as_sf(
  cell_counts,
  coords = c("center_lon", "center_lat"),
  crs = 4326
)

# Plot
plot(cell_sf["count"], pch = 16, main = "Observation counts per cell")
```

## Best Practices

### 1. Choose aperture based on use case

- **Aperture 3**: Best for fine resolution control, ecological studies
- **Aperture 4**: Good default, familiar power-of-2 scaling
- **Aperture 7**: Fast resolution jumps, coarse regional analysis

### 2. Store indices efficiently

- Use string indices for interoperability
- Use seqnum for database storage
- Store resolution alongside indices

### 3. Use hierarchical structure

- Pre-compute parent indices for multi-scale analysis
- Navigate hierarchy rather than re-computing

### 4. Handle edge cases

- Check for NA coordinates before processing
- Polar regions (lat > 89°) may have projection artifacts
- Date line (lon = ±180°) works correctly

```{r}
# Example: handling NA values
data_with_na <- data.frame(
  lon = c(16.37, NA, 2.35),
  lat = c(48.21, 48.86, NA)
)

# Filter before processing
valid_rows <- complete.cases(data_with_na[, c("lon", "lat")])
data_with_na$cell_id <- NA
data_with_na$cell_id[valid_rows] <- hexify_lonlat_to_index(
  data_with_na$lon[valid_rows],
  data_with_na$lat[valid_rows],
  resolution = 5, aperture = 3
)

print(data_with_na)
```
