---
title: "Practical Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Practical Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hexify)
```

## Overview

This vignette demonstrates practical workflows for common spatial analysis tasks
using hexify's discrete global grid system and S4 classes.

## One Grid, Many Datasets

The most powerful pattern in hexify is defining a grid once and reusing it across
multiple datasets. This ensures spatial consistency and eliminates parameter repetition.

### The Problem

You often have:

- Several independent datasets (observations, sensors, surveys)
- All in longitude/latitude coordinates
- Collected at different times or from different sources

You want to:

- Put everything on one common global grid
- Be sure the grids actually match
- Combine results later without subtle errors

### The Solution: Shared Grid Objects

```{r one-grid-many-datasets}
library(hexify)

# Step 1: Define the grid once
# This is your shared spatial reference system - like a CRS, but discrete and equal-area
grid <- hex_grid(area_km2 = 5000)
print(grid)

# Step 2: Create multiple datasets with different structures
set.seed(123)

# Dataset 1: Bird observations (note different column names)
bird_obs <- data.frame(
  species = sample(c("Passer domesticus", "Turdus merula", "Parus major"), 200, replace = TRUE),
  longitude = runif(200, -10, 30),
  latitude = runif(200, 35, 60)
)

# Dataset 2: Mammal records
mammal_obs <- data.frame(
  species = sample(c("Vulpes vulpes", "Meles meles", "Sciurus vulgaris"), 150, replace = TRUE),
  lon = runif(150, -10, 30),
  lat = runif(150, 35, 60)
)

# Dataset 3: Climate stations
climate_data <- data.frame(
  station_id = paste0("WS", 1:50),
  x = runif(50, -10, 30),
  y = runif(50, 35, 60),
  temp_c = rnorm(50, 12, 5)
)

# Step 3: Attach all datasets to the SAME grid
# No aperture, resolution, or area parameters needed - the grid carries them
birds <- hexify(bird_obs, lon = "longitude", lat = "latitude", grid = grid)
mammals <- hexify(mammal_obs, lon = "lon", lat = "lat", grid = grid)
climate <- hexify(climate_data, lon = "x", lat = "y", grid = grid)

cat("Birds:  ", nrow(as.data.frame(birds)), "observations in", n_cells(birds), "cells\n")
cat("Mammals:", nrow(as.data.frame(mammals)), "observations in", n_cells(mammals), "cells\n")
cat("Climate:", nrow(as.data.frame(climate)), "stations in", n_cells(climate), "cells\n")
```

### Working at the Cell Level

Once data are hexified, longitude/latitude no longer matter for analysis.
The `cell_id` becomes the shared spatial key:

```{r cell-level-analysis}
# Extract data frames with cell IDs
birds_df <- as.data.frame(birds)
birds_df$cell_id <- birds@cell_id

mammals_df <- as.data.frame(mammals)
mammals_df$cell_id <- mammals@cell_id

climate_df <- as.data.frame(climate)
climate_df$cell_id <- climate@cell_id

# Aggregate each dataset by cell
bird_richness <- aggregate(
  species ~ cell_id,
  data = birds_df,
  FUN = function(x) length(unique(x))
)
names(bird_richness)[2] <- "bird_species"

mammal_richness <- aggregate(
  species ~ cell_id,
  data = mammals_df,
  FUN = function(x) length(unique(x))
)
names(mammal_richness)[2] <- "mammal_species"

mean_temp <- aggregate(
  temp_c ~ cell_id,
  data = climate_df,
  FUN = mean
)
names(mean_temp)[2] <- "mean_temp"

# Join datasets by cell_id - guaranteed to align because same grid
combined <- merge(bird_richness, mammal_richness, by = "cell_id", all = TRUE)
combined <- merge(combined, mean_temp, by = "cell_id", all = TRUE)

head(combined)
```

### Visual Confirmation

All datasets produce identical grid overlays when plotted:

```{r visual-confirmation, fig.width=8, fig.height=4}
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))

# Both plots use the same grid automatically
plot(birds, main = "Bird Observations", basemap = TRUE)
plot(mammals, main = "Mammal Observations", basemap = TRUE)
```

### Key Benefits

1. **No parameter repetition**: Define aperture, resolution, area once
2. **Guaranteed consistency**: All datasets share the exact same grid
3. **Error prevention**: Can't accidentally mix incompatible grids
4. **Clean code**: The grid travels with the data

---

## Workflow 1: Point Aggregation

### Problem

You have a large dataset of point observations and want to aggregate them into
equal-area cells for analysis or visualization.

### Solution

```{r workflow-aggregation}
library(hexify)

# Simulate species occurrence data
set.seed(42)
observations <- data.frame(
  species = sample(c("Species A", "Species B", "Species C"), 1000, replace = TRUE),
  lon = runif(1000, -10, 30),  # Europe
  lat = runif(1000, 35, 60)
)

# Create grid and hexify data
grid <- hex_grid(area_km2 = 10000)
obs_hex <- hexify(observations, lon = "lon", lat = "lat", grid = grid)

# Extract with cell IDs
obs_df <- as.data.frame(obs_hex)
obs_df$cell_id <- obs_hex@cell_id

# Count observations per cell
cell_counts <- aggregate(
  species ~ cell_id,
  data = obs_df,
  FUN = length
)
names(cell_counts)[2] <- "count"

head(cell_counts)
```

### Get cell centers for mapping

```{r cell-centers}
# Add cell center coordinates
cell_centers <- cell_to_lonlat(cell_counts$cell_id, grid)
cell_counts$center_lon <- cell_centers$lon_deg
cell_counts$center_lat <- cell_centers$lat_deg

head(cell_counts)
```

## Workflow 2: Multi-Resolution Analysis

### Problem

You want to analyze data at multiple spatial scales.

### Solution

Use different target areas to get different resolutions.

```{r multi-resolution}
# Fine resolution (~100 km2 cells)
grid_fine <- hex_grid(area_km2 = 100)
obs_fine <- hexify(observations, lon = "lon", lat = "lat", grid = grid_fine)

# Coarse resolution (~10000 km2 cells)
grid_coarse <- hex_grid(area_km2 = 10000)
obs_coarse <- hexify(observations, lon = "lon", lat = "lat", grid = grid_coarse)

cat(sprintf("Fine resolution: %d unique cells (area: %.1f km2)\n",
            n_cells(obs_fine), grid_fine@area_km2))
cat(sprintf("Coarse resolution: %d unique cells (area: %.1f km2)\n",
            n_cells(obs_coarse), grid_coarse@area_km2))
```

### Aggregate at multiple scales

```{r multi-scale-aggregate}
# Extract data with cell IDs
fine_df <- as.data.frame(obs_fine)
fine_df$cell_id <- obs_fine@cell_id

coarse_df <- as.data.frame(obs_coarse)
coarse_df$cell_id <- obs_coarse@cell_id

# Species richness at fine scale
richness_fine <- aggregate(
  species ~ cell_id,
  data = fine_df,
  FUN = function(x) length(unique(x))
)
names(richness_fine)[2] <- "n_species"

# Species richness at coarse scale
richness_coarse <- aggregate(
  species ~ cell_id,
  data = coarse_df,
  FUN = function(x) length(unique(x))
)
names(richness_coarse)[2] <- "n_species"

cat(sprintf("Fine scale: mean %.2f species per cell\n", mean(richness_fine$n_species)))
cat(sprintf("Coarse scale: mean %.2f species per cell\n", mean(richness_coarse$n_species)))
```

## Workflow 3: Spatial Joins

### Problem

You have two datasets with different point locations and want to join them
based on shared grid cells.

### Solution

```{r spatial-joins}
# Dataset 1: Weather stations
stations <- data.frame(
  station_id = paste0("ST", 1:50),
  lon = runif(50, -10, 30),
  lat = runif(50, 35, 60),
  temperature = rnorm(50, 15, 5)
)

# Dataset 2: Cities
cities <- data.frame(
  city = c("Vienna", "Paris", "London", "Berlin", "Rome",
           "Madrid", "Prague", "Warsaw", "Budapest", "Amsterdam"),
  lon = c(16.37, 2.35, -0.12, 13.40, 12.50,
          -3.70, 14.42, 21.01, 19.04, 4.90),
  lat = c(48.21, 48.86, 51.51, 52.52, 41.90,
          40.42, 50.08, 52.23, 47.50, 52.37)
)

# Use a coarse grid for joining disparate points
grid <- hex_grid(area_km2 = 50000)

# Hexify both datasets with the same grid
stations_hex <- hexify(stations, lon = "lon", lat = "lat", grid = grid)
cities_hex <- hexify(cities, lon = "lon", lat = "lat", grid = grid)

# Extract with cell IDs
stations_df <- as.data.frame(stations_hex)
stations_df$cell_id <- stations_hex@cell_id

cities_df <- as.data.frame(cities_hex)
cities_df$cell_id <- cities_hex@cell_id

# Join by cell_id
city_weather <- merge(
  cities_df[, c("city", "cell_id")],
  aggregate(temperature ~ cell_id, data = stations_df, FUN = mean),
  by = "cell_id",
  all.x = TRUE
)

city_weather
```

## Workflow 4: Using hexify() for Complete Workflows

### Problem

You want a simple one-function approach that handles everything.

### Solution

Use `hexify()` which returns a HexData object containing grid info and cell assignments.

```{r complete-workflow}
# Simple workflow with hexify()
grid <- hex_grid(area_km2 = 1000)
result <- hexify(observations, lon = "lon", lat = "lat", grid = grid)

# Access grid info
cat("Resolution:", grid(result)@resolution, "\n")
cat("Cell area:", grid(result)@area_km2, "km2\n")

# Access cell information
cat("Unique cells:", n_cells(result), "\n")

# Cell centers are stored in cell_center slot
cat("First 3 cell centers:\n")
print(head(result@cell_center, 3))
```

## Workflow 5: Choosing Resolution

### Problem

You want cells of approximately a specific area.

### Solution

Use `hex_grid()` to create a grid with the appropriate resolution.

```{r choosing-resolution}
# Target: 100 km2 cells
grid_100 <- hex_grid(area_km2 = 100, aperture = 3)
cat(sprintf("For ~100 km2 cells: resolution %d (actual area: %.1f km2)\n",
            grid_100@resolution, grid_100@area_km2))

# Target: 1000 km2 cells
grid_1000 <- hex_grid(area_km2 = 1000, aperture = 3)
cat(sprintf("For ~1000 km2 cells: resolution %d (actual area: %.1f km2)\n",
            grid_1000@resolution, grid_1000@area_km2))

# Target: 10000 km2 cells
grid_10000 <- hex_grid(area_km2 = 10000, aperture = 3)
cat(sprintf("For ~10000 km2 cells: resolution %d (actual area: %.1f km2)\n",
            grid_10000@resolution, grid_10000@area_km2))
```

### Compare apertures for target area

```{r compare-apertures}
target_area <- 500  # km2

# Compare aperture options
for (ap in c(3, 4, 7)) {
  grid <- hex_grid(area_km2 = target_area, aperture = ap)
  n_cells <- 10 * (ap^grid@resolution) + 2
  cat(sprintf("Aperture %d: resolution %d, area %.1f km2, %.0f cells\n",
              ap, grid@resolution, grid@area_km2, n_cells))
}
```

## Workflow 6: Working with sf

### Problem

You want to create sf polygons for visualization.

### Solution

```{r sf-workflow, eval=requireNamespace("sf", quietly = TRUE)}
library(sf)

# Hexify some data
grid <- hex_grid(area_km2 = 20000)
result <- hexify(cities, lon = "lon", lat = "lat", grid = grid)

# Convert to sf points (fast) - uses cell centers
sf_points <- as_sf(result, geometry = "point")
print(class(sf_points))

# Convert to sf polygons (for choropleth maps)
sf_polys <- as_sf(result, geometry = "polygon")
print(class(sf_polys))

# Or generate polygons directly from cell IDs
unique_cells <- cells(result)
cell_polys <- cell_to_sf(unique_cells, grid)
```

### Visualize with ggplot2

```{r sf-plot, fig.width=7, fig.height=5, message=FALSE}
library(ggplot2)

# Plot polygons with basemap and city labels
europe <- hexify_world[hexify_world$continent == "Europe", ]

ggplot() +
  geom_sf(data = europe, fill = "ivory", color = "gray70") +
  geom_sf(data = cell_polys, fill = "steelblue", alpha = 0.5, color = "darkblue") +
  coord_sf(xlim = c(-10, 25), ylim = c(35, 58)) +
  labs(title = "European Cities - Hexagonal Grid") +
  theme_minimal()
```

## Workflow 7: Visualization Options

hexify provides multiple visualization approaches for different needs.

### Base R plot() Method

```{r viz-base, fig.width=7, fig.height=5}
# Simplest approach - just plot the HexData
plot(result, basemap = TRUE, main = "Base R Plot")
```

### With hexify_ggplot() (ggplot2)

```{r viz-ggplot, fig.width=7, fig.height=5, message=FALSE}
library(ggplot2)

# Returns a ggplot object you can customize
hexify_ggplot(result, basemap = TRUE, title = "ggplot2 hexify_ggplot") +
  theme(legend.position = "none")
```

### Heatmaps for Aggregated Data

```{r viz-heatmap, fig.width=7, fig.height=5, message=FALSE, warning=FALSE}
# Add count data for heatmap
obs_df <- as.data.frame(result)
obs_df$cell_id <- result@cell_id
obs_df$count <- sample(10:100, nrow(obs_df), replace = TRUE)

# Hexify with count column
result_with_count <- hexify(obs_df, lon = "lon", lat = "lat", grid = grid)

# Create heatmap
hexify_heatmap(
  result_with_count,
  value = "count",
  basemap = "world",
  colors = "YlOrRd",
  title = "City Counts",
  xlim = c(-15, 30),
  ylim = c(35, 60)
)
```

## Best Practices

### 1. Choose aperture based on use case

| Aperture | Best For | Trade-offs |
|----------|----------|------------|
| 3 | Fine resolution control, ecological studies, dggridR compatibility | Slowest cell growth |
| 4 | Power-of-2 scaling, GIS workflows | Moderate resolution steps |
| 7 | Rapid cell count growth, coarse analysis | Largest resolution jumps |
| 4/3 | Balance of 4's fast start + 3's fine control | More complex indexing |

### 2. Store indices efficiently

- Cell IDs are numeric (1-based DGGRID-compatible SEQNUM)
- Store resolution alongside cell IDs for later retrieval
- Use `cell_to_lonlat()` to recover coordinates

### 3. Use hex_grid() + hexify() for workflows

```{r best-practice, eval=FALSE}
# Good: Define grid once, reuse
grid <- hex_grid(area_km2 = 1000)
result1 <- hexify(df1, lon = "lon", lat = "lat", grid = grid)
result2 <- hexify(df2, lon = "x", lat = "y", grid = grid)

# Also good: Let hexify create grid internally
result <- hexify(df1, lon = "lon", lat = "lat", area_km2 = 1000)
```

### 4. Handle edge cases

- Check for NA coordinates before processing
- Polar regions (lat > 89 deg) may have projection artifacts
- Date line (lon = +/-180 deg) works correctly

```{r edge-cases}
# Example: handling NA values with hexify()
data_with_na <- data.frame(
  lon = c(16.37, NA, 2.35, 13.40),
  lat = c(48.21, 48.86, NA, 52.52)
)

# hexify handles NAs gracefully with a warning
grid <- hex_grid(area_km2 = 1000)
result <- hexify(data_with_na, lon = "lon", lat = "lat", grid = grid)

# Check which rows have valid cell assignments
cat("Cell IDs:", result@cell_id, "\n")
```

## Summary: Key Functions

| Task | Function |
|------|----------|
| Create grid specification | `hex_grid(area_km2 = ...)` |
| Assign points to cells | `hexify(df, lon, lat, grid)` |
| Get grid from HexData | `grid(result)` |
| Get unique cell IDs | `cells(result)` |
| Count cells | `n_cells(result)` |
| Extract data frame | `as.data.frame(result)` |
| Convert to sf | `as_sf(result, geometry = "polygon")` |
| Generate polygons | `cell_to_sf(cell_ids, grid)` |
| Grid over region | `grid_rect(bbox, grid)` |
| Coordinate conversion | `lonlat_to_cell()`, `cell_to_lonlat()` |
| Visualize | `plot()`, `hexify_ggplot()`, `hexify_heatmap()` |
