---
title: "Practical Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Practical Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates practical workflows for common spatial analysis tasks
using hexify's discrete global grid system.

## One Grid, Many Datasets

The most powerful pattern in hexify is defining a grid once and reusing it across
multiple datasets. This ensures spatial consistency and eliminates parameter repetition.

### The Problem

You often have:

- Several independent datasets (observations, sensors, surveys)
- All in longitude/latitude coordinates
- Collected at different times or from different sources

You want to:

- Put everything on one common global grid
- Be sure the grids actually match
- Combine results later without subtle errors

### The Solution: Shared Grid Objects

```{r one-grid-many-datasets}
library(hexify)

# Step 1: Define the grid once
# This is your shared spatial reference system - like a CRS, but discrete and equal-area
grid <- hex_grid(area_km2 = 5000)
print(grid)

# Step 2: Create multiple datasets with different structures
set.seed(123)

# Dataset 1: Bird observations (note different column names)
bird_obs <- data.frame(
  species = sample(c("Passer domesticus", "Turdus merula", "Parus major"), 200, replace = TRUE),
  longitude = runif(200, -10, 30),
  latitude = runif(200, 35, 60)
)

# Dataset 2: Mammal records
mammal_obs <- data.frame(
  species = sample(c("Vulpes vulpes", "Meles meles", "Sciurus vulgaris"), 150, replace = TRUE),
  lon = runif(150, -10, 30),

  lat = runif(150, 35, 60)
)

# Dataset 3: Climate stations
climate_data <- data.frame(
  station_id = paste0("WS", 1:50),
  x = runif(50, -10, 30),
  y = runif(50, 35, 60),
  temp_c = rnorm(50, 12, 5)
)

# Step 3: Attach all datasets to the SAME grid
# No aperture, resolution, or area parameters needed - the grid carries them
birds <- hexify(bird_obs, lon = "longitude", lat = "latitude", grid = grid)
mammals <- hexify(mammal_obs, lon = "lon", lat = "lat", grid = grid)
climate <- hexify(climate_data, lon = "x", lat = "y", grid = grid)

cat("Birds:  ", nrow(birds), "observations in", n_cells(birds), "cells\n")
cat("Mammals:", nrow(mammals), "observations in", n_cells(mammals), "cells\n")
cat("Climate:", nrow(climate), "stations in", n_cells(climate), "cells\n")
```

### Working at the Cell Level

Once data are hexified, longitude/latitude no longer matter for analysis.
The `cell_id` becomes the shared spatial key:

```{r cell-level-analysis}
# Extract data frames for aggregation
birds_df <- as.data.frame(birds)
mammals_df <- as.data.frame(mammals)
climate_df <- as.data.frame(climate)

# Aggregate each dataset by cell
bird_richness <- aggregate(
  species ~ cell_id,
  data = birds_df,
  FUN = function(x) length(unique(x))
)
names(bird_richness)[2] <- "bird_species"

mammal_richness <- aggregate(
  species ~ cell_id,
  data = mammals_df,
  FUN = function(x) length(unique(x))
)
names(mammal_richness)[2] <- "mammal_species"

mean_temp <- aggregate(
  temp_c ~ cell_id,
  data = climate_df,
  FUN = mean
)
names(mean_temp)[2] <- "mean_temp"

# Join datasets by cell_id - guaranteed to align because same grid
combined <- merge(bird_richness, mammal_richness, by = "cell_id", all = TRUE)
combined <- merge(combined, mean_temp, by = "cell_id", all = TRUE)

head(combined)
```

### Visual Confirmation

All datasets produce identical grid overlays when plotted:

```{r visual-confirmation, fig.width=8, fig.height=4}
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))

# Both plots use the same grid automatically
plot(birds, main = "Bird Observations", pch = 16, cex = 0.5)
plot(mammals, main = "Mammal Observations", pch = 16, cex = 0.5)
```

### Key Benefits

1. **No parameter repetition**: Define aperture, resolution, area once
2. **Guaranteed consistency**: All datasets share the exact same grid

3. **Error prevention**: Can't accidentally mix incompatible grids
4. **Clean code**: The grid travels with the data

---

## Workflow 1: Point Aggregation

### Problem

You have a large dataset of point observations and want to aggregate them into
equal-area cells for analysis or visualization.

### Solution

```{r}
library(hexify)

# Simulate species occurrence data
set.seed(42)
observations <- data.frame(
  species = sample(c("Species A", "Species B", "Species C"), 1000, replace = TRUE),
  lon = runif(1000, -10, 30),  # Europe
  lat = runif(1000, 35, 60)
)

# Assign each observation to a grid cell (vectorized)
observations$cell_id <- hexify_lonlat_to_cell(
  observations$lon, observations$lat,
  resolution = 5, aperture = 3
)

# Count observations per cell
cell_counts <- aggregate(
  species ~ cell_id,
  data = observations,
  FUN = length
)
names(cell_counts)[2] <- "count"

head(cell_counts)
```

### Get cell centers for mapping

```{r}
# Add cell center coordinates
cell_centers <- hexify_cell_to_lonlat(cell_counts$cell_id, resolution = 5, aperture = 3)
cell_counts$center_lon <- cell_centers$lon_deg
cell_counts$center_lat <- cell_centers$lat_deg

head(cell_counts)
```

## Workflow 2: Multi-Resolution Analysis

### Problem

You want to analyze data at multiple spatial scales.

### Solution

Use different target areas to get different resolutions.

```{r}
# Fine resolution (~100 km² cells)
obs_fine <- as.data.frame(hexify(observations, lon = "lon", lat = "lat", area_km2 = 100))

# Coarse resolution (~10000 km² cells)
obs_coarse <- as.data.frame(hexify(observations, lon = "lon", lat = "lat", area_km2 = 10000))

cat(sprintf("Fine resolution: %d unique cells (area: %.1f km²)\n",
            length(unique(obs_fine$cell_id)), obs_fine$cell_area_km2[1]))
cat(sprintf("Coarse resolution: %d unique cells (area: %.1f km²)\n",
            length(unique(obs_coarse$cell_id)), obs_coarse$cell_area_km2[1]))
```

### Aggregate at multiple scales

```{r}
# Species richness at fine scale
richness_fine <- aggregate(
  species ~ cell_id,
  data = obs_fine,
  FUN = function(x) length(unique(x))
)
names(richness_fine)[2] <- "n_species"

# Species richness at coarse scale
richness_coarse <- aggregate(
  species ~ cell_id,
  data = obs_coarse,
  FUN = function(x) length(unique(x))
)
names(richness_coarse)[2] <- "n_species"

cat(sprintf("Fine scale: mean %.2f species per cell\n", mean(richness_fine$n_species)))
cat(sprintf("Coarse scale: mean %.2f species per cell\n", mean(richness_coarse$n_species)))
```

## Workflow 3: Spatial Joins

### Problem

You have two datasets with different point locations and want to join them
based on shared grid cells.

### Solution

```{r}
# Dataset 1: Weather stations
stations <- data.frame(
  station_id = paste0("ST", 1:50),
  lon = runif(50, -10, 30),
  lat = runif(50, 35, 60),
  temperature = rnorm(50, 15, 5)
)

# Dataset 2: Cities
cities <- data.frame(
  city = c("Vienna", "Paris", "London", "Berlin", "Rome",
           "Madrid", "Prague", "Warsaw", "Budapest", "Amsterdam"),
  lon = c(16.37, 2.35, -0.12, 13.40, 12.50,
          -3.70, 14.42, 21.01, 19.04, 4.90),
  lat = c(48.21, 48.86, 51.51, 52.52, 41.90,
          40.42, 50.08, 52.23, 47.50, 52.37)
)

# Assign both to the same grid (coarse for joining disparate points)
stations$cell_id <- hexify_lonlat_to_cell(
  stations$lon, stations$lat,
  resolution = 4, aperture = 3
)

cities$cell_id <- hexify_lonlat_to_cell(
  cities$lon, cities$lat,
  resolution = 4, aperture = 3
)

# Join by cell_id
city_weather <- merge(
  cities[, c("city", "cell_id")],
  aggregate(temperature ~ cell_id, data = stations, FUN = mean),
  by = "cell_id",
  all.x = TRUE
)

city_weather
```

## Workflow 4: Using hexify() for Complete Workflows

### Problem

You want a simple one-function approach that handles everything.

### Solution

Use `hexify()` which returns all cell metadata in one call.

```{r}
# Simple workflow with hexify()
result <- as.data.frame(hexify(observations, lon = "lon", lat = "lat", area_km2 = 1000))

# Result includes all cell information
head(result[, c("species", "lon", "lat", "cell_id", "cell_cen_lon", "cell_cen_lat", "cell_area_km2")])
```

## Workflow 5: Choosing Resolution

### Problem

You want cells of approximately a specific area.

### Solution

Use `hexify_grid()` to find the appropriate resolution.

```{r}
# Target: 100 km² cells
grid_100 <- hexify_grid(area = 100, aperture = 3)
cat(sprintf("For ~100 km² cells: resolution %d (actual area: %.1f km²)\n",
            grid_100$resolution, grid_100$area))

# Target: 1000 km² cells
grid_1000 <- hexify_grid(area = 1000, aperture = 3)
cat(sprintf("For ~1000 km² cells: resolution %d (actual area: %.1f km²)\n",
            grid_1000$resolution, grid_1000$area))

# Target: 10000 km² cells
grid_10000 <- hexify_grid(area = 10000, aperture = 3)
cat(sprintf("For ~10000 km² cells: resolution %d (actual area: %.1f km²)\n",
            grid_10000$resolution, grid_10000$area))
```

### Compare apertures for target area

```{r}
target_area <- 500  # km²

# Compare aperture options
for (ap in c(3, 4, 7)) {
  grid <- hexify_grid(area = target_area, aperture = ap)
  cat(sprintf("Aperture %d: resolution %d, area %.1f km², %.0f cells\n",
              ap, grid$resolution, grid$area, grid$total_cells))
}
# Note: Mixed aperture "4/3" is available via hexify() but not hexify_grid()
```

## Workflow 6: Working with sf

### Problem

You want to create sf polygons for visualization.

### Solution

```{r eval=requireNamespace("sf", quietly = TRUE)}
library(sf)

# Hexify some data
result <- hexify(cities, lon = "lon", lat = "lat", area_km2 = 5000)
result_df <- as.data.frame(result)

# Convert to sf points (fast)
sf_points <- hexify_to_sf(result_df, geometry = "point")

# Convert to sf polygons (for choropleth maps)
sf_polys <- hexify_to_sf(result_df, geometry = "polygon")

# Plot polygons with basemap and city labels
europe <- hexify_world[hexify_world$continent == "Europe", ]
plot(st_geometry(europe), col = "ivory", border = "gray70",
     xlim = c(-10, 25), ylim = c(35, 58), main = "European Cities - Hexagonal Grid")
plot(st_geometry(sf_polys), col = "steelblue", border = "darkblue", add = TRUE)
text(result_df$lon, result_df$lat, result_df$city, cex = 0.7, pos = 3, font = 2)
```

## Best Practices

### 1. Choose aperture based on use case

| Aperture | Best For | Trade-offs |
|----------|----------|------------|
| 3 | Fine resolution control, ecological studies, dggridR compatibility | Slowest cell growth |
| 4 | Power-of-2 scaling, GIS workflows | Moderate resolution steps |
| 7 | Rapid cell count growth, coarse analysis | Largest resolution jumps |
| 4/3 | Balance of 4's fast start + 3's fine control | More complex indexing |

### 2. Store indices efficiently

- Use `cell_id` (integer) for database storage
- Store resolution alongside cell IDs for later retrieval

### 3. Use hexify() for simple workflows

- Returns all metadata in one call
- Handles sf objects automatically
- Auto-calculates resolution from target area

### 4. Handle edge cases

- Check for NA coordinates before processing
- Polar regions (lat > 89°) may have projection artifacts
- Date line (lon = ±180°) works correctly

```{r}
# Example: handling NA values with hexify()
data_with_na <- data.frame(
  lon = c(16.37, NA, 2.35, 13.40),
  lat = c(48.21, 48.86, NA, 52.52)
)

# hexify handles NAs gracefully with a warning
result <- as.data.frame(hexify(data_with_na, lon = "lon", lat = "lat", area_km2 = 1000))
print(result[, c("lon", "lat", "cell_id")])
```
