---
title: "Interactive Globe"
author: "Gilles Colling"
date: "`r Sys.Date()`"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(hexify)
```

`hex_globe()` draws a grid on a globe that turns under the mouse, rendered on
the graphics card through WebGPU. Drag to turn the globe; drag with Shift held
to turn the view about the line of sight; the mouse wheel zooms. The slider
folds the polyhedron the grid is built on into the sphere, and the pointer
shows the ID and value of the cell under it.

The globes on this page need a browser with WebGPU (see
[Where the globe draws](#where-the-globe-draws)); without it each one shows a
notice in its place.

## Installation

```{r, eval = FALSE}
# Install from CRAN
install.packages("hexify")

# Or install development version from GitHub
# install.packages("pak")
pak::pak("gcol33/hexify")
```

## Values per cell

`values` holds one number per cell of `cells`, or one per cell of the grid
when `cells` is `NULL`. Here each cell is filled by the latitude of its centre.

```{r values}
grid <- hex_grid(resolution = 3, aperture = 3)
cells <- seq_len(n_cells(grid))
centres <- cell_to_lonlat(cells, grid)
hex_globe(grid, values = centres$lat_deg, palette = "Blue-Red 3",
          width = "100%", height = 560)
```

The values go to the graphics card as a texture laid out like the grid's
quads, one block per quad, padded with the cells across each quad's edges,
at the grid's resolution and at every coarser one (Jubair et al. 2016 store
icosahedral data in such blocks). Where the cells on screen shrink below a
pixel, the fill is read from the coarser grid whose cells are about a pixel
wide; its cells hold the mean of the values in them, and are as opaque as
the share of them the given cells cover. A fine grid seen whole shows the
pattern of its values rather than noise, and a few scattered cells fade
rather than flicker.

`smooth = TRUE` blends the values of neighbouring cells linearly across the
triangles of cell centres, as a field sampled at the centres.

```{r smooth}
hex_globe(grid, values = centres$lat_deg, palette = "Blue-Red 3", smooth = TRUE,
          width = "100%", height = 560)
```

## The solid

`surface = "solid"` opens on the flat faces of the icosahedron. Moving the
slider to the right folds the faces onto the sphere.

```{r solid}
hex_globe(grid, surface = "solid", center = "pacific", land = FALSE,
          width = "100%", height = 560)
```

An octahedral grid folds the same way.

```{r octahedron}
octa <- hex_grid(resolution = 3, aperture = 4, polyhedron = "octahedron")
hex_globe(octa, surface = "solid", land = FALSE, width = "100%", height = 560)
```

## A fine grid

The shader finds the cell of each pixel from the grid's description, so the
page carries no cell outlines and a fine grid draws as fast as a coarse one.
This grid has `r format(n_cells(hex_grid(resolution = 12, aperture = 3)), big.mark = ",")`
cells. The view opens about 30 km above Europe, with the camera tilted towards the
horizon; the mouse wheel moves it out.

```{r fine}
fine <- hex_grid(resolution = 12, aperture = 3)
hex_globe(fine, center = "europe", land = FALSE,
          projection = "perspective", distance = 1.005, tilt = 50,
          width = "100%", height = 560)
```

## H3

H3 cells are drawn from their outlines, on the sphere only.

```{r h3, message = FALSE}
h3 <- hex_grid(resolution = 1, type = "h3")
hex_globe(h3, land = FALSE, width = "100%", height = 560)
```

## A still image

`hex_globe_png()` draws a globe with the same shader and saves the view it
opens with as a PNG file. It draws through the companion package
[hexglobe](https://github.com/gcol33/hexglobe), which runs WebGPU's shaders
through 'wgpu' in R's own process, or with `renderer = "chrome"` in headless
Chrome. `plot()` on a grid draws the
same views as a static plot without a browser.

```{r png, eval = FALSE}
globe <- hex_globe(grid, surface = "solid", center = "pacific")
hex_globe_png(globe, "globe.png", scale = 2)
```

## Where the globe draws

WebGPU ships by default in these browsers, according to the
[WebGPU implementation status](https://github.com/gpuweb/gpuweb/wiki/Implementation-Status)
kept by the W3C GPU for the Web group:

| Browser | Windows | macOS | Linux |
|---|---|---|---|
| Chrome, Edge | 113 | 113 | 144 (Intel Gen12+), 147 (NVIDIA on Wayland) |
| Firefox | 141 | 145 (Apple Silicon, macOS 26), 147 | not yet |
| Safari | | 26 | |

RStudio's viewer draws the globe: RStudio 2026.08 is built on Electron 42
(Chromium 148). Where WebGPU is missing, open the widget in one of the
browsers above, for example with `htmlwidgets::saveWidget()` and a browser.

## Reference

Jubair, M. I., Alim, U., Röber, N., Clyne, J., and Mahdavi-Amiri, A. (2016).
Icosahedral maps for a multiresolution representation of Earth data. In
*Vision, Modeling, and Visualization (VMV 2016)*.
