---
title: "Lambert's Construction of Snyder's Projection"
author: "Gilles Colling"
date: "`r Sys.Date()`"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 3.6)
library(hexify)
```

Snyder's projection (Snyder 1992) maps each face of the icosahedron, as a
spherical triangle, onto a plane triangle of the same area. Snyder built it
from Lambert's azimuthal equal-area projection centred on the face, and
changed the azimuths and radii of that projection so that the face fills the
triangle exactly. Lambert's projection has a construction with compass and
ruler. In this article I follow the points of one face through that
construction and through Snyder's change of it, step by step, with three
functions:

- `projection_stages()` returns each point's place at every step;
- `projection_distortion(stage = )` returns Tissot's indicatrix at every step;
- `hex_globe(lambert = )` draws the steps on a globe, with a slider.

The sphere has radius 1 throughout. For a face, T is its centre on the
sphere, S a point of the face, z the arc from T to S, and Az the azimuth of S
seen from T, measured from the face's first vertex.

```{r draw-helpers, include = FALSE}
# Drawing helpers for one face at one step: the cell walls, the face outline
# and Tissot's ellipses, as plane coordinates from the face centre.
blue <- "#0072B2"
purple <- "#CC79A7"
grey <- "#9AA4AC"
face_frame <- function(xlim = c(-0.72, 0.72), ylim = c(-0.42, 0.8)) {
  plot.new()
  plot.window(xlim, ylim, asp = 1)
}
draw_face <- function(g, stage, walls = TRUE, ellipses = TRUE, border = "black") {
  if (walls) lines(hexify:::stage_walls(g, 0L, stage), col = blue, lwd = 0.8)
  if (ellipses && stage != "sphere") {
    lines(hexify:::stage_ellipses(g, 0L, stage, steps = 8L, radius = 0.035), col = purple,
          lwd = 1)
  }
  polygon(hexify:::stage_outline(g, 0L, stage), border = border, lwd = 1.5)
}
# A triangular lattice on face 0, `steps` to an edge, as lon/lat
face_lattice <- function(g, steps) {
  ij <- expand.grid(i = 0:steps, j = 0:steps)
  ij <- ij[ij$i + ij$j <= steps, ]
  tx <- (ij$i + 0.5 * ij$j) / steps
  ty <- ij$j * sqrt(3) / 2 / steps
  ic <- hexify:::icosa_arg(g)
  t(vapply(seq_along(tx), function(k) hexify:::cpp_face_xy_to_ll(ic, tx[k], ty[k], 0L),
           numeric(2)))
}
grid <- hex_grid(resolution = 3, aperture = 3)
g_deg <- hexify:::solid_info("icosahedron")$g_deg
```

## One point

Lambert's projection is built on the plane that touches the sphere at T.
The point S lies on a circle with centre T and radius |TS|, in the plane that
holds S and the normal of the sphere at T. S swings along that circle down
onto the touching plane, and lands at the Lambert point L, so that

$$|TL| = |TS| = 2 \sin(z / 2).$$

L keeps the azimuth of S. `projection_stages()` returns the point's plane
coordinates `u` and `v` at each step, along the x and y axes of the face's
triangle, from the face centre:

```{r one-point}
s <- projection_stages(grid, lon = 20, lat = 40)
s[, c("face", "arc", "az", "lambert_u", "lambert_v")]
c(TL = sqrt(s$lambert_u^2 + s$lambert_v^2), chord = 2 * sin(s$arc * pi / 360))
```

The figure shows the construction in the plane of S, T and the sphere's
centre: the sphere's circle, the touching plane at T (the horizontal line
at height 1), and the circle about T that S swings along.

```{r one-point-figure, echo = FALSE, fig.height = 3.4}
z <- s$arc * pi / 180
r <- 2 * sin(z / 2)
par(mar = c(1, 1, 1, 1))
plot.new()
plot.window(c(-0.15, 0.6), c(0.75, 1.08), asp = 1)
th <- seq(-0.1, 0.7, length.out = 200)
lines(sin(th), cos(th), col = grey, lwd = 1.5)
lines(c(-0.15, 0.6), c(1, 1), lwd = 1)
beta <- seq(-z / 2, 0, length.out = 100)
lines(r * cos(beta), 1 + r * sin(beta), col = blue, lwd = 1.5)
segments(0, 1, sin(z), cos(z), lty = 2)
points(c(0, sin(z), r), c(1, cos(z), 1), pch = 19, cex = 0.8)
text(c(0, sin(z), r), c(1, cos(z), 1), c("T", "S", "L"), pos = c(3, 4, 3))
```

## One face

Every point of the face swings down the same way. The face lands on the
touching plane as a triangle whose edges are curved: its corners lie at
distance `r round(2 * sin(g_deg * pi / 360), 4)` from T, and the
middle of each edge farther out than the straight line between two corners.
Lambert's projection is equal-area, so this curved triangle has the face's
area, `4 * pi / 20` = `r round(4 * pi / 20, 4)`, and every cell keeps its
area.

```{r lambert-face, echo = FALSE}
par(mar = c(0.5, 0.5, 0.5, 0.5))
face_frame()
draw_face(grid, "lambert")
```

The ellipses are Tissot's indicatrices: the images of small circles of equal
size on the sphere. On Lambert's projection, at arc z from the centre, a
circle is stretched by $1 / \cos(z / 2)$ across the radius and shrunk by
$\cos(z / 2)$ along it, so its area does not change:

```{r lambert-tissot}
corner <- projection_stages(grid, lon = 11.25, lat = 58.282525588538995, face = 0L)
d <- projection_distortion(grid, c(20, 11.25), c(40, 58.282525588538995), stage = "lambert")
d[, c("a", "b", "areal", "angular")]
```

The second row is a corner of the face (a vertex of the icosahedron, z =
`r round(corner$arc, 2)` degrees), where Lambert's projection deforms angles
most on the face, by `r round(d$angular[2], 2)` degrees.

Cells that lie wholly on the face keep their area on the curved triangle.
For this grid's hexagons on the face, their areas there over their areas on
Snyder's plane triangle, which are their areas on the sphere:

```{r lambert-cell-areas}
paths <- hexify:::grid_surface_paths(grid, NULL, 0.002)
rows <- split(seq_len(nrow(paths)), paths[, "cell"])
inside <- vapply(rows, function(i) all(paths[i, "face"] == 0), logical(1))
inside <- inside & !is_pentagon(as.numeric(names(rows)), grid)
sum(inside)
cell_area <- function(i, stage) {
  ll <- hexify:::vec_lonlat(paths[i, c("sphere_x", "sphere_y", "sphere_z")])
  st <- projection_stages(grid, ll[, 1], ll[, 2], face = 0L)
  u <- st[[paste0(stage, "_u")]]
  v <- st[[paste0(stage, "_v")]]
  abs(sum(u * c(v[-1], v[1]) - c(u[-1], u[1]) * v)) / 2
}
range(vapply(rows[inside], cell_area, numeric(1), stage = "lambert") /
        vapply(rows[inside], cell_area, numeric(1), stage = "snyder"))
```

The walls are drawn through points 0.002 radians apart on the sphere, and
the ratios differ from 1 by that sampling only.

The edges of the curved triangle are not those of a plane triangle, so the
faces of the icosahedron do not fit together on Lambert's projection, and
cells there that cross a face edge do not close up.

## The nudge

Snyder changes Lambert's projection within the touching plane. He moves the
point from the azimuth Az to an azimuth Az' and multiplies its radius by a
factor f, which depends on the azimuth only:

$$\rho = 2 f \sin(z / 2).$$

He chooses Az' so that the ray from the centre at Az' cuts off as much area
of the plane triangle as the ray at Az cuts off of the face, in each of the
six right triangles between the centre, a corner and the middle of an edge
(Snyder 1992, eqs. 6 to 11), and f so that the face edge lands on the
triangle's edge. I call the new point N, the nudged point. The figure shows
the curved triangle in grey and the straight one the nudge makes of it, with
an arrow from L to N at points of the face.

```{r nudge-figure, echo = FALSE}
ll <- face_lattice(grid, 7)
lattice <- projection_stages(grid, ll[, 1], ll[, 2], face = 0L)
par(mar = c(0.5, 0.5, 0.5, 0.5))
face_frame(ylim = c(-0.45, 0.85))
polygon(hexify:::stage_outline(grid, 0L, "lambert"), border = grey, lwd = 1.5)
polygon(hexify:::stage_outline(grid, 0L, "nudge"), border = "black", lwd = 1.5)
arrows(lattice$lambert_u, lattice$lambert_v, lattice$nudge_u, lattice$nudge_v,
       length = 0.04, col = blue)
```

How far the nudge moves the points, over a lattice of points 60 steps to a
face edge. Snyder's projection is equal-area, so a lattice on the plane
triangle spreads the points evenly over the face's area:

```{r nudge-size}
ll <- face_lattice(grid, 60)
dense <- projection_stages(grid, ll[, 1], ll[, 2], face = 0L)
r1 <- attr(dense, "r1")
edge <- sqrt(16 * pi / (sqrt(3) * 20)) / r1
moved <- sqrt((dense$nudge_u - dense$lambert_u)^2 +
              (dense$nudge_v - dense$lambert_v)^2) / edge
turn <- abs((dense$az_prime - dense$az + 180) %% 360 - 180)
c(largest = max(moved), mean = mean(moved), largest_turn_deg = max(turn))
```

`moved` is |LN| over the edge of the nudged triangle. The largest is at the
corners. `paper/bench/bench_lambert_nudge.R` measures the same over 20,301
points and on the octahedron. The nudge moves the corners by about a tenth
of an edge, which shows on a globe without magnification, and the points
near the centre by much less.

The nudge keeps areas in proportion: the nudged triangle has
$1 / R'^2$ = `r round(1 / r1^2, 4)` times the face's area, everywhere on
it. It does not keep the shape of Tissot's ellipses. Snyder's projection
bends along the three lines from the centre to the corners (Snyder 1992,
p. 11): the ellipses on the two sides of such a line differ, and a line on
the sphere that crosses it gets a kink there.

```{r nudge-tissot, echo = FALSE}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5))
face_frame(ylim = c(-0.45, 0.85))
draw_face(grid, "lambert", walls = FALSE)
face_frame(ylim = c(-0.45, 0.85))
draw_face(grid, "nudge", walls = FALSE)
```

On the left Lambert's ellipses, on the right those after the nudge, at the
same points. The largest angular deformation after the nudge is at the
centre, approached along the lines to the corners, 17.27 degrees (Snyder
1992, Table 1):

```{r nudge-centre}
centre <- hexify:::cpp_face_xy_to_ll(hexify:::icosa_arg(grid), 0.5,
                                     1 / (2 * sqrt(3)) + 1e-7, 0L)
projection_distortion(grid, centre[1], centre[2], stage = "nudge")[, c("a", "b", "angular")]
```

## The scaling

The last step scales the touching plane about the sphere's centre by
$R'$ = `r round(r1, 4)`. This takes the plane at distance 1 from the centre
onto the face plane, at distance $R'$, and the nudged triangle onto the
plane triangle with the face's area. It is Snyder's projection:
$\rho = 2 R' f \sin(z / 2)$. The scaling multiplies every length by $R'$, so
Tissot's ellipses keep their shape and the areal scale becomes 1.

```{r scaled-face, echo = FALSE}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5))
face_frame(ylim = c(-0.45, 0.85))
draw_face(grid, "lambert")
face_frame(ylim = c(-0.45, 0.85))
draw_face(grid, "snyder")
```

The last step is the forward projection itself: `tx` and `ty` of
`projection_stages()` are the triangle coordinates `hexify_forward()` gives.

```{r forward}
p <- hexify_forward(20, 40)
c(p[["icosa_triangle_x"]] - s$tx, p[["icosa_triangle_y"]] - s$ty)
```

## Fuller's projection beside it

Fuller's projection (`projection = "fuller"`) maps the same face onto the
same triangle without Lambert's projection behind it, and is not
equal-area. The figure shows the face under Snyder's projection (left) and
Fuller's (right), with Tissot's ellipses of each at the same points.

```{r fuller, echo = FALSE}
fuller <- hex_grid(resolution = 3, aperture = 3, projection = "fuller")
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5))
for (g in list(grid, fuller)) {
  face_frame(ylim = c(-0.45, 0.85))
  draw_face(g, "face")
}
```

Over 5,000 random points of the sphere, the range of the areal scale and
the largest angular deformation of each:

```{r fuller-numbers}
set.seed(1)
lon <- runif(5000, -180, 180)
lat <- asin(runif(5000, -1, 1)) * 180 / pi
df <- projection_distortion(fuller, lon, lat)
ds <- projection_distortion(grid, lon, lat)
rbind(snyder = c(range(ds$areal), max_angular = max(ds$angular)),
      fuller = c(range(df$areal), max_angular = max(df$angular)))
```

## All faces

On the globe every face swings down onto its own touching plane, so the
faces come apart along their edges, and the gaps close again only on the
solid. `hex_globe(lambert = )` opens the globe at a step of the
construction, from 0 (the sphere) through 1 (the Lambert points) and 2 (the
nudged points) to 3 (the faces); the slider moves along it. The globes need
a browser with WebGPU (see the [Interactive Globe](globe.html) article).

```{r globe-lambert}
hex_globe(grid, lambert = 1, center = c(lon = 20, lat = 40),
          width = "100%", height = 560)
```

At step 1 the curved triangles of the faces leave gaps between them, and
coastlines and cell walls stop at the face edges. Moving the slider to 2
straightens the edges; at 3 the faces close into the icosahedron whose
faces have the area of theirs on the sphere, whose corners lie
`r round(r1 / cos(g_deg * pi / 180), 4)` radii from the
centre, outside the sphere.

## One path

The menu beside the slider switches from the three steps to one path per
point: the angle below the touching plane, the azimuth, the radius and the
height of the plane change together, so that the point leaves S along the
compass circle and lands on Snyder's point.

```{r globe-path}
hex_globe(grid, lambert = 0, lambert_path = "single", land = TRUE,
          width = "100%", height = 560)
```

## Earlier pictures of the construction

Wikipedia's article on Lambert's projection describes the construction with
the circle centred at the point of tangency and draws it in a cross-section
figure; it also has an animation in which the sphere opens at a pole into a
disc while every cell of a grid keeps its area
(<https://en.wikipedia.org/wiki/Lambert_azimuthal_equal-area_projection>,
read 2026-10-10). Snyder (1992) describes his projection as Lambert's
azimuthal equal-area projection with the azimuths and the radial distances
adjusted. In web searches on 2026-10-10 I did not find an animation that
carries out Lambert's construction face by face on a polyhedron or shows
Snyder's adjustment as a step of its own.

## Reference

Snyder, J. P. (1992). An equal-area map projection for polyhedral globes.
*Cartographica* 29(1), 10-21. <https://doi.org/10.3138/27H7-8K88-4882-1752>
